---
title: "Lecture 5: Monte Carlo Simulation"
author: "Mark Himmelstein"
format: 
  revealjs:
    slide-width: 1920
    slide-height: 1080
    theme: ../../../gt-theme.scss
    slide-number: true
    slide-number-format: "c/t"
    html-math-method: mathjax
    include-in-header:
      text: |
        <script>
        window.MathJax = {
          tex: {
            packages: {'[+]': ['cancel']}
          },
          loader: {
            load: ['[tex]/cancel']
          }
        };
        </script>
params:
  v0: false 
editor: visual
---

```{r}
#| echo: false


library(knitr)
library(formatR)
library(tidyverse)
opts_chunk$set(out.width = '100%', options(width=60))

opts_knit$set(global.par = TRUE)
opts_chunk$set(echo=FALSE, warning = FALSE, message = FALSE, comment = '', class.output='sh',
               fig.width=8, fig.height=5, fig.align = "center")

my_red <- "#E41A1C"
my_blue <- "#377EB8"
my_green <- "#4DAF4A"
my_purple <- "#984EA3"
my_orange <- "#FF7F00"
my_yellow <- "#FFFF33"
my_red2 <- "#FBB4AE" 
my_blue2 <- "#B3CDE3" 
my_green2 <- "#CCEBC5" 
my_purple2 <- "#DECBE4"
my_orange2 <- "#FED9A6"
my_yellow2 <- "#FFFFCC"

```



## Probability and Statistics

```{r}
#| fig.width: 6
#| fig.height: 2
#| out.width: "100%"

# Zero out standard plot margins
par(mar = c(0, 0, 0, 0))

# Initialize empty canvas
plot(0, 0, col = "white", xlim = c(0, 5), ylim = c(0, 3.5),
     bty = "n", xlab = "", ylab = "", xaxt = "n", yaxt = "n")

# --- Left Box (Population) ---
polygon(x = c(0.2, 1.7, 1.7, 0.2), y = c(2, 2, 3, 3), border = my_blue, lwd = 2)

# --- Right Box (Sample) ---
polygon(x = c(3.3, 4.8, 4.8, 3.3), y = c(.5, .5, 1.5, 1.5), border = my_purple, lwd = 2)

# --- Top Curve (Probability) ---
x_top <- seq(0.95, 4.05, length = 30)
y_top <- -0.478 * x_top^2 + 1.905 * x_top + 1.621
points(x_top, y_top, type = 'l', col = my_blue, lwd = 2)

# Top Arrow (pointing to Sample)
arrows(x0 = x_top[28], x1 = x_top[30], 
       y0 = y_top[28], y1 = y_top[30], 
       col = my_blue, lwd = 2)

# --- Bottom Curve (Statistics) ---
x_bot <- seq(0.95, 4.05, length = 30)
y_bot <- 0.435 * x_bot^2 - 2.656 * x_bot + 4.132
points(x_bot, y_bot, type = 'l', col = my_purple, lwd = 2)

# Bottom Arrow (pointing to Population)
arrows(x0 = x_bot[3], x1 = x_bot[1], 
       y0 = y_bot[3], y1 = y_bot[1], 
       col = my_purple, lwd = 2)

# --- Text Labels ---
text(x = 0.95, y = 2.65, "Population", col = my_blue)
text(x = 0.95, y = 2.35, "(Parameter)", col = my_blue)

text(x = 4.05, y = 1.15, "Sample", col = my_purple)
text(x = 4.05, y = 0.85, "(Data)", col = my_purple)

text(x = 3.1, y = 3.3, "Probability", col = my_blue)
text(x = 1.9, y = 0.3, "Statistics", col = my_purple)
```

::: {style="margin-top: -10px;"}
Monte Carlo simulation is a method of **generating** *random* data using the rules of probability

- Powerful tool for helping to understand the interplay between probability and statistics
- Many methodological developments rely purely on simulation!
- We will use simulation to
    - Boost our intuitions about statistical methods
    - Develop our computational chops

:::


## What Is Randomness?

[Is the outcome of a coin flip random?]{.fragment} [Yes! At least at the limit of human knowledge and perception]{.answer .fragment .content-visible unless-meta="params.v0"}



[Is the number of hours in a day random?]{.fragment} [Not really. It's possible some cosmic event will change this, but we treat a day as a cycle of the earth around the sun that takes 24 hours]{.answer .fragment .content-visible unless-meta="params.v0"}


[Is the amount time it takes for you to get from your home to campus random?]{.fragment} [Sure is! It probably takes you roughly the same amount of time every day, but varies due to unpredictable conditions like traffic, weather, your mood, etc.]{.answer .fragment .content-visible unless-meta="params.v0"}



[**Random** $\neq$ **Arbitrary**]{.fragment}

[Anything that isn't perfectly predictable is subject to randomness, even things with some regularity. Statistics is all about separating the regularity from the randomness as best we can.]{.fragment}

## Psychology and Randomness

Imagine I flip three different coins 10 times.

Which, if any, of the following sequences seems the most/least "random" to you:

- H T H H T H T T T H

- H T H T H T H T H T

- H H H H H T T T T T

[**These sequences are each equally (un)likely.**]{.answer .fragment .content-visible unless-meta="params.v0"}



## Random Variable

A **random variable** is any variable defined by

- A domain: possible values it can take on
- A function: A way of mapping possible values onto numerical outcomes
- A probability distribution: rules governing how likely each numerical outcome is

For example, consider a coin flip:

- **Domain:** \{heads, tails\}
- **Function:**: $X(\text{coin flip}) = \text{heads} = 1; \text{tails} = 0$
- **Probability distribution:** $P(X = 1) = \pi = .5$

## Computational Randomness

Computers can generate sequences of *pseudo* random numbers, which we can use to represent random variables

For example, I can generate numbers from a Bernoulli distribution with $\pi = .5$ to represent my coin flips.

Random number generating functions in `R` typically start with `r` and then the name of the probability distribution. In this case, I'm actually going to use the `rbinom()` function. Why not `rbernoulli()`?

[The Bernoulli distribution is a special case of the binomial where there is only one trial]{.answer .fragment .content-visible unless-meta="params.v0"}

## Simulating Single Bernoullis

```{r}
#| echo: true

set.seed(1204)

rbinom(n = 1, size = 1, prob = .5)

set.seed(5632)

rbinom(n = 1, size = 1, prob = .5)

set.seed(11234)

rbinom(n = 1, size = 1, prob = .5)

set.seed(865)

rbinom(n = 1, size = 1, prob = .5)

```

## `rbinom()`

In the `rbinom()` function I have three arguments

- `n` is the number of "draws"
- `size` is the number of trials per draw
- `prob` is the probability of success on each trial

By increasing `n` I can sample a sequence of zeros and ones, and map those onto heads or tails, per my random variable function

```{r}
#| echo: true

set.seed(4210)
(zeros_and_ones <- rbinom(n = 10, size = 1, prob = .5))
```

The `ifelse()` function takes a vector and checks if a `test` condition is met. If it is, it returns the output specified as `yes`, otherwise it specifies the output returned as `no`.


```{r}

(heads_or_tals <- ifelse(test = zeros_and_ones == 1, 
                         yes = "heads", no = "tails"))
```

## Pseudo Randomness

- Computers are not capable of generating truly random sequences of numbers
- However, the sequence before a repetition can be so large that it becomes functionally random
- The random number generator in `R` will repeat after $2^{19937} - 1$ numbers
    - More than the number of seconds in the history of the universe!
    
Before drawing a sequence of random numbers, we can `set.seed()`

- Determines where in the (extremely long) sequence the random number generation will start
- Makes it so the same sequence can be drawn again to make code precisely reproducible

## Monte Carlo Simulation

- Originally proposed in the 1940s by Stanislaw Ulam, John von Neumann, and Nicholas Metropolis
- Named for Monte Carlo Casino in Monaco

Procedure:

- Define domain
- Define probability distribution
- Draw random numbers
- **Perform computation on random numbers**
- Repeat many times
- Analyze distribution of computation outputs from each iteration

## Loops

There are different techniques for simulating data. One that is common across many programming languages is to use "loops"

- Example: a `for` loop
- Repeats a procedure a finite number of times
- each iteration is indexed
- Indexed outputs can thus be saved and analyzed after loop is complete

## Example

I am going to

- Simulate 10 coin flips (1 = heads, 0 =  tails)
- Change the outcomes so the fifth result is **always** tails
- Calculate the proportion of remaining cases that are heads
- Repeat 1000 times
- Plot the distribution of proportions of remaining heads

## Deleting Odd Heads

Let's first look at how this simulation would look if I just did it *once*

```{r}
flips  <- 10 ## number of flips
p <- .5 ## probability heads

set.seed(521)
raw_outcome <- rbinom(n = flips, size = 1, prob = p)
raw_outcome
mean(raw_outcome) ## proportion heads
```

In this sequence, the fifth outcome was heads, so I will change it to tails

```{r}
raw_outcome[5] <- 0
raw_outcome
mean(raw_outcome)
```

Note that the proportion dropped because I "lost" one heads.



## Another Case

```{r}
set.seed(331)
raw_outcome <- rbinom(n = flips, size = 1, prob = p)
raw_outcome
mean(raw_outcome)

```

In this one, the fifth outcome was **not** heads, so changing it does nothing

```{r}
raw_outcome[5] <- 0
raw_outcome
mean(raw_outcome)

```

## Monte Carloing

Now let's do this 1000 times, and each time, save the proportion of heads that resulted. 

To do this, I'll start by creating an empty vector, then initialize my `for` loop. The for loop uses an index variable, which I am naming `i`, to index iterations. I will save my result into the `i`th entry of my empty vector

I'm going to use a variable `i` to index each iteration

```{r}
iter <- 1000 ## how many iterations in my simulation
raw_result <- numeric(iter) ## empty vector to save results
altered_result <- numeric(iter) ## empty vector to save results
set.seed(124)
for(i in 1:iter){
  raw_outcome <- rbinom(n = 10, size = 1, prob = p)
  raw_result[i] <- mean(raw_outcome) ## save raw result
  raw_outcome[5] <- 0
  altered_result[i] <- mean(raw_outcome) # save altered result
}

```

## Results

I can check descriptive statistics for my results

```{r}
summary(raw_result)
summary(altered_result)
```

Unsurprisingly, the mean before altering our data is approximately 0.50, and is smaller in the cases where we did change our result.

## Results Visualized


```{r}
df <- data.frame(
  value = factor(c(raw_result, altered_result), levels = seq(0, 1, by = 0.1)),
  group = rep(c("Raw", "Altered"), each = 1000)
)

ggplot(df, aes(x = value, fill = group)) +
  geom_bar(position = position_dodge(preserve = "single"), alpha = 0.85) +
  scale_fill_manual(values = c("Raw" = "#377eb8", "Altered" = "#e41a1c")) +
  scale_x_discrete(drop = FALSE) +
  labs(
    x = "Value",
    y = "Count",
    fill = "Version"
  ) +
  theme_minimal()


```

## Other Functions

We will make use of Monte Carlo throughout the semester, often using the normal distribution functions, such as

- `rnorm(n, mean, sd)`

These functions also have associated `p` (probability), `q` (quantile), and `d` (density) functions that are all part of the same families that we will make use of throughout the semester

```{r}
#| eval: false
#| echo: true

?rnorm()
```

```{r}

rd_file <- utils:::.getHelpFile(help("rnorm"))
tools::Rd2HTML(rd_file, out = "rnorm_help.html")
```

<iframe src="rnorm_help.html" width="100%" height="500" style="border:none;"></iframe>