---
title: "Bayesian *t*-Tests and Diagnostics"
subtitle: "PSYC 6019-A01 / PSYC 6022-A01 <br> 2025-10-09 | Lab 7"
author: "Jess Helmer"
footer: "[jhelmer3.github.io/PSYC6019-2026](https://jhelmer3.github.io/PSYC6019-2026/)"
format: 
  revealjs:
    code-link: true
    slide-number: true
    code-line-numbers: false
    theme: ../../gt-theme.scss
    width: 1200
execute: 
  echo: true
  dpi: 600
  freeze: auto
knitr:
  opts_chunk:
    dev: svg
---

## Outline

- Bayesian *t*-tests

- Diagnostics

```{r setup}
#| echo: false
#| output: false
library(tidyverse)
library(gt)
```



# Bayesian *t*-Tests

## Bayesian *t*-Tests

A very user-friendly interface to Bayesian models with `Stan` is the `rstanarm` R package!

```{r}
# install.packages("rstanarm")
library(rstanarm)

filtered_mpg <- mpg |>
  filter(when_any(cyl == 4, cyl == 8))
```

<br>

:::: {.columns}
::: {.column width="50%"}
```{r}
freq_t_test <- t.test(hwy ~ cyl, data = filtered_mpg)
freq_t_test
```
:::

::: {.column width="50%"}
```{r}
#| message: false
#| eval: false
bayes_t_test <- stan_glm(hwy ~ cyl, data = filtered_mpg,
                         family = "gaussian")
summary(bayes_t_test)
```

```{r}
#| echo: false
bayes_t_test <- stan_glm(hwy ~ cyl, data = filtered_mpg,
                         family = "gaussian", refresh = 0)
summary(bayes_t_test)
```
:::
::::


## Bayesian *t*-Tests

```{r}
prior_summary(bayes_t_test)
```


## Model

$$
\begin{aligned}
\text{highway MPG}_i &= \beta_0 + \beta_1 \text{cylinders}_i + \epsilon_i && \text{likelihood: a linear model} \\
\beta_0 &\sim N(24, 17) && \text{prior for intercept} \\
\beta_1 &\sim N(0, 8.6) && \text{prior for slope} \\
\epsilon &\sim \text{Exponential}(0.15)  && \text{prior for residual variance}
\end{aligned}
$$

## Credible Intervals

:::: {.columns}
::: {.column width="50%"}
**Equal Tailed Interval**

```{r}
posterior_interval(bayes_t_test)
```
:::

::: {.column width="50%"}

**Highest Density Interval**

```{r}
# install.packages("bayestestR")
bayestestR::hdi(bayes_t_test, component = "all")
```
:::
::::


<!-- ```{r} -->
<!-- etis <- posterior_interval(bayes_t_test) |> -->
<!--   as.data.frame() -->
<!-- hdpis <- bayestestR::hdi(bayes_t_test, component = "all") |> -->
<!--   as.data.frame() -->


<!-- bayes_t_test |> -->
<!--   bayesplot::mcmc_dens(pars = c("(Intercept)")) + -->
<!--   geom_vline(aes(xintercept = etis["(Intercept)", "5%"])) + -->
<!--   geom_vline(aes(xintercept = etis["(Intercept)", "95%"])) + -->
<!--   # geomtextpath::geom_textsegment(aes(x = etis["(Intercept)", "5%"], xend = etis["(Intercept)", "95%"],  -->
<!--   #                                    y = Inf, yend = Inf, label = "90% CTI", vjust = 1)) + -->
<!--   geom_rect(aes(xmin = hdpis[hdpis$Parameter == "(Intercept)", "CI_low"], xmax = hdpis[hdpis$Parameter == "(Intercept)", "CI_high"], ymin = 0, ymax = 0.5), alpha = 0.3, fill = "pink") -->
<!--   geom_vline(aes(xintercept = hdpis[hdpis$Parameter == "(Intercept)", "CI_low"])) + -->
<!--   geom_vline(aes(xintercept = hdpis[hdpis$Parameter == "(Intercept)", "CI_high"])) + -->
<!--   geomtextpath::geom_textsegment(aes(x = hdpis[hdpis$Parameter == "(Intercept)", "CI_low"], xend = hdpis[hdpis$Parameter == "(Intercept)", "CI_high"],  -->
<!--                                      y = Inf, yend = Inf, label = "90% HDPI", vjust = 1)) -->
<!-- ``` -->

# Diagnostics

## Diagnostics with Bayes

Most of the time, Bayes' theorem $P(\theta | D) = \frac{P(\theta) P(D | \theta)}{P(D)}$ is too complex to solve analytically. So, we estimate Bayesian models with Markov Chain Monte Carlo (MCMC).

The job of MCMC is to explore and sample from the posterior probability distribution $P(\theta | D)$ for all $\theta$ at a rate proportional to each probability.

So, after many MCMC samples, we end up with a density distribution of samples with height matching $P(\theta | D)$, and we can use those samples to make inferences about our parameters and outcomes!

::: subpoint
○ E.g., we believe there is a 90% chance the difference in highway MPG between four- and eight-cylinder cars is within the interval [`r posterior_interval(bayes_t_test)["cyl", ] |> round(2) |> paste(collapse = ", ")`].
:::

But, all this hinges on the MCMC sampler---we have to check it for evidence that it *sampled well*.


## Looking at the Samples

Often, a good first step for checking the sampler is just looking at what it sampled!

::: fragment
```{r}
#| fig-height: 2
#| fig-width: 6
scale_colors <-  c("steelblue", "#D98282", "#6BAE9D", "#8C7AAE")
x <- bayesplot::example_mcmc_draws(chains = 1, params = 1)
bayesplot::mcmc_trace(x) +
  scale_color_manual(values = scale_colors)
```
:::

[This is called a *trace plot*. It shows the samplers' samples for each iteration it runs. Can you tell if this sampler sampled well?]{.fragment}


## Looking at the Samples

Often, a good first step for checking the sampler is just looking at what it sampled!

```{r}
#| fig-height: 2
#| fig-width: 6
x <- bayesplot::example_mcmc_draws(chains = 2, params = 1)
bayesplot::mcmc_trace(x) +
  scale_color_manual(values = scale_colors)
```

[What if I told you we ran another, completely separate run, and its samples looked like this?]{.fragment}


## Looking at the Samples

Often, a good first step for checking the sampler is just looking at what it sampled!

```{r}
#| fig-height: 2
#| fig-width: 6
x <- bayesplot::example_mcmc_draws(chains = 3, params = 1)
bayesplot::mcmc_trace(x) +
  scale_color_manual(values = scale_colors)
```

[Or a third?]{.fragment}


## Looking at the Samples

Often, a good first step for checking the sampler is just looking at what it sampled!

```{r}
#| fig-height: 2
#| fig-width: 6
x <- bayesplot::example_mcmc_draws(chains = 4, params = 1)
bayesplot::mcmc_trace(x) +
  scale_color_manual(values = scale_colors)
```

[A fourth? How much confidence do you have in the sampler now?]{.fragment}

[They cannot guarantee the sampler is **correctly** sampling from the distribution, but if we run multiple *chains* (separate runs of MCMC, each doing their own thing independently), we can at least check if the sampler is sampling **consistently**.]{.fragment}

## House Layout Example


## Looking at the Samples

Our example's trace plots!

```{r}
#| fig-height: 3
#| fig-width: 12
# install.packages("bayesplot")
library(bayesplot)

mcmc_trace(bayes_t_test)
```

[Good! They look like "fuzzy caterpillars".]{.fragment}


## Trace Rank Plots

These are called "trace rank" plots. They're very similar to trace plots, except they *order* the samples in magnitude.

It's mostly personal preference, but these are sometimes easier to see bad chains.

```{r}
mcmc_rank_overlay(bayes_t_test)
```

[What does a bad chain look like?]{.fragment}


## Bad Sampling Example

Messing with the algorithm on purpose to make it sample poorly...

```{r}
#| message: false
#| fig-height: 2.5
#| fig-width: 10
#| echo: false
bad_example <- stan_glm(hwy ~ cyl, data = filtered_mpg, seed = 12345, adapt_delta = 0.3)
mcmc_trace(bad_example) +
  scale_color_manual(values = scale_colors)
mcmc_rank_overlay(bad_example) +
  scale_color_manual(values = scale_colors)
```

[Ick!]{.fragment}


## Rhat

Looking at traceplots is great, but it's also helpful to have a numerical measure of how well the chains mixed.

Rhat ($\hat{R}$, read "R-hat") is a measure of convergence among different MCMC *chains*.

The main idea of $\hat{R}$ is to compare the variance *between* chains to the variance *within* chains.


## Rhat Visualized: Bad

The main idea of $\hat{R}$ is to compare the variance *between* chains to the variance *within* chains.

```{r}
#| echo: false
set.seed(1)
n_samples <- 500
d <- tibble(chain = 1:4) |>
  mutate(
    chain_mu = c(3, 3.3, 2.3, 5),
    samples = map(
      chain_mu, \(chain_mu) tibble(
        rep = 1:n_samples, 
        x = rnorm(n_samples, chain_mu, 0.3)
      )
    )
  ) |>
  unnest(samples)

chain_means <- d |>
  summarize(.by = chain, x = mean(x))

d |>
  ggplot(aes(x = rep, y = x, color = factor(chain), group = chain)) +
  geom_line() +
  guides(x = guide_axis(cap = T),
         y = guide_none(NULL),
         color = guide_none(),
         fill = guide_none()) +
  scale_x_continuous("Rep", limits = c(-50, n_samples + 50), breaks = seq(0, 500, 100)) +
  scale_color_manual(values = scale_colors, aesthetics = c("color", "fill")) +
  coord_cartesian(ylim = c(1.2, 6)) +
  theme_classic()
```


## Rhat Visualized: Bad

The main idea of $\hat{R}$ is to compare the variance *between* chains to the variance *within* chains.

```{r}
#| echo: false
d |>
  ggplot(aes(x = rep, y = x, color = factor(chain), group = chain)) +
  geom_line() +
  geom_point(
    data = chain_means |> mutate(rep = 0 - 15),
    aes(x = rep, y = x),
    size = 3
  ) +
  guides(x = guide_axis(cap = T),
         y = guide_none(NULL),
         color = guide_none(),
         fill = guide_none()) +
  scale_x_continuous("Rep", limits = c(-50, n_samples + 50), breaks = seq(0, 500, 100)) +
  scale_color_manual(values = scale_colors, aesthetics = c("color", "fill")) +
  coord_cartesian(ylim = c(1.2, 6)) +
  theme_classic()
```


## Rhat Visualized: Bad

The main idea of $\hat{R}$ is to compare the variance *between* chains to the variance *within* chains.

```{r}
#| echo: false
d |>
  ggplot(aes(x = rep, y = x, color = factor(chain), group = chain)) +
  geom_line() +
  ggdist::stat_slab(
    data = d |> mutate(rep = n_samples + 10),
    aes(x = rep, y = x, fill = factor(chain)),
    color = NA,
    side = "right",
    scale = 30,
    alpha = 0.7
  ) +
  ggdist::stat_slab(
    data = d |> mutate(rep = 0 - 15),
    aes(x = rep, y = x),
    color = NA,
    side = "left",
    scale = 30,
    alpha = 0.7,
    inherit.aes = F
  ) +
  geom_point(
    data = chain_means |> mutate(rep = 0 - 15),
    aes(x = rep, y = x),
    size = 3
  ) +
  guides(x = guide_axis(cap = T),
         y = guide_none(NULL),
         color = guide_none(),
         fill = guide_none()) +
  scale_x_continuous("Rep", limits = c(-50, n_samples + 50), breaks = seq(0, 500, 100)) +
  scale_color_manual(values = scale_colors, aesthetics = c("color", "fill")) +
  coord_cartesian(ylim = c(1.2, 6)) +
  theme_classic()
```

Variance between chain means is sooo different from variance within chains!


## Rhat Visualized: Good

The main idea of $\hat{R}$ is to compare the variance *between* chains to the variance *within* chains.

```{r}
#| echo: false
set.seed(1)
n_samples <- 500
d <- tibble(chain = 1:4) |>
  mutate(
    chain_mu = mean(c(3, 3.3, 2.3, 5)),
    samples = map(
      chain_mu, \(chain_mu) tibble(
        rep = 1:n_samples, 
        x = rnorm(n_samples, chain_mu, 0.3)
      )
    )
  ) |>
  unnest(samples)

chain_means <- d |>
  summarize(.by = chain, x = mean(x))

d |>
  ggplot(aes(x = rep, y = x, color = factor(chain), group = chain)) +
  geom_line() +
  guides(x = guide_axis(cap = T),
         y = guide_none(NULL),
         color = guide_none(),
         fill = guide_none()) +
  scale_x_continuous("Rep", limits = c(-50, n_samples + 50), breaks = seq(0, 500, 100)) +
  scale_color_manual(values = scale_colors, aesthetics = c("color", "fill")) +
  coord_cartesian(ylim = c(1.2, 6)) +
  theme_classic()
```


## Rhat Visualized: Good

The main idea of $\hat{R}$ is to compare the variance *between* chains to the variance *within* chains.

```{r}
#| echo: false
d |>
  ggplot(aes(x = rep, y = x, color = factor(chain), group = chain)) +
  geom_line() +
  geom_point(
    data = chain_means |> mutate(rep = 0 - 15),
    aes(x = rep, y = x),
    size = 3
  ) +
  guides(x = guide_axis(cap = T),
         y = guide_none(NULL),
         color = guide_none(),
         fill = guide_none()) +
  scale_x_continuous("Rep", limits = c(-50, n_samples + 50), breaks = seq(0, 500, 100)) +
  scale_color_manual(values = scale_colors, aesthetics = c("color", "fill")) +
  coord_cartesian(ylim = c(1.2, 6)) +
  theme_classic()
```


## Rhat Visualized: Good

The main idea of $\hat{R}$ is to compare the variance *between* chains to the variance *within* chains.

```{r}
#| echo: false
d |>
  ggplot(aes(x = rep, y = x, color = factor(chain), group = chain)) +
  geom_line() +
  ggdist::stat_slab(
    data = d |> mutate(rep = n_samples + 10),
    aes(x = rep, y = x, fill = factor(chain)),
    color = NA,
    side = "right",
    scale = 30,
    alpha = 0.7
  ) +
  ggdist::stat_slab(
    data = d |> mutate(rep = 0 - 15),
    aes(x = rep, y = x),
    color = NA,
    side = "left",
    scale = 30,
    alpha = 0.7,
    inherit.aes = F
  ) +
  geom_point(
    data = chain_means |> mutate(rep = 0 - 15),
    aes(x = rep, y = x),
    size = 3
  ) +
  guides(x = guide_axis(cap = T),
         y = guide_none(NULL),
         color = guide_none(),
         fill = guide_none()) +
  scale_x_continuous("Rep", limits = c(-50, n_samples + 50), breaks = seq(0, 500, 100)) +
  scale_color_manual(values = scale_colors, aesthetics = c("color", "fill")) +
  coord_cartesian(ylim = c(1.2, 6)) +
  theme_classic()
```

Variance between chain means is very close to variance within chains!


## Rhat

In practice, you want to see $\hat{R}$ values less than 1.01. 

::: fragment
```{r}
summary(bayes_t_test)[, "Rhat"]
```
:::


## Effective Sample Size (ESS)

In order to closely approximate the posterior ($P(\theta | D)$, the point of this all!), we need to make sure we have a large number of MCMC samples.


## Lab 07 Assignment























