---
title: "Data Simulations and Chi-Squared Tests"
subtitle: "PSYC 6019-A01 / PSYC 6022-A01 \n 2025-09-11 | Lab 3"
author: "Jess Helmer"
footer: "[jhelmer3.github.io/PSYC6019-2026/](https://jhelmer3.github.io/PSYC6019-2026/)"
format: 
  revealjs:
    code-link: true
    code-line-numbers: false
    theme: ../../gt-theme.scss
execute: 
  echo: true
  dpi: 600
---

## Outline

- Simulating data introduction
- Central Limit Theorem example
- $\chi^2$ distribution
  
```{r setup}
#| echo: false
#| output: false
# targets::tar_load(store = here::here("_targets"))
```


## Simulation

why do we like sims


## Simulation Example: CLT

The Central Limit Theorem (CLT) says that the sampling distribution of the mean of any distribution with finite mean and variance will converge to be normal with increasing sample size.

::: subpoint
○ Imagine: take a sample of $N$ observations from the population, calculate some statistics (e.g., the mean), and toss them back.
:::

::: subpoint
○ Repeat a whole bunch of times
:::

::: subpoint
○ Sampling distribution: the distribution of those calculated statistics (e.g., means)
:::

Let's say we're not so sure, and we want to test this out with different distributions and sample sizes $N$.


## Simulation Example: CLT

:::: {.columns}
::: {.column width="50%"}
Before simulating, define your goals / questions.

Does the sampling distribution of the mean really converge to normal for wonky asymmetrical distributions like the $\chi^2$ distribution?

At what sample sizes does this fail?
:::

::: {.column width="50%"}
Then, consider your plan.


:::
::::





## Simulation Example: CLT



```{r}

```





























