---
title: "Effect Size and Power Analysis"
subtitle: "PSYC 6019-A01 / PSYC 6022-A01 <br> 2025-10-02 | Lab 6"
author: "Jess Helmer"
footer: "[jhelmer3.github.io/PSYC6019-2026](https://jhelmer3.github.io/PSYC6019-2026/)"
format: 
  revealjs:
    code-link: true
    slide-number: true
    code-line-numbers: false
    theme: ../../gt-theme.scss
    width: 1200
execute: 
  echo: true
  dpi: 600
  freeze: auto
knitr:
  opts_chunk:
    dev: svg
---

## Outline

- Effect size

- Power Analysis

```{r setup}
#| echo: false
#| output: false
library(tidyverse)
library(gt)
```



# Effect Size

## Effect Size

Effect sizes provide ways to estimate the magnitude of an effect

::: subpoint
○ Standardized, so we can compare differences across scales
:::

::: fragment
Cohen's $d$: measure of difference (not a test)

::: subpoint
○ Transforms to the *z*-scale, can interpret in standard deviation units
:::
:::


## Cohen's *d*

:::: {.columns}
::: {.column width="28%"}
#### One Sample

$$ d = \frac{\bar{x} - \mu_0}{s_x} $$


::: {style="font-size:85%}
```{r}
mu_cty <- 30

cty_bar <- mean(mpg$cty)
cty_sd <- sd(mpg$cty)

(cty_bar - mu_cty) / cty_sd
```
:::
:::

::: {.column width="44%"}
#### Independent Samples

$$ d = \frac{\bar{x}_1 - \bar{x}_2}{s_p} $$

::: {style="font-size:85%}

::: panel-tabset
### Pull Data
```{r}
four_cyl <- mpg |>
  filter(cyl == 4) |>
  pull(hwy)
eight_cyl <- mpg |>
  filter(cyl == 8) |>
  pull(hwy)
```


### Calculate $d$
```{r}
x_bar_four <- mean(four_cyl)
sd_four <- sd(four_cyl)
n_four <- length(four_cyl)
x_bar_eight <- mean(eight_cyl)
sd_eight <- sd(eight_cyl)
n_eight <- length(eight_cyl)

pooled_sd <- sqrt( 
  ((n_four - 1) * sd_four^2 + (n_eight - 1) * sd_eight^2) /
    ((n_four - 1) + (n_eight - 1)) 
  )

(x_bar_four - x_bar_eight) / pooled_sd
```
:::

:::


:::

::: {.column width="28%"}
#### Paired Sample

$$ d = \frac{\bar{D}}{s_D} $$

::: {style="font-size:85%}
```{r}
differences <- mpg$cty - mpg$hwy

d_bar <- mean(differences)
d_sd <- sd(differences)

d_bar / d_sd
```
:::

:::
::::


## Cohen's *d*

:::: {.columns}
::: {.column width="33%"}
#### One Sample

$$ d = \frac{\bar{x} - \mu_0}{s_x} $$


::: {style="font-size:85%}
```{r}
effectsize::cohens_d(mpg$cty, mu = 30)
```
:::
:::

::: {.column width="33%"}
#### Independent Samples

$$ d = \frac{\bar{x}_1 - \bar{x}_2}{s_p} $$

::: {style="font-size:85%}
```{r}
four_cyl <- mpg |>
  filter(cyl == 4) |>
  pull(hwy)
eight_cyl <- mpg |>
  filter(cyl == 8) |>
  pull(hwy)

effectsize::cohens_d(four_cyl, eight_cyl)
```
:::


:::

::: {.column width="33%"}
#### Paired Sample

$$ d = \frac{\bar{D}}{s_D} $$

::: {style="font-size:85%}
```{r}
effectsize::cohens_d(mpg$cty, mpg$hwy, paired = T)
```
:::

:::

::::


## Cohen's *D* Strength Visualization

```{r}
#| echo: false
x_bar <- mean(mpg$hwy)
x_sd <- sd(mpg$hwy)
```


Based on mean ($\bar{x}$ = `r round(x_bar, 2)` = `r round(x_bar, 2)`) and SD ($\bar{\sigma}$ = `r round(x_sd, 2)` = `r round(x_sd, 2)`) of cars' highway MPG (`hwy`).

Solved backwards for ($\mu$) based on different potential levels of *d*. "What would the hypothesized mean have to have been to get different levels of Cohen's *d*?"

$$\bar{x} - d \times s = \mu$$

```{r}
dat <- data.frame(x = rnorm(100000, x_bar, x_sd))

mu_ex1 <- x_bar - .2 * x_sd
mu_ex2 <- x_bar - .5 * x_sd
mu_ex3 <- x_bar - .8 * x_sd
```


## Cohen's *D* Strength Visualization

"What would the hypothesized mean have to have been to get different levels of Cohen's *d*?"

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| fig-align: center
ggplot(dat, aes(x = x)) +
  geom_histogram(fill = 'skyblue3', alpha = .4, bins = 16) +
  geom_vline(aes(xintercept = x_bar),
             color = "skyblue3", linewidth = 1, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 60), expand = F, clip = "off") +
  guides(y = "none", x = NULL) +
  labs(y = NULL, x = "Highway MPG") +
  theme_classic(base_size = 18)
```


## Cohen's *D* Strength Visualization

"What would the hypothesized mean have to have been to get different levels of Cohen's *d*?"


```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| layout-ncol: 3
#| fig-align: center
ggplot(dat, aes(x = x)) +
  geom_histogram(fill = 'skyblue3', alpha = .4, bins = 16) +
  geom_vline(aes(xintercept = x_bar),
             color = "skyblue3", linewidth = 1, linetype = "dashed") +
  geomtextpath::geom_textvline(aes(xintercept = mu_ex1, label = "d = 0.2"),
                               hjust = .3, size = 10, color = "steelblue4", linewidth = 1.5) +
  coord_cartesian(xlim = c(0, 60), expand = F, clip = "off") +
  guides(y = "none", x = NULL) +
  labs(y = NULL, x = "Highway MPG") +
  theme_classic(base_size = 18)

ggplot(dat, aes(x = x)) +
  geom_histogram(fill = 'skyblue3', alpha = .4, bins = 16) +
  geom_vline(aes(xintercept = x_bar),
             color = "skyblue3", linewidth = 1, linetype = "dashed") +
  geomtextpath::geom_textvline(aes(xintercept = mu_ex2, label = "d = 0.5"),
                               hjust = .3, size = 10, color = "steelblue4", linewidth = 1.5) +
  coord_cartesian(xlim = c(0, 60), expand = F, clip = "off") +
  guides(y = "none", x = NULL) +
  labs(y = NULL, x = "Highway MPG") +
  theme_classic(base_size = 18)

ggplot(dat, aes(x = x)) +
  geom_histogram(fill = 'skyblue3', alpha = .4, bins = 16) +
  geom_vline(aes(xintercept = x_bar),
             color = "skyblue3", linewidth = 1, linetype = "dashed") +
  geomtextpath::geom_textvline(aes(xintercept = mu_ex3, label = "d = 0.8"),
                               hjust = .3, size = 10, color = "steelblue4", linewidth = 1.5) +
  coord_cartesian(xlim = c(0, 60), expand = F, clip = "off") +
  guides(y = "none", x = NULL) +
  labs(y = NULL, x = "Highway MPG") +
  theme_classic(base_size = 18)

```



# Power Analysis


## Power Analysis

*Power*: the proportion of times we'd correctly reject the null hypothesis $H_0$ when something else is true, over many hypothetical repeated experiments

*Power analysis*: the chance of correctly rejecting $H_0$ if $H_0$ is not the reality.

::: {.subpoint .fragment}
○ Given that our null hypothesis is $H_0: \mu = \mu_0$,
:::

::: {.subpoint .fragment}
○ How often would we correctly reject the null if $\mu$ was not $\mu_0$?
:::

$\beta$: probability of incorrectly retaining the null when $H_0$ is false
$1 - \beta$: power (probability of correctly rejecting the null when $H_0$ is false)


## Power Analysis

What can we change in power analysis?

::: fragment
| $1 - \beta$ | $N$         | $\alpha$           | $\delta$         |
|-------------|-------------|--------------------|------------------|
| Power       | Sample size | significance level | true effect size |
:::

::: fragment
If know three, can solve for the fourth. 
:::

::: fragment
Can manipulate potential values to create power curves for different combinations
:::


## Power Analysis

We can use the `pwr` package in `R` to conduct power analyses! For a *t*-test, 

`pwr.t.test(n, d, sig.level, power)`

<div class="subpoint">○ `n` = sample size **in each group** </div>

<div class="subpoint">○ `d` = true effect size </div>

<div class="subpoint">○ `sig.level` = significance level, $\alpha$ </div>

<div class="subpoint">○ `power` = power, $1 - \beta$ </div>

Specify exactly three of the four, it will fill in the fourth

<div class="subpoint">○ `alternative` = type of test, default is `"two.sided"` </div>

:::: {.columns}
::: {.column width="33%"}
```{r}
pwr::pwr.t.test(n = 10, d = 0.3, sig.level = 0.05)
```
:::

::: {.column width="33%"}
```{r}
pwr::pwr.t.test(n = 50, sig.level = 0.05, power = 0.8)
```
:::

::: {.column width="33%"}
```{r}
pwr::pwr.t.test(d = 0.3, sig.level = 0.05, power = 0.8)
```
:::
::::


## Power Analysis

Object contains a list of each argument, including the one it calculated

```{r}
power_test <- pwr::pwr.t.test(n = 10, d = 0.3, sig.level = 0.05)

power_test |>
  glimpse()
```

Can pull the calculated statistic out with `$`

```{r}
power_test$power
```


## Power Analysis

Can create power curves by iterating over candidate values for one of the inputs and calculating the remaining.

Example: You are running a study that you believe has a true effect size of $\delta = 0.25$. You want to test your hypothesis with a *t*-test at $\alpha = .05$. What will your power be at varying levels of sample size $N$ per group? 

:::: {.columns}
::: {.column width="50%"}
```{r}
sample_sizes <- c(10, 30, 100, 300)
power_values <- map_dbl(
  sample_sizes, 
  function(sample_size) 
    pwr::pwr.t.test(
      n = sample_size, 
      d = 0.25, 
      sig.level = 0.05)$power
)
```
:::

::: {.column width="50%"}
::: {.panel-tabset}
### Plot
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| fig-align: center
tibble(n = sample_sizes,
       power = power_values) |>
  ggplot(aes(x = n, y = power)) +
  geom_line() +
  coord_cartesian(ylim = c(0, 1))
```

### Code
```{r}
#| eval: false
tibble(n = sample_sizes,
       power = power_values) |>
  ggplot(aes(x = n, y = power)) +
  geom_line()
```
:::

:::
::::


## Power Curves

:::: {.columns}
::: {.column width="50%"}

::: {.code-stack}
```{.r code-line-numbers="2-3,9"}
sample_sizes <- c(10, 30, 100, 300)
power_values <- map_dbl(
  sample_sizes, 
  function(sample_size) 
    pwr::pwr.t.test(
      n = sample_size, 
      d = 0.25, 
      sig.level = 0.05)$power
)
```

::: {.fragment fragment-index="0" .code-overlay}
```{.r code-line-numbers="4"}
sample_sizes <- c(10, 30, 100, 300)
power_values <- map_dbl(
  sample_sizes, 
  function(sample_size) 
    pwr::pwr.t.test(
      n = sample_size, 
      d = 0.25, 
      sig.level = 0.05)$power
)
```
:::

::: {.fragment fragment-index="1" .code-overlay}
```{.r code-line-numbers="5-8"}
sample_sizes <- c(10, 30, 100, 300)
power_values <- map_dbl(
  sample_sizes, 
  function(sample_size) 
    pwr::pwr.t.test(
      n = sample_size, 
      d = 0.25, 
      sig.level = 0.05)$power
)
```
:::

:::
:::

::: {.column width="50%"}
**`map_dbl()`**: iterate over every item in `sample_sizes`, do something to it, and return a `double` (a number with decimal points).

::: {.fragment fragment-index="0"}
**`function(sample_size)`**: defines a function — this tells `map_dbl()` what we want to *do* to each `sample_size`.
:::

::: {.fragment fragment-index="1"}
**`pwr::pwr.t.test(...)$power`**: the same function we were using before.
:::
:::
::::

```{=html}
<style>
.code-stack { position: relative; }
.code-overlay {
  position: absolute;
  inset: 0;
  background-color: var(--r-background-color, #fff);
}
</style>
```


## Power Analysis

Add more candidate $N$ values with `seq(min, max, by)` and clean up the plot

:::: {.columns}
::: {.column width="33%"}
```{r}
seq(1, 10, by = 1)
```

:::

::: {.column width="33%"}
```{r}
seq(1, 10, by = 2)
```
:::

::: {.column width="33%"}
```{r}
seq(1, 10, by = 5)
```
:::
::::

:::: {.columns}
::: {.column width="50%"}
```{r}
sample_sizes <- seq(10, 300, by = 20)
power_values <- map_dbl(
  sample_sizes, 
  function(sample_size) 
    pwr::pwr.t.test(
      n = sample_size, 
      d = 0.25, 
      sig.level = 0.05)$power
)
```
:::

::: {.column width="50%"}
::: {.panel-tabset}
### Plot
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| fig-align: center
tibble(n = sample_sizes,
       power = power_values) |>
  ggplot(aes(x = n, y = power)) +
  geom_hline(aes(yintercept = 0.8), linetype = "dashed", color = "gray75") +
  geom_line() +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic(base_size = 14)
```

### Code
```{r}
#| eval: false
tibble(n = sample_sizes,
       power = power_values) |>
  ggplot(aes(x = n, y = power)) +
  geom_hline(aes(yintercept = 0.8), linetype = "dashed", color = "gray75") +
  geom_line() +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic(base_size = 14)
```
:::

:::
::::


## Power Curves

:::: {.columns}
::: {.column width="50%"}
```{r}
sample_sizes <- c(10, 30, 100, 300)
power_values <- map_dbl(
  sample_sizes, 
  function(sample_size) 
    pwr::pwr.t.test(
      n = sample_size, 
      d = 0.25, 
      sig.level = 0.05)$power
)
```
:::

::: {.column width="50%"}

:::
::::


## Power Curves: Bonus

Even better, do it all in a data frame

:::: {.columns}
::: {.column width="50%"}
```{r}
power_values_df <- tibble(
  sample_sizes = seq(10, 500, by = 20)
) |>
  mutate(
    power = map_dbl(
      sample_sizes, 
      function(sample_size) 
        pwr::pwr.t.test(
          n = sample_size, 
          d = 0.25, 
          sig.level = 0.05)$power
    )
  )
```

:::

::: {.column width="50%"}
::: {.panel-tabset}
### Plot
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| fig-align: center
power_values_df |>
  ggplot(aes(x = sample_sizes, y = power)) +
  geom_line() +
  geom_hline(aes(yintercept = 0.8), linetype = "dashed", color = "gray75") +
  geom_line() +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic(base_size = 14)
```

### Code
```{r}
#| eval: false
power_values_df |>
  ggplot(aes(x = sample_sizes, y = power)) +
  geom_line() +
  geom_hline(aes(yintercept = 0.8), linetype = "dashed", color = "gray75") +
  geom_line() +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic(base_size = 14)
```
:::

:::
::::


## Power Curves: Bonus

Let's you easily add more conditions...

:::: {.columns}
::: {.column width="50%"}
```{r}
power_values_df <- tibble(
  sample_sizes = seq(10, 500, by = 20)
) |>
  expand_grid(d = c(0.2, 0.25, 0.3, 0.4)) |>
  mutate(
    power = map2_dbl(
      sample_sizes, d,
      function(sample_size, d) 
        pwr::pwr.t.test(
          n = sample_size, 
          d = d, 
          sig.level = 0.05)$power
    )
  )
```

:::

::: {.column width="50%"}
::: {.panel-tabset}
### Plot
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| fig-align: center
power_values_df |>
  mutate(d = factor(d)) |>
  ggplot(aes(x = sample_sizes, y = power, color = d, group = d)) +
  geom_line() +
  geom_hline(aes(yintercept = 0.8), linetype = "dashed", color = "gray75") +
  geom_line() +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic(base_size = 14)
```

### Code
```{r}
#| eval: false
power_values_df |>
  mutate(d = factor(d)) |>
  ggplot(aes(x = sample_sizes, y = power, color = d, group = d)) +
  geom_line() +
  geom_hline(aes(yintercept = 0.8), linetype = "dashed", color = "gray75") +
  geom_line() +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic(base_size = 14)
```
:::

:::
::::




<!-- ## Confidence Intervals for $d$ -->

<!-- $$ CI = [d - t_{crit} \times SE, \ d + t_{crit} \times SE] $$ -->


<!-- <!-- :::: {.columns} -->
<!-- <!-- ::: {.column width="50%"} -->
<!-- <!-- ::: -->

<!-- <!-- ::: {.column width="50%"} -->
<!-- <!-- $$ SE = s_\text{pooled} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}} $$ -->
<!-- <!-- ::: -->
<!-- <!-- :::: -->

<!-- Since we transform to *z*-units, $s_\text{pooled} = 1$  -->

<!-- <!-- If $s_\text{pooled} = 1$, $CI = [d - t_{crit} \times \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, \ d + t_{crit} \times \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}]$ --> 


<!-- ```{r} -->
<!-- (t_crit <- qt(c(0.025, 0.975), df = nrow(mpg) - 1)) -->
<!-- ``` -->


<!-- :::: {.columns} -->
<!-- ::: {.column width="25%"} -->
<!-- #### One Sample -->

<!-- $$ SE = \frac{s_d}{\sqrt{N}} = \frac{1}{\sqrt{N}} $$ -->


<!-- ::: {style="font-size:85%} -->
<!-- ```{r} -->
<!-- t_crit <- qt(c(0.025, 0.975), df = nrow(mpg) - 1) -->
<!-- d_one_sample + t_crit * (1 / sqrt(nrow(mpg))) -->
<!-- ``` -->
<!-- ::: -->
<!-- ::: -->

<!-- ::: {.column width="50%"} -->
<!-- #### Independent Samples -->

<!-- $$ d = \frac{\bar{x}_1 - \bar{x}_2}{s_p} $$ -->

<!-- ::: {style="font-size:85%} -->
<!-- ```{r} -->
<!-- d_independent_samples + t_crit * sqrt(1/n_four + 1/n_eight) -->
<!-- ``` -->
<!-- ::: -->


<!-- ::: -->

<!-- ::: {.column width="25%"} -->
<!-- #### Paired Sample -->

<!-- $$ d = \frac{\bar{D}}{s_D} $$ -->

<!-- ::: {style="font-size:85%} -->
<!-- ```{r} -->
<!-- d_paired_sample + t_crit * (sqrt(1 / length(differences))) -->
<!-- ``` -->
<!-- ::: -->

<!-- ::: -->
<!-- :::: -->





























