---
title: "Chi-Squared Tests"
subtitle: "PSYC 6019-A01 / PSYC 6022-A01 <br> 2025-09-18 | Lab 4"
author: "Jess Helmer"
footer: "[jhelmer3.github.io/PSYC6019-2026](https://jhelmer3.github.io/PSYC6019-2026/)"
lang: en
format: 
  revealjs:
    code-link: true
    slide-number: true
    code-line-numbers: false
    theme: ../../gt-theme.scss
    width: 1200
execute: 
  echo: true
  dpi: 600
  freeze: auto
knitr:
  opts_chunk:
    dev: svg
---

## Outline

- $\chi^2$ tests

  - $\chi^2$ Test of Goodness of Fit
  
  - $\chi^2$ Test of Independence
  
- Working with count data

```{r setup}
#| echo: false
#| output: false
library(tidyverse)
library(gt)
```


## Chi-Squared ($\chi^2$) Test

Testing inferences about proportions

[○ Involve categorical variables]{.subpoint}

Comparing observed frequencies (counts) to some expected (null hypothesis) frequencies

**$\chi^2$ Test of Goodness of Fit**: expected frequencies are set by researcher (e.g., null hypothesis is equal frequencies across all groups)

**$\chi^2$ Test of Independence**: expected frequencies are those implied by independence


## Chi-Squared ($\chi^2$) Test

Instead of a continuous variable, our outcome is a **count**

:::: {.columns}
::: {.column width="50%"}
**Continuous**

* Height

* Petal length

* Test score

:::

::: {.column width="50%"}
**Counts**

* Coin flip being heads

* Number of pets

* Goals scored

Only integers

:::
::::



# $\chi^2$ Test of Goodness of Fit


## $\chi^2$ Test of Goodness of Fit

Do the observed frequencies of a categorical variable differ from what expected *a priori*?

Typically, this expectation is equal occurrence across groups (but doesn't have to be).

If equal occurrence, what would be our expected frequencies?

:::: {.columns}
::: {.column .fragment width="45%"}
**100 coin flips**

| Heads           | Tails           |
|-----------------|-----------------|
| [50]{.fragment} | [50]{.fragment} |
:::

::: {.column .fragment width="55%"}
**Favorite primary color of 33 students**

| [Red]{.red}     | [Blue]{.blue}   | [Yellow]{.yellow} |
|-----------------|-----------------|-------------------|
| [11]{.fragment} | [11]{.fragment} | [11]{.fragment}   |
:::
::::

[Can also hypothesize proportions and convert to frequencies once you have a sample size.]{.fragment}


## $\chi^2$ Test of Goodness of Fit: Hypotheses

$H_0$: Observed data match the expected frequencies for the population

$H_1$: Observed data do not match the expected frequencies for the population

[○ Observed frequency is significantly different than expected]{.subpoint}

::: fragment
$H_0$: $\pi_j = \pi_{j_0}$  for all categories $j$ (i.e., difference for all categories is 0), where

[$\pi_j$ = observed proportion]{.subpoint}

[$\pi_{j_0}$ = expected proportion]{.subpoint}

$H_1$: $\pi_j \neq \pi_{j_0}$ for any category $j$
:::


## $\chi^2$ Test of Goodness of Fit

$$ \chi^2 = \sum_{j=1}^{J} \frac{(O_j – N\pi_{j_0})^2}{N\pi_{j_0}} = \sum_{j=1}^{J} \frac{(O_j – E_j)^2}{E_j} $$

where 

[$O_j$ = observed frequency]{.subpoint}

[$N$ = total sample size]{.subpoint}

[$E_j$ = expected frequency]{.subpoint}

[$j$ = individual category out of $J$ total categories]{.subpoint}

[*The sum of the squared differences between the observed and expected frequencies, divided by the expected frequency, for each group.*]{.fragment}


## $\chi^2$ Test of Goodness of Fit

$\text{df} = J - 1$, where

[$J$ = number of categories]{.subpoint}

Use to identify critical $\chi^2$ value from R

For $\chi^2$, as $\text{df}$ increases, so does the critical-$\chi^2$ value (at same $\alpha$)

<!-- Only ever a one-tailed test! For the upper end of the distribution. -->


## $\chi^2$ Distribution

```{r}
#| echo: false
#| fig-align: center
tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 35)) +
  coord_cartesian(clip = "off", xlim = c(1, 35), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16) 
```


## $\chi^2$ Distribution

```{r}
#| echo: false
#| fig-align: center
tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  geomtextpath::geom_textline(aes(y = chi1, label = "chi^{2}*'(df=1)'"), parse = T,
                              hjust = .125, size = 6, linewidth = 1.5, color = "skyblue",
                              padding = unit(15, "points")) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 35)) +
  coord_cartesian(clip = "off", xlim = c(1, 35), ylim = c(0, 0.3), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16) 
```


## $\chi^2$ Distribution

```{r}
#| echo: false
#| fig-align: center
tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  geomtextpath::geom_textline(aes(y = chi1, label = "chi^{2}*'(df=1)'"), parse = T,
                              hjust = .125, size = 6, linewidth = 1.5, color = "skyblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi2, label = "chi^{2}*'(df=2)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue",
                              padding = unit(15, "points")) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 35)) +
  coord_cartesian(clip = "off", xlim = c(1, 35), ylim = c(0, 0.3), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16) 
```


## $\chi^2$ Distribution

```{r}
#| echo: false
#| fig-align: center
tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  geomtextpath::geom_textline(aes(y = chi1, label = "chi^{2}*'(df=1)'"), parse = T,
                              hjust = .125, size = 6, linewidth = 1.5, color = "skyblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi2, label = "chi^{2}*'(df=2)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi4, label = "chi^{2}*'(df=4)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue4",
                              padding = unit(15, "points")) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 35)) +
  coord_cartesian(clip = "off", xlim = c(1, 35), ylim = c(0, 0.3), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16) 
```


## $\chi^2$ Distribution

```{r}
#| echo: false
#| fig-align: center
tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  geomtextpath::geom_textline(aes(y = chi1, label = "chi^{2}*'(df=1)'"), parse = T,
                              hjust = .125, size = 6, linewidth = 1.5, color = "skyblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi2, label = "chi^{2}*'(df=2)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi4, label = "chi^{2}*'(df=4)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue4",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi8, label = "chi^{2}*'(df=8)'"), parse = T,
                              hjust = .435, size = 6, linewidth = 1.5, color = "royalblue4",
                              padding = unit(15, "points")) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 35)) +
  coord_cartesian(clip = "off", xlim = c(1, 35), ylim = c(0, 0.3), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16) 
```


## $\chi^2$ Distribution

```{r}
#| echo: false
#| fig-align: center
tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  geomtextpath::geom_textline(aes(y = chi1, label = "chi^{2}*'(df=1)'"), parse = T,
                              hjust = .125, size = 6, linewidth = 1.5, color = "skyblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi2, label = "chi^{2}*'(df=2)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi4, label = "chi^{2}*'(df=4)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue4",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi8, label = "chi^{2}*'(df=8)'"), parse = T,
                              hjust = .435, size = 6, linewidth = 1.5, color = "royalblue4",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi16, label = "chi^{2}*'(df=16)'"), parse = T,
                              hjust = .65, size = 6, linewidth = 1.5, color = "navy",
                              padding = unit(15, "points")) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 35)) +
  coord_cartesian(clip = "off", xlim = c(1, 35), ylim = c(0, 0.3), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16) 
```


<!-- ## $\chi^2$ Test of Goodness of Fit Cutoffs -->

<!-- :::: {.columns} -->
<!-- ::: {.column width="50%"} -->
<!-- Cutoff for test of heads vs. tails -->

<!-- <br> -->

<!-- ::: fragment -->
<!-- $J =$ 2 -->

<!-- $\text{df} = J - 1 =$ 1 -->
<!-- ::: -->

<!-- ::: fragment -->
<!-- ```{r} -->
<!-- qchisq(.95, 1) -->

<!-- qchisq(.05, 1, lower.tail = F) -->
<!-- ``` -->
<!-- ::: -->
<!-- ::: -->

<!-- ::: {.column width="50%"} -->
<!-- Cutoff for test of favorite color in the rainbow -->

<!-- ::: fragment -->
<!-- $J =$ 7 (ROYGBIV) -->

<!-- $\text{df} = J - 1 =$ 6 -->
<!-- ::: -->

<!-- ::: fragment -->
<!-- ```{r} -->
<!-- qchisq(.95, 6) -->

<!-- qchisq(.05, 6, lower.tail = F) -->
<!-- ``` -->
<!-- ::: -->
<!-- ::: -->
<!-- :::: -->


## $\chi^2$ Test of Goodness of Fit Example

::: {style="font-size:80%"}
You are an education psychologist interested in college students' choice of majors. You've grouped majors into five categories: STEM, Social Sciences, Humanities, Arts, and Business. You'd like to test whether there are differences in the population of students choosing those majors. You sample 200 students and ask what major they chose.

[First, what are your expected frequencies for each category?]{.fragment}
:::

```{r}
#| echo: false
tibble(type = c("Expected", "Observed"),
       STEM = c(NA, NA),
       `Social Sciences` = c(NA, NA),
       Humanities = c(NA, NA),
       Arts = c(NA, NA), 
       Business = c(NA, NA)) |>
  gt(rowname_col = "type") |>
  cols_label(`Social Sciences` = "Social Sciences") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```


## $\chi^2$ Test of Goodness of Fit Example

::: {style="font-size:80%"}
You are an education psychologist interested in college students' choice of majors. You've grouped majors into five categories: STEM, Social Sciences, Humanities, Arts, and Business. You'd like to test whether there are differences in the population of students choosing those majors. You sample 200 students and ask what major they chose.

First, what are your expected frequencies for each category?
:::

```{r}
#| echo: false
tibble(type = c("Expected", "Observed"),
       STEM = c(40, NA),
       `Social Sciences` = c(40, NA),
       Humanities = c(40, NA),
       Arts = c(40, NA), 
       Business = c(40, NA)) |>
  gt(rowname_col = "type") |>
  cols_label(`Social Sciences` = "Social Sciences") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_body(rows = "Expected"))) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```


## $\chi^2$ Test of Goodness of Fit Example

::: {style="font-size:80%"}
You are an education psychologist interested in college students' choice of majors. You've grouped majors into five categories: STEM, Social Sciences, Humanities, Arts, and Business. You'd like to test whether there are differences in the population of students choosing those majors. You sample 200 students and ask what major they chose.

[Next, what is your $\chi^2$ cutoff value?]{.fragment}
:::

```{r}
#| echo: false
tibble(type = c("Expected", "Observed"),
       STEM = c(40, NA),
       `Social Sciences` = c(40, NA),
       Humanities = c(40, NA),
       Arts = c(40, NA), 
       Business = c(40, NA)) |>
  gt(rowname_col = "type") |>
  cols_label(`Social Sciences` = "Social Sciences") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_body(rows = "Expected"))) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```


## $\chi^2$ Test of Goodness of Fit Example

Cutoff in R

::: fragment
```{r}
df <- 5 - 1
chisq_crit <- qchisq(.95, df)
chisq_crit
```
:::


## $\chi^2$ Distribution

```{r}
#| echo: false
#| fig-align: center
tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  geomtextpath::geom_textline(aes(y = chi1, label = "chi^{2}*'(df=1)'"), parse = T,
                              hjust = .125, size = 6, linewidth = 1.5, color = "skyblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi2, label = "chi^{2}*'(df=2)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi4, label = "chi^{2}*'(df=4)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue4",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi8, label = "chi^{2}*'(df=8)'"), parse = T,
                              hjust = .435, size = 6, linewidth = 1.5, color = "royalblue4",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi16, label = "chi^{2}*'(df=16)'"), parse = T,
                              hjust = .65, size = 6, linewidth = 1.5, color = "navy",
                              padding = unit(15, "points")) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 35)) +
  coord_cartesian(clip = "off", xlim = c(1, 35), ylim = c(0, 0.3), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16) 
```


##  $\chi^2(df \ = \ 4)$ Rejection Region

```{r}
#| echo: false
#| fig-align: center
## Claude-generated plot 
## if you can't tell by all the comments

chisq_crit <- qchisq(.95, 4)

tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  # shaded rejection region (under df=4 curve, right of critical value)
  geom_area(data = ~ filter(.x, i >= chisq_crit),
            aes(y = chi4), fill = "steelblue4", alpha = 0.3) +
  # critical value line
  geom_vline(xintercept = chisq_crit, linetype = "dashed",
             color = "steelblue4", linewidth = 0.8) +
  annotate("text", x = chisq_crit, y = 0.29,
           label = paste0("chi[crit]^2 == ", round(chisq_crit, 2)),
           parse = TRUE, hjust = -0.1, size = 5, color = "steelblue4") +
  # faded background curves
  geomtextpath::geom_textline(aes(y = chi1, label = "chi^{2}*'(df=1)'"), parse = T,
                              hjust = .125, size = 6, linewidth = 1.5, color = alpha("grey80", .5),
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi2, label = "chi^{2}*'(df=2)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = alpha("grey80", .5),
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi8, label = "chi^{2}*'(df=8)'"), parse = T,
                              hjust = .435, size = 6, linewidth = 1.5, color = alpha("grey80", .5),
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi16, label = "chi^{2}*'(df=16)'"), parse = T,
                              hjust = .65, size = 6, linewidth = 1.5, color = alpha("grey80", .5),
                              padding = unit(15, "points")) +
  # the star of the show
  geomtextpath::geom_textline(aes(y = chi4, label = "chi^{2}*'(df=4)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue4",
                              padding = unit(15, "points")) +
  scale_x_continuous(breaks = c(1, chisq_crit, 20, 30, 35),
                     labels = c("1", "9.49", "20", "30", "35")) +
  coord_cartesian(clip = "off", xlim = c(1, 35), ylim = c(0, 0.3), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16)
```


## $\chi^2$ Test of Goodness of Fit Example

::: {style="font-size:80%"}
You are an education psychologist interested in college students' choice of majors. You've grouped majors into five categories: STEM, Social Sciences, Humanities, Arts, and Business. You'd like to test whether there are differences in the population of students choosing those majors. You sample 200 students and ask what major they chose.

[Now, what are the observed values (after collecting data)?]{.fragment}
:::

```{r}
#| echo: false
tibble(type = c("Expected", "Observed"),
       STEM = c(40, NA),
       `Social Sciences` = c(40, NA),
       Humanities = c(40, NA),
       Arts = c(40, NA), 
       Business = c(40, NA)) |>
  gt(rowname_col = "type") |>
  cols_label(`Social Sciences` = "Social Sciences") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_body(rows = "Expected"))) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```



## $\chi^2$-test in R

Let's get some observed values!

Most of the time, we'd get data that looks like this:

```{r}
major_data <- read_csv(here::here("lab", "data", "lab_04", "major_data.csv"),
                       show_col_types = F)

major_data |> head(10)
```

And does not show the frequencies yet


## Count Data: `table()`

Some ways to generate frequencies from categorical variables

```{r}
table(major_data$major)
```


<!-- ## Count Data: `summarize()` -->

<!-- Some ways to generate frequencies from categorical variables -->

<!-- :::: {.columns} -->
<!-- ::: {.column width="50%"} -->
<!-- ```{r} -->
<!-- major_data |> -->
<!--   summarize(.by = c(major), -->
<!--             n = n()) -->
<!-- ``` -->
<!-- ::: -->

<!-- ::: {.column width="50%"} -->
<!-- ```{r} -->
<!-- major_data |> -->
<!--   summarize(.by = c(major), -->
<!--             n = n()) |> -->
<!--   arrange(desc(n)) -->
<!-- ``` -->
<!-- ::: -->
<!-- :::: -->


## Count Data: `count()`

`count(df, grouping_var1, grouping_var2, etc.)`

[○ `df` = dataframe]{.subpoint}

<div class="subpoint">○ `grouping_vars` = variable names to group by</div>

Basically shorthand for:

```{r}
#| eval: false
df |>
  summarise(.by = grouping_var,
            n = n())
```

Also seems to `arrange` by default


## Count Data: `count()`

`count(df, grouping_var1, grouping_var2, etc.)`

[○ `df` = dataframe]{.subpoint}

<div class="subpoint">○ `grouping_vars` = variable names to group by</div>

::: {.columns}
::: {.column width="50%"}
```{r}
major_data |>
  summarize(.by = major,
            n = n())
```
:::

::: {.column width="50%"}
```{r}
major_data |>
  count(major)
```
:::
::::


## $\chi^2$ Test of Goodness of Fit Example

::: {style="font-size:80%"}
You are an education psychologist interested in college students' choice of majors. You've grouped majors into five categories: STEM, Social Sciences, Humanities, Arts, and Business. You'd like to test whether there are differences in the population of students choosing those majors. You sample 200 students and ask what major they chose.

Here are the observed values:
:::

```{r}
#| echo: false
counted_major_dat <- major_data |>
  count(major)

tibble(type = c("Expected", "Observed"),
       STEM = c(40, counted_major_dat[counted_major_dat$major == "STEM", "n"][[1]]),
       `Social Sciences` = c(40, counted_major_dat[counted_major_dat$major == "Social Sciences", "n"][[1]]),
       Humanities = c(40, counted_major_dat[counted_major_dat$major == "Humanities", "n"][[1]]),
       Arts = c(40, counted_major_dat[counted_major_dat$major == "Arts", "n"][[1]]), 
       Business = c(40, counted_major_dat[counted_major_dat$major == "Business", "n"][[1]])) |>
  gt(rowname_col = "type") |>
  cols_label(`Social Sciences` = "Social Sciences") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_body(rows = "Expected"))) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```


## $\chi^2$ Test of Goodness of Fit Example

$$ \chi^2 = \sum_{j=0}^{J} \frac{(O_j – N\pi_j)^2}{N\pi_j} = \sum_{j=0}^{J} \frac{(O_j – E_j)^2}{E_j} $$

:::: {.columns style="font-size:80%"}
::: {.column .fragment fragment-index=1 width="70%"}
```{r}
#| echo: false
counted_major_dat <- major_data |>
  count(major)

tibble(type = c("Expected", "Observed"),
       STEM = c(40, counted_major_dat[counted_major_dat$major == "STEM", "n"][[1]]),
       `Social Sciences` = c(40, counted_major_dat[counted_major_dat$major == "Social Sciences", "n"][[1]]),
       Humanities = c(40, counted_major_dat[counted_major_dat$major == "Humanities", "n"][[1]]),
       Arts = c(40, counted_major_dat[counted_major_dat$major == "Arts", "n"][[1]]), 
       Business = c(40, counted_major_dat[counted_major_dat$major == "Business", "n"][[1]])) |>
  gt(rowname_col = "type") |>
  cols_label(`Social Sciences` = "Social Sciences") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_body(rows = "Expected"))) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```

[We can reject the null in favor of the alternative that the number of students is not the same across majors.]{.fragment fragment-index=3}
:::

::: {.column .fragment fragment-index=2 width="30%"}
```{r}
obs_chisq <- (51 - 40)^2 / 40 +
  (33 - 40)^2 / 40 +
  (32 - 40)^2 / 40 +
  (32 - 40)^2 / 40 +
  (52 - 40)^2 / 40
obs_chisq

obs_chisq > chisq_crit

pchisq(obs_chisq, df = df)
1 - pchisq(obs_chisq, df = df) # test p-value
```
:::
::::


## R Function

R has a function for $\chi^2$-tests, `chisq.test(x)`. This function wants counted data, either a numeric vector of counts, or a `table` of the data.

:::: {.columns}
::: {.column .fragment width="33%"}
```{r}
counted_major_data <- major_data |>
  count(major)
counted_major_data
```
:::

::: {.column .fragment width="66%"}
```{r}
tabled_major_data <- major_data |>
  select(major) |>
  table()
tabled_major_data
```
:::
::::

:::: {.columns}
::: {.column .fragment width="40%"}
```{r}
chisq <- counted_major_data |>
  select(-major) |>
  chisq.test()
chisq
```

:::

::: {.column .fragment width="60%"}
```{r}
chisq <- tabled_major_data |>
  chisq.test()
chisq
```

:::
::::


## R Function

:::: {.columns .fragment }
::: {.column width="70%"}
```{r}
chisq$expected
```
:::

::: {.column width="15%"}
```{r}
chisq$statistic
```
:::

::: {.column width="15%"}
```{r}
chisq$p.value
```
:::
::::


:::: {.columns}
::: {.column .fragment width="50%"}
A pretend Quarto document like this:

~~~~markdown
---
title: Chi-Squared Test
author: Jess
---

```{{r}}
chisq <- counted_major_data |>
  select(-major) |>
  chisq.test()
```

The $\chi^2$ statistic was `{{r}} chisq$statistic` with 
a p-value of `{{r}} round(chisq$p.value, 2)`.
~~~~
:::

::: {.column .fragment width="50%"}
Will render the text like this:


The $\chi^2$ statistic was `{r} chisq$statistic` with a p-value of `{r} round(chisq$p.value, 2)`.

[This is called "in-line code".]{.fragment}
:::
::::



## $\chi^2$ Test of Goodness of Fit Example

::: {style="font-size:80%"}
Let's restart and say that you were a educational psychologist interested in how students in STEM schools choose their major. In this situation, you had expected more students to select a STEM major than other majors.

These are your new expected frequencies (60% STEM, evenly divided across the rest):
:::

```{r}
#| echo: false
tibble(type = c("Expected", "Observed"),
       STEM = c(120, counted_major_dat[counted_major_dat$major == "STEM", "n"][[1]]),
       `Social Sciences` = c(20, counted_major_dat[counted_major_dat$major == "Social Sciences", "n"][[1]]),
       Humanities = c(20, counted_major_dat[counted_major_dat$major == "Humanities", "n"][[1]]),
       Arts = c(20, counted_major_dat[counted_major_dat$major == "Arts", "n"][[1]]), 
       Business = c(20, counted_major_dat[counted_major_dat$major == "Business", "n"][[1]])) |>
  gt(rowname_col = "type") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_body(rows = "Expected"))) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```


## R Function

:::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: false
chisq.test(x = c(51, 33, 32, 32, 52),
           p = c(120, 20, 20, 20, 20),
           rescale.p = T)

chisq.test(x = c(51, 33, 32, 32, 52),
           p = c(.60, .10, .10, .10, .10))
```
:::

::: {.column width="50%"}
[○ `x` = data of frequencies]{.subpoint}

[○ `p` = expected probabilities]{.subpoint}

[○ `rescale.p` = [`T`/`F`], will rescale `p` to probabilities if you prefer to input frequencies]{.subpoint}

::: fragment
We can verify that it calculated the expected frequencies correctly:

```{r}
chisq.test(x = c(51, 33, 32, 32, 52),
           p = c(.60, .10, .10, .10, .10))$expected
```
:::
:::
::::



# $\chi^2$ Test of Independence


## $\chi^2$ Test of Independence

Do the observed frequencies of a categorical variable differ from what is expected via properties of independence?

[○ Now considering more than one type of category]{.subpoint}

[○ E.g., Number of pets by cat vs. dog people and undergrad vs. post-graduate]{.subpoint}

::: {style="font-size:80%"}
|               | Cat Person | Dog Person |
|---------------|------------|------------|
| Undergrad     |            |            | 
| Post-Graduate |            |            |
:::

Expected frequencies calculated from the marginal totals of the contingency table

[○ Expected values determined from data]{.subpoint}


## $\chi^2$ Test of Independence: Hypotheses

$H_0$: Categorical Variable 1 and Categorical Variable 2 are independent (i.e., have no association)

$H_1$: Categorical Variable 1 and Categorical Variable 2 are not independent (i.e., have some association)

[○ Do expected frequencies match those expected via independence?]{.subpoint}

::: fragment
Mathematically the same as before: 

$H_0$: $\pi_j = \pi_{j_0}$  for all categories $j$; difference for all categories is 0, where

[$\pi_j$ = observed proportion]{.subpoint}

[$\pi_{j_0}$ = expected proportion]{.subpoint}

$H_1$: $\pi_j \neq \pi_{j_0}$ for any category $j$
:::

::: {.my-callout .fragment .absolute bottom=0 right=0 width="400px"}
Just changing the expected frequencies
:::


## $\chi^2$ Test of Independence

$$ \chi^2 = \sum_{j_1=1}^{J_1}\sum_{j_2=1}^{J_2} \frac{(O_j – N\pi_{j_0})^2}{N\pi_{j_0}} = \sum_{j_1=1}^{J_1}\sum_{j_2=1}^{J_2} \frac{(O_j – E_j)^2}{E_j} $$

Basically identical formula to calculate the $\chi^2$.

The difference is that our expected frequencies $E_j$ are derived from the data. Just like before, we calculate the difference from each cell, but now we have two categories that make up the cell.

[*The sum of the squared differences between the observed and expected frequencies, divided by the expected frequency, for each cell.*]{.fragment}


<!-- ## $\chi^2$ Test of Independence Cutoffs -->

<!-- What is the critical $\chi^2$ value in a test of independence for a 3x2? -->

<!-- <br> -->

<!-- ::: fragment -->
<!-- $\text{df} = (J_1 - 1) \times (J_2 - 1)$ or -->

<!-- $\text{df} = (n_{rows} - 1) \times (n_{cols} - 1)$ -->
<!-- ::: -->

<!-- <br> -->

<!-- :::: {.columns} -->
<!-- ::: {.column .fragment width="50%"} -->
<!-- ```{r} -->
<!-- df <- (3 - 1) * (2 - 1) -->
<!-- df -->
<!-- qchisq(.95, df) -->
<!-- ``` -->
<!-- ::: -->

<!-- ::: {.column .fragment width="50%"} -->
<!-- ```{r} -->
<!-- pchisq(6, df) -->
<!-- 1 - pchisq(6, df) -->
<!-- ``` -->
<!-- ::: -->
<!-- :::: -->


## $\chi^2$ Test of Independence Example

::: {style="font-size:80%"}
Let's say that you were a educational psychologist interested in how students choose STEM or non-STEM schools based on whether or not they had a parent go to a STEM or non-STEM school. You sample 200 students.
:::

```{r}
#| echo: false
major_data_wide <- major_data |>
  count(school_type, parent_in_stem) |>
  pivot_wider(names_from = school_type, values_from = n)

major_data_wide |>
  gt(rowname_col = "parent_in_stem") |>
  cols_label(`Non-STEM School` = "Non-Stem School",
             `STEM School` = "STEM School",
             parent_in_stem = "Parent in STEM") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  sub_values(values = c(59, 56, 32, 53), replacement = " ") |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```

:::: {.columns}
::: {.column .fragment width="50%"}
First, what is your critical $\chi^2$ value?

::: fragment
$\text{df} = (J_1 - 1) \times (J_2 - 1)$ or

$\text{df} = (n_{rows} - 1) \times (n_{cols} - 1)$
:::
:::

::: {.column .fragment width="50%"}
```{r}
df <- (2 - 1) * (2 - 1)
qchisq(0.95, df)
```
:::
::::


## $\chi^2$ Distribution

```{r}
#| echo: false
#| fig-align: center
tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  geomtextpath::geom_textline(aes(y = chi1, label = "chi^{2}*'(df=1)'"), parse = T,
                              hjust = .125, size = 6, linewidth = 1.5, color = "skyblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi2, label = "chi^{2}*'(df=2)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi4, label = "chi^{2}*'(df=4)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = "steelblue4",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi8, label = "chi^{2}*'(df=8)'"), parse = T,
                              hjust = .435, size = 6, linewidth = 1.5, color = "royalblue4",
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi16, label = "chi^{2}*'(df=16)'"), parse = T,
                              hjust = .65, size = 6, linewidth = 1.5, color = "navy",
                              padding = unit(15, "points")) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 35)) +
  coord_cartesian(clip = "off", xlim = c(1, 35), ylim = c(0, 0.3), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16) 
```


##  $\chi^2(df \ = \ 1)$ Rejection Region

```{r}
#| echo: false
#| fig-align: center
## Claude-generated plot 
## if you can't tell by all the comments

chisq_crit <- qchisq(.95, 1)  # ~3.84

tibble(i = seq(1, 35, length.out = 1000)) |>
  mutate(chi1 = dchisq(i, 1),
         chi2 = dchisq(i, 2),
         chi4 = dchisq(i, 4),
         chi8 = dchisq(i, 8),
         chi16 = dchisq(i, 16)) |>
  ggplot(aes(x = i)) +
  # shaded rejection region (under df=1 curve, right of critical value)
  geom_area(data = ~ filter(.x, i >= chisq_crit),
            aes(y = chi1), fill = "skyblue", alpha = 0.3) +
  # critical value line
  geom_vline(xintercept = chisq_crit, linetype = "dashed",
             color = "skyblue3", linewidth = 0.8) +
  annotate("text", x = chisq_crit, y = 0.29,
           label = paste0("chi[crit]^2 == ", round(chisq_crit, 2)),
           parse = TRUE, hjust = -0.1, size = 5, color = "skyblue3") +
  # faded background curves
  geomtextpath::geom_textline(aes(y = chi2, label = "chi^{2}*'(df=2)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = alpha("grey80", .5),
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi4, label = "chi^{2}*'(df=4)'"), parse = T,
                              hjust = .3, size = 6, linewidth = 1.5, color = alpha("grey80", .5),
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi8, label = "chi^{2}*'(df=8)'"), parse = T,
                              hjust = .435, size = 6, linewidth = 1.5, color = alpha("grey80", .5),
                              padding = unit(15, "points")) +
  geomtextpath::geom_textline(aes(y = chi16, label = "chi^{2}*'(df=16)'"), parse = T,
                              hjust = .65, size = 6, linewidth = 1.5, color = alpha("grey80", .5),
                              padding = unit(15, "points")) +
  # the star of the show
  geomtextpath::geom_textline(aes(y = chi1, label = "chi^{2}*'(df=1)'"), parse = T,
                              hjust = .125, size = 6, linewidth = 1.5, color = "skyblue3",
                              padding = unit(15, "points")) +
  scale_x_continuous(breaks = c(1, chisq_crit, 10, 20, 30, 35),
                     labels = c("1", "3.84", "10", "20", "30", "35")) +
  coord_cartesian(clip = "off", xlim = c(1, 35), ylim = c(0, 0.3), expand = F) +
  guides(y = "none", x = guide_axis(cap = T)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 16)
```


## $\chi^2$ Test of Independence Example

::: {style="font-size:80%"}
Let's say that you were a educational psychologist interested in how students choose STEM or non-STEM schools based on whether or not they had a parent go to a STEM or non-STEM school. You sample 200 students.
:::

```{r}
#| echo: false
major_data_wide <- major_data |>
  count(school_type, parent_in_stem) |>
  pivot_wider(names_from = school_type, values_from = n)

major_data_wide |>
  gt(rowname_col = "parent_in_stem") |>
  cols_label(`Non-STEM School` = "Non-Stem School",
             `STEM School` = "STEM School",
             parent_in_stem = "Parent in STEM") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  sub_values(values = c(59, 56, 32, 53), replacement = " ") |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```

[Next, let's get our observed values!]{.fragment}


## $\chi^2$ Test of Independence Observed Values

Let's look at creating frequency data for more than one variable

```{r}
major_data |> head(10)
```


## Count Data: `table()`

```{r}
table(major_data$parent_in_stem, major_data$school_type)
```


## Count Data: `count()`

`count(df, grouping_var1, grouping_var2, etc.)`

[○ `df` = dataframe]{.subpoint}

<div class="subpoint">○ `grouping_vars` = variable names to group by</div>

::: {.columns}
::: {.column width="50%"}
```{r}
major_data |>
  summarize(.by = c(parent_in_stem, school_type),
            n = n())
```
:::

::: {.column width="50%"}
```{r}
major_data |>
  count(parent_in_stem, school_type)
```
:::
::::


## $\chi^2$ Test of Independence Example

::: {style="font-size:80%"}
Let's say that you were a educational psychologist interested in how students choose STEM or non-STEM schools based on whether or not they had a parent go to a STEM or non-STEM school. You sample 200 students, and these are the frequencies you observe.
:::

```{r}
#| echo: false
major_data_wide <- major_data |>
  count(school_type, parent_in_stem) |>
  pivot_wider(names_from = school_type, values_from = n)

major_data_wide |>
  gt(rowname_col = "parent_in_stem") |>
  cols_label(`Non-STEM School` = "Non-Stem School",
             `STEM School` = "STEM School",
             parent_in_stem = "Parent in STEM") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```

::: {.fragment style="font-size:80%"}
First, what are your expected frequencies for each category?
:::


## Calculating Expected Frequencies

First, get the marginal totals for each column

```{r}
#| echo: false
major_data_wide |>
  mutate(Total = NA) |>
  add_row(tibble(parent_in_stem = "Total", `Non-STEM School`= NA, `STEM School` = NA, Total = NA)) |>
  gt(rowname_col = "parent_in_stem") |>
  cols_label(`Non-STEM School` = "Non-Stem School",
             `STEM School` = "STEM School") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_column_labels(columns = "Total"),
                             cells_body(columns = "Total"),
                             # cells_body(rows = in_state >= 30),
                             cells_stub(rows = "Total")))
```

:::: {.columns}
::: {.column .fragment width="50%"}
```{r}
major_data |> 
  count(parent_in_stem)
```
:::

::: {.column .fragment width="50%"}
```{r}
major_data |> 
  count(school_type)
```
:::
::::


## Calculating Expected Frequencies

First, get the marginal totals for each column

```{r}
#| echo: false
major_data_wide |>
  mutate(Total = `Non-STEM School` + `STEM School`) |>
  add_row(tibble(parent_in_stem = "Total", `Non-STEM School`= 59+56, `STEM School` = 32+53, Total = 91+109)) |>
  gt(rowname_col = "parent_in_stem") |>
  cols_label(`Non-STEM School` = "Non-Stem School",
             `STEM School` = "STEM School") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_column_labels(columns = "Total"),
                             cells_body(columns = "Total"),
                             cells_body(rows = `Non-STEM School` >= 100),
                             cells_stub(rows = "Total")))
```



## Calculating Expected Frequencies

:::: {.columns}
::: {.column width="50%"}
```{r}
#| echo: false
major_data_wide |>
  mutate(Total = `Non-STEM School` + `STEM School`) |>
  add_row(tibble(parent_in_stem = "Total", `Non-STEM School`= 59+56, `STEM School` = 32+53, Total = 91+109)) |>
  gt(rowname_col = "parent_in_stem") |>
  cols_label(`Non-STEM School` = "Non-Stem School",
             `STEM School` = "STEM School") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_column_labels(columns = "Total"),
                             cells_body(columns = "Total"),
                             cells_body(rows = `Non-STEM School` >= 100),
                             cells_stub(rows = "Total")))
```
:::

::: {.column .fragment width="50%"}
```{r}
#| echo: false
tibble(parent_in_stem = c("No", "Yes", "Total"),
       `Non-STEM School` = c(91*115/200, 109*115/200, 59+56) |> round(1),
       `STEM School` = c(91*85/200, 109*85/200, 109) |> round(1),
       Total = c(115, 85, 91+109)) |>
  gt(rowname_col = "parent_in_stem") |>
  cols_label(`Non-STEM School` = "Non-Stem School",
             `STEM School` = "STEM School") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_column_labels(),
                             cells_body(),
                             cells_stub())) 
```
:::
::::

Expected cell frequency = $E_j$ = $\frac{\text{column total} \times \text{row total}}{N}$

:::: {.columns}
::: {.column width="50%"}
[○ Parent non-STEM & non-STEM school = $\frac{91 \times 115}{200} = 52.3$]{.subpoint-narrow}

[○ Parent STEM & non-STEM school = $\frac{109 \times 115}{200} = 62.7$]{.subpoint-narrow}
:::

::: {.column width="50%"}
[○ Parent non-STEM & STEM school = $\frac{91 \times 85}{200} = 38.7$]{.subpoint-narrow}

[○ Parent STEM & STEM school = $\frac{109 \times 85}{200} = 46.3$]{.subpoint-narrow}
:::
::::


## $\chi^2$ Test of Independence

$$ \chi^2 = \sum_{j_1=1}^{J_1}\sum_{j_2=1}^{J_2} \frac{(O_j – N\pi_{j_0})^2}{N\pi_{j_0}} = \sum_{j_1=1}^{J_1}\sum_{j_2=1}^{J_2} \frac{(O_j – E_j)^2}{E_j} $$

:::: {.columns}
::: {.column width="50%"}
::: fragment
```{r}
#| echo: false
major_data_wide |>
  gt(rowname_col = "parent_in_stem") |>
  cols_label(`Non-STEM School` = "Non-Stem School",
             `STEM School` = "STEM School",
             parent_in_stem = "Parent in STEM") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26)
```
:::

::: fragment
```{r}
#| echo: false
tibble(parent_in_stem = c("No", "Yes"),
       `Non-STEM School` = c(91*115/200, 109*115/200) |> round(1),
       `STEM School` = c(91*85/200, 109*85/200) |> round(1)) |>
  gt(rowname_col = "parent_in_stem") |>
  cols_label(`Non-STEM School` = "Non-Stem School",
             `STEM School` = "STEM School") |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = 26) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = list(cells_column_labels(), cells_stub())) |>
  tab_style(style = cell_text(color = "steelblue3", style = "italic"),
            locations = list(cells_column_labels(),
                             cells_body(),
                             cells_stub())) 
```
:::
:::

::: {.column .fragment width="50%"}
```{r}
#| code-line-numbers: false
obs_chisq <- (59 - 52.3)^2 / 52.3 +
  (32 - 38.7)^2 / 38.7 +
  (56 - 62.7)^2 / 62.7 +
  (53 - 46.3)^2 / 46.3
obs_chisq

obs_chisq > chisq_crit
1 - pchisq(obs_chisq, # test p-value
          df = (2 - 1) * (2 - 1)) 
```
:::
::::


<!-- ## Count Data: `summarize()` -->

<!-- Some ways to generate frequencies from categorical variables -->

<!-- :::: {.columns} -->
<!-- ::: {.column width="50%"} -->
<!-- ```{r} -->
<!-- major_data |> -->
<!--   summarize(.by = c(parent_in_stem, school_type), -->
<!--             n = n()) -->
<!-- ``` -->
<!-- ::: -->

<!-- ::: {.column width="50%"} -->
<!-- ```{r} -->
<!-- major_data |> -->
<!--   summarize(.by = c(parent_in_stem, school_type), -->
<!--             n = n()) |> -->
<!--   arrange(parent_in_stem, school_type) -->
<!-- ``` -->
<!-- ::: -->
<!-- :::: -->


## R Function

If going to put into the `chisq.test()` function, want to pivot **wider** and get rid of additional columns.

Many ways to do this, but one example:

:::: {.columns style="font-size:90%"}
::: {.column .fragment width="50%"}
```{r}
head(major_data, 2)

chisqdat <- major_data |>
  count(parent_in_stem, school_type) |>
  pivot_wider(names_from = parent_in_stem,
              values_from = n) |>
  select(-school_type)

chisqdat
```
:::

::: {.column .fragment width="50%"}
```{r}
chisq_test <- chisq.test(chisqdat)
chisq_test$expected
chisq_test
```

[We retain the null hypothesis that students' school type (STEM vs. non-STEM) and whether they choose a STEM or non-STEM major are independent.]{.fragment}
:::
::::




## Visualizing Count Data: `geom_bar()`

::: {.panel-tabset}
### Plot
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
major_data |>
  ggplot(aes(x = major)) +
  geom_bar() +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 15, vjust = .7))
```

### Code
```{r}
#| eval: false
major_data |>
  ggplot(aes(x = major)) +
  geom_bar() +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 15, vjust = .7))
```
:::

## Working With Count Data: `geom_bar()`

::: {.panel-tabset}
### Plot
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
major_data |>
  ggplot(aes(y = major)) +
  geom_bar() +
  theme_classic(base_size = 14)
```

### Code
```{r}
#| eval: false
major_data |>
  ggplot(aes(y = major)) +
  geom_bar() +
  theme_classic(base_size = 14)
```
:::


## Working With Count Data: `geom_bar()`

::: {.panel-tabset}
### Plot
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
major_data |>
  ggplot(aes(y = major, fill = factor(parent_in_stem))) +
  geom_bar(position = "dodge") +
  theme_classic(base_size = 14)
```

### Code
```{r}
#| eval: false
major_data |>
  ggplot(aes(y = major, fill = factor(parent_in_stem))) +
  geom_bar(position = "dodge") +
  theme_classic(base_size = 14)
```
:::


## Lab 04 Activity
