---
title: "Effect Size and Power Analysis"
subtitle: "PSYC 6019-A01 / PSYC 6022-A01 <br> 2025-09-25 | Lab 5"
author: "Jess Helmer"
footer: "[jhelmer3.github.io/PSYC6019-2026](https://jhelmer3.github.io/PSYC6019-2026/)"
format: 
  revealjs:
    code-link: true
    slide-number: true
    code-line-numbers: false
    theme: ../../gt-theme.scss
    width: 1200
execute: 
  echo: true
  dpi: 600
  freeze: auto
---

## Outline

- Effect size
- Power analysis

```{r setup}
#| echo: false
#| output: false
library(tidyverse)
library(gt)
```


## Cohen's *D*

Standardized mean difference, interpretable effect size

Derived with two means (represented as $\mu_1$ and $\mu_2$) and standard deviation ($s$)

$$d = \frac{\mu_1 - \mu_2}{s}$$

For a one-sample *t* test, $\mu_1$ would be our sample mean $\bar{x}$, $\mu_2$ would be our population mean $\mu$, and the standard deviation $\sigma$ would be our observed standard deviation $s$.

Not a statistical test---just a measure of difference.
## Cohen's *D* Example

Let's say a researcher claims the average highway miles per gallon across all cars is 30mpg. We found earlier that the sample had a significant difference between the hypothesized and sample mean. What is the effect size of that difference?

$$d = \frac{\bar{x} - \mu}{s}$$

::: fragment
```{r}
x_bar <- mean(mpg$hwy)
x_sd <- sd(mpg$hwy)

d <- (x_bar - 30) / x_sd
d
```
:::

[How would be characterize this difference? What is the direction?]{.fragment}


## Cohen's *D* Strength Visualization

Based on mean ($\bar{x}$ = `round(x_bar, 2)` = `r round(x_bar, 2)`) and SD ($\bar{\sigma}$ = `round(x_sd, 2)` = `r round(x_sd, 2)`) of cars' highway MPG (`hwy`).

Solved backwards for ($\mu$) based on different potential levels of *d*. "What would the hypothesized mean have to have been to get different levels of Cohen's *d*?"

$$\bar{x} - d \times s = \mu$$

```{r}
dat <- data.frame(x = rnorm(100000, x_bar, x_sd))

mu_ex1 <- x_bar - .2 * x_sd
mu_ex2 <- x_bar - .5 * x_sd
mu_ex3 <- x_bar - .8 * x_sd
```


## Cohen's *D* Strength Visualization

"What would the hypothesized mean have to have been to get different levels of Cohen's *d*?"

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| fig-align: center
ggplot(dat, aes(x = x)) +
  geom_histogram(fill = 'skyblue3', alpha = .4, bins = 16) +
  geom_vline(aes(xintercept = x_bar),
             color = "skyblue3", linewidth = 1, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 60), expand = F, clip = "off") +
  guides(y = "none", x = NULL) +
  labs(y = NULL, x = "Highway MPG") +
  theme_classic(base_size = 18)
```


## Cohen's *D* Strength Visualization

"What would the hypothesized mean have to have been to get different levels of Cohen's *d*?"


```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| layout-ncol: 3
#| fig-align: center
ggplot(dat, aes(x = x)) +
  geom_histogram(fill = 'skyblue3', alpha = .4, bins = 16) +
  geom_vline(aes(xintercept = x_bar),
             color = "skyblue3", linewidth = 1, linetype = "dashed") +
  geomtextpath::geom_textvline(aes(xintercept = mu_ex1, label = "d = 0.2"),
                               hjust = .3, size = 10, color = "steelblue4", linewidth = 1.5) +
  coord_cartesian(xlim = c(0, 60), expand = F, clip = "off") +
  guides(y = "none", x = NULL) +
  labs(y = NULL, x = "Highway MPG") +
  theme_classic(base_size = 18)

ggplot(dat, aes(x = x)) +
  geom_histogram(fill = 'skyblue3', alpha = .4, bins = 16) +
  geom_vline(aes(xintercept = x_bar),
             color = "skyblue3", linewidth = 1, linetype = "dashed") +
  geomtextpath::geom_textvline(aes(xintercept = mu_ex2, label = "d = 0.5"),
                               hjust = .3, size = 10, color = "steelblue4", linewidth = 1.5) +
  coord_cartesian(xlim = c(0, 60), expand = F, clip = "off") +
  guides(y = "none", x = NULL) +
  labs(y = NULL, x = "Highway MPG") +
  theme_classic(base_size = 18)

ggplot(dat, aes(x = x)) +
  geom_histogram(fill = 'skyblue3', alpha = .4, bins = 16) +
  geom_vline(aes(xintercept = x_bar),
             color = "skyblue3", linewidth = 1, linetype = "dashed") +
  geomtextpath::geom_textvline(aes(xintercept = mu_ex3, label = "d = 0.8"),
                               hjust = .3, size = 10, color = "steelblue4", linewidth = 1.5) +
  coord_cartesian(xlim = c(0, 60), expand = F, clip = "off") +
  guides(y = "none", x = NULL) +
  labs(y = NULL, x = "Highway MPG") +
  theme_classic(base_size = 18)

```


## Cohen's *D* in R: One Sample

For one sample, simplest to just program it ourselves `(x_bar - mu) / x_sd`

Example: effect size of observed iris petal length against a hypothesized population mean of 3.

```{r}
(mean(iris$Petal.Length) - 3) / sd(iris$Petal.Length)
```


## Cohen's *D*

The `cohens_d()` function from the `effectsize` package allows for many formats of inputs.

:::: {.columns}
::: {.column width="50%"}
```{r}
#| code-line-numbers: false
# install.packages("effectsize")

(x_bar_vanilla - x_bar_cookie_dough) / pooled_sd

effectsize::cohens_d(sales ~ candy,
                     data = candy)
```
:::

::: {.column width="50%"}
```{r}
#| code-line-numbers: false
mean_diff / sd_diff

effectsize::cohens_d(pumpkin$oct1, pumpkin$oct31,
                     paired = T)
```
:::
::::

