---
title: "Simulations and Chi-Squared Tests"
subtitle: "PSYC 6019-A01 / PSYC 6022-A01 \n 2025-09-11 | Lab 3"
author: "Jess Helmer"
footer: "[jhelmer3.github.io/PSYC6019-2026/](https://jhelmer3.github.io/PSYC6019-2026/)"
format: 
  revealjs:
    code-link: true
    code-line-numbers: false
    theme: ../../gt-theme.scss
execute: 
  echo: true
  dpi: 600
---

## Outline

- Simulating data introduction
- Central Limit Theorem example
- $\chi^2$ distribution
  
```{r setup}
#| echo: false
#| output: false
# targets::tar_load(store = here::here("_targets"))
```


## Simulation

why do we like sims


## Simulation Example: CLT

The Central Limit Theorem (CLT) says that the sampling distribution of the mean of any distribution with finite mean and variance will converge to be normal with increasing sample size.

::: subpoint
○ Imagine: take a sample of $N$ observations from the population, calculate some statistics (e.g., the mean), and toss them back.
:::

::: subpoint
○ Repeat a whole bunch of times
:::

::: subpoint
○ Sampling distribution: the distribution of those calculated statistics (e.g., means)
:::

Let's say we're not so sure, and we want to test this out with different distributions and sample sizes $N$.


## Simulation Example: CLT

Before simulating, define your goals / questions.

:::: {.columns}
::: {.column width="50%"}
### Questions

Does the sampling distribution of the mean really converge to normal for wonky asymmetrical distributions like the $\chi^2$ distribution?

At what sample sizes does this fail?
:::

::: {.column width="50%"}
### Goals

See if the distribution of means from repeatedly sampling data looks normal

::: subpoint-narrow
○ Does this change with different original distributions from which the data is drawn?
:::

::: subpoint-narrow
○ Does this change with sample size? 
:::

:::
::::


## Simulation Example: CLT

Then, sketch out a plan. Try to build things up step-by-step, not all at once.

1. Start with data sampled from a normal distribution. Generate one sample of data, take its mean.

2. Repeat this process multiple times, generating replications of data samples and sample means.

3. Visualize both: the original distribution of the data (sample distribution), the distribution of the the samples' means (sampling distribution).

4. Incorporate different distributions.

5. Incorporate different sample sizes.


## Simulation Example: CLT

1. Start with data sampled from a normal distribution. Generate one sample of data, take its mean.


:::: {.columns style="font-size:80%"}
::: {.column width="50%"}
`rnorm(x, mean, sd)` where `x` is the number of samples

```{r}
#| message: false
library(tidyverse)
samples <- rnorm(3, mean = 0, sd = 1)
samples
mean(samples)
```
:::

::: {.column width="50%"}
Instead of having it in a vector by itself, lets put it in a dataframe

```{r}
sim_data <- tibble(x = rnorm(3, mean = 0, sd = 1))
sim_data
mean(sim_data$x)
```

:::
::::

::: {.callout-note style="font-size:80%"}
### `tibble()`
A `tibble()` is just like a dataframe, but better. The best part is its printing---`tibble()`s will never crowd your screen by trying to show every single value of a dataframe. Also, they work better with making `tibble`s in `tibble`s in `tibble`s... [More on `tibble`s here](https://tibble.tidyverse.org/articles/tibble.html).
:::


## Simulation Example: CLT

2. Repeat this process multiple times, generating replications of data samples and sample means.

:::: {.columns style="font-size:80%"}
::: {.column width="50%"}
Let's specify the sample size and how many replications we want. Start small!

```{r}
n_reps <- 2
sample_size <- 3
# total number of simulated draws
total_n <- n_reps * sample_size
```
:::

::: {.column width="50%"}
Let's create a `tibble` with a row for every index in every replication.
```{r}
sim_data <- tibble(
  n_reps = n_reps,
  sample_size = sample_size
) |>
  uncount(n_reps, .id = "rep") |>
  uncount(sample_size, .id = "index")
sim_data
```

:::
::::


## Simulation Example: CLT

2. Repeat this process multiple times, generating replications of data samples and sample means.

:::: {.columns style="font-size:80%"}
::: {.column width="50%"}
We can then draw `n_reps * sample_size = total_n` values from a normal distribution with `rnorm()`.

```{r}
sim_data <- tibble(
  n_reps = n_reps,
  sample_size = sample_size
) |>
  uncount(n_reps, .id = "rep") |>
  uncount(sample_size, .id = "index") |>
  mutate(
    draw = rnorm(total_n, mean = 0, sd = 1)
  )
sim_data
```
:::

::: {.column width="50%"}
And, we can plot those draws (getting a head start on #3).

```{r}
#| fig-width: 5
#| fig-height: 3
sim_data |>
  ggplot(aes(x = draw)) +
  geom_density()
```

:::
::::


## Simulation Example: CLT

2. Repeat this process multiple times, generating replications of data samples and sample means.

3. Visualize both: the original distribution of the data (sample distribution), the distribution of the the samples' means (sampling distribution).

:::: {.columns style="font-size:80%"}
::: {.column width="50%"}
Let' finish out #2 by `summarize()`ing our `sim_data` into one sample mean per replication.

```{r}
means <- sim_data |>
  summarize(.by = rep,
            sample_mean = mean(draw))
means
```
:::

::: {.column width="50%"}
And, we can plot those draws (getting a head start on #4).

```{r}
#| fig-width: 5
#| fig-height: 3
means |>
  ggplot(aes(x = sample_mean)) +
  geom_density()
```
:::
::::


## Simulation Example: CLT

Before we start incorporating other conditions, let's scale this up.

:::: {.columns style="font-size:80%"}
::: {.column width="50%"}
We can then draw `n_reps * sample_size = total_n` values from a normal distribution with `rnorm()`.

```{r}
n_reps <- 1000
sample_size <- 10
total_n <- n_reps * sample_size

sim_data <- tibble(
  n_reps = n_reps,
  sample_size = sample_size
) |>
  uncount(n_reps, .id = "rep") |>
  uncount(sample_size, .id = "index") |>
  mutate(
    draw = rnorm(total_n, mean = 0, sd = 1)
  )
sim_data
```
:::

::: {.column width="50%"}
And grab their means.

```{r}
means <- sim_data |>
  summarize(.by = rep,
            sample_mean = mean(draw))
means
```
:::
::::


## Simulation Example: CLT

Plotting both the sample distribution and sampling distributions...

:::: {.columns}
::: {.column width="50%"}
::: {.panel-tabset}
### Plot
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-align: center
#| layout-ncol: 1

sim_data |>
  ggplot(aes(x = draw)) +
  geom_density() +
  theme_classic(base_size = 14) +
  coord_cartesian(xlim = c(-4, 4)) +
  guides(x = guide_axis(cap = T),
         y = guide_axis(cap = T),
         color = guide_none()) +
  labs(title = "Sample Distribution")
```

### Code
```{r}
#| eval: false
sim_data |>
  ggplot(aes(x = draw)) +
  geom_density() +
  theme_classic(base_size = 14) +
  coord_cartesian(xlim = c(-4, 4)) +
  guides(x = guide_axis(cap = T),
         y = guide_axis(cap = T),
         color = guide_none()) +
  labs(title = "Sample Distribution")
```
:::
:::

::: {.column width="50%"}
::: {.panel-tabset}
### Plot
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 3
#| fig-align: center
means |>
  ggplot(aes(x = sample_mean)) +
  geom_density() +
  theme_classic(base_size = 14) +
  coord_cartesian(xlim = c(-4, 4)) +
  guides(x = guide_axis(cap = T),
         y = guide_axis(cap = T),
         color = guide_none()) +
  labs(title = "Sampling Distribution")
```

### Code
```{r}
#| eval: false
means |>
  ggplot(aes(x = sample_mean)) +
  geom_density() +
  theme_classic(base_size = 14) +
  coord_cartesian(xlim = c(-4, 4)) +
  guides(x = guide_axis(cap = T),
         y = guide_axis(cap = T),
         color = guide_none()) +
  labs(title = "Sampling Distribution")
```
:::
:::

::::


















