---
title: "Lab 01 Activity Key"
author: "[Your Name]"
date: "[Today's Date]"
---

### Activity Objectives:

- Do some basic R!

- Get familiar with Quarto

### Instructions:

- Fill in the code chunks

- For each code chunk, provide some accompanying text

- Render the Quarto document and submit both the `.qmd` and `.html` file on Canvas


## Activity

### 1) Display the first few rows of the iris dataset.

```{r}
head(iris)
```


### 2) Calculate the mean and SD for the Sepal Width and Sepal Length variables. How do they compare to each other? (1--2 sentences).

```{r}
mean(iris$Sepal.Width)
mean(iris$Sepal.Length)
sd(iris$Sepal.Width)
sd(iris$Sepal.Length)
```

The mean and variance of Sepal Length are both larger than those of Sepal Width.

### 3) What kinds of R topics are you hoping to learn this semester? If you have experience coding in R, are there any challenges you've encountered with it?


### 4) The Quarto documentation has a [Gallery Page](https://quarto.org/docs/gallery/). Visit the link, and look through the neat things people have made with Quarto! For whichever is your favorite, paste the link below. Also, check out its source code by visiting its accompanying `</>` "View Source Code" link.








