---
title: "Lecture 7: More Chi-Square"
author: "Mark Himmelstein"
format: 
  revealjs:
    slide-width: 1920
    slide-height: 1080
    theme: ../../../gt-theme.scss
    slide-number: true
    slide-number-format: "c/t"
    html-math-method: mathjax
    include-in-header:
      text: |
        <script>
        window.MathJax = {
          tex: {
            packages: {'[+]': ['cancel']}
          },
          loader: {
            load: ['[tex]/cancel']
          }
        };
        </script>
params:
  v0: false 
editor: visual
---

```{r}
#| echo: false


library(knitr)
library(formatR)
library(tidyverse)
library(gganimate)
library(gifski)
opts_chunk$set(out.width = '100%', options(width=60))

opts_knit$set(global.par = TRUE)
opts_chunk$set(echo=FALSE, warning = FALSE, message = FALSE, comment = '', class.output='sh',
               fig.width=8, fig.height=5, fig.align = "center")

my_red <- "#E41A1C"
my_blue <- "#377EB8"
my_green <- "#4DAF4A"
my_purple <- "#984EA3"
my_orange <- "#FF7F00"
my_yellow <- "#FFFF33"
my_red2 <- "#FBB4AE" 
my_blue2 <- "#B3CDE3" 
my_green2 <- "#CCEBC5" 
my_purple2 <- "#DECBE4"
my_orange2 <- "#FED9A6"
my_yellow2 <- "#FFFFCC"


bball <- read.csv("../../data/2026NBA_data.csv") |> filter(G >= max(G)/2,
                                                draft_year != "")

```

## Last Time

-   We learned the conceptual foundation of null hypothesis significance testing (NHST)
-   We saw two examples
    -   $Z$-test
        -   Test of a hypothesis about a continuous variable
        -   Were our sampled NBA players consistent with a population that scored 12 points per game?
    -   $\chi^2$ test (aka $\chi^2$ "goodness of fit" test)
        -   Were our sampled gummy bears consistent with equal categories, or consistent with the more raspberries hypothesis?

## Setting Up Hypotheses

-   Set up a null hypothesis $H_0$

    -   $H_0: \mu_{PPG} = 12$
    -   $H_0: \pi_{raspberry} = \pi_{pineapple} = \pi_{orange} = \pi_{strawberry} = \pi_{lemon} = \frac{1}{5}$
        -   $\pi$ = notation for parameter representing probability in the population

-   Set up alternative hypothesis $H_a$

    -   $H_0: \mu_{PPG} \neq 12$
    -   $H_0: \text{At least one } \pi_k \neq \frac{1}{5}$
        -   $\pi$ = notation for parameter representing probability in the population

## Conducting Test

-   Construct sampling distribution
-   Calculate test statistic
-   Determine how unlikely test statistic is for sampling distribution
-   If $P(\text{statistic}) < .05$, reject null hypothesis.
    -   For a "two-tailed" test (e.g. a $Z$-test) we multiply by two to account for possibility that under the proposed sampling distribution the statistic was equally likely to have been above or below $\mu$

## More Tests

-   The $Z$-test allows us to test a specific hypothesis about a continuous variable, but **assumes we know the population value of** $\sigma$
    -   Major limitation! But good for building a conceptual foundation
-   The $\chi^2$ goodness of fit test allows us to test a hypothesis about the probability of each category for a categorical variable in the population
    -   E.g. hypothesis about how common each flavor of gummy bears are

In this module we're going to discuss more things you can do with a $\chi^2$ sampling distribution:

-   The $\chi^2$ test of independence for two categorical variables

## Contingency Tables

What if we have more than one categorical variable?

We can arrange these data in a **contingency table**.

*Example: Race and death penalty sentences*\
*Data from Unah & Boger (2001), death-eligible cases in North Carolina from 1993 to 1997, white victim.*

|                      | Death Sentence |         |           |
|:---------------------|:--------------:|:-------:|:---------:|
| **Defendant's Race** |    **Yes**     | **No**  | **Total** |
| Nonwhite             |       33       |   251   |  **284**  |
| White                |       33       |   508   |  **541**  |
| **Total**            |     **66**     | **759** |  **825**  |

What might we want to compare in this table?

## Contingency Tables

::: {style="font-size: 0.825em;"}
Contingency tables give information about the joint, marginal, and conditional distributions of the observed variables

|                      | Death Sentence |         |           |
|:---------------------|:--------------:|:-------:|:---------:|
| **Defendant's Race** |    **Yes**     | **No**  | **Total** |
| Nonwhite             |       33       |   251   |  **284**  |
| White                |       33       |   508   |  **541**  |
| **Total**            |     **66**     | **759** |  **825**  |

-   $P(\text{white} \cap \text{death sentence}) =$ [$\frac{33}{825} = .04$ (joint)]{.fragment}

-   $P(\text{nonwhite} \cap \text{death sentence}) =$ [$\frac{33}{825} = .04$ (joint)]{.fragment}

-   $P(\text{white}) =$ [$\frac{541}{825} = .66$ (marginal)]{.fragment}

-   $P(\text{nonwhite}) =$ [$\frac{284}{825} = .34$ (marginal)]{.fragment}

-   $P(\text{death sentence} | \text{white}) =$ [$\frac{33}{541} = .06$ (conditional)]{.fragment}

-   $P(\text{death sentence} | \text{nonwhite}) =$ [$\frac{33}{284} = .12$ (conditional)]{.fragment}
:::

## Independence

:::::: {style="font-size: 0.95em;"}
In the $\chi^2$ test of independence, we want to know if two categorical variables are related (dependent on each other)

$$H_0: \text{Categorical variables are independent}$$

$$H_1: \text{Categorical variables are related}$$

How do we define **independence**?

::::: {.content-visible .answer unless-meta="params.v0"}
:::: answer
::: fragment
If variables $X$ and $Y$ are **independent**, then $P(X \cap Y) = P(X)P(Y)$

-   Joint distribution is the product of marginal distributions

-   Similar to flipping two coins:

$$P\left[\text{(flip 1 = heads)} \cap \text{(flip 2 = heads)}\right] = \frac{1}{2}\times\frac{1}{2} = \frac{1}{4}$$
:::
::::
:::::
::::::

## Expected Frequency Calculations

We next need to calculate expected frequencies based on $H_0$

First, a bit of notation: We often index *rows* with a subscript *i*, and *columns* with a subscript *j*

-   $R_i$: category $i$ of row variable $R$

-   $C_j$: category $j$ of column variable $C$

-   $N$: total number of observations

-   $f_{Oi}$: observed marginal frequency for row $i$

-   $f_{Oj}$: observed marginal frequency for column $j$

-   $f_{Eij}$: expected joint frequency for row $i$ and column $j$

## Expected Frequency Calculations

We next need to calculate expected frequencies based on $H_0$

To get expected frequencies, we use the **observed marginal** frequencies to determine the **expected joint** frequencies.

Thus, under the assumption of independence,

$$P(R_i \cap C_j) = P(R_i)P(C_j) = \bigg(\frac{f_{Oi}}{N}\bigg) \bigg(\frac{f_{Oj}}{N}\bigg) = \frac{f_{Oi}f_{Oj}}{N^2}$$ $$f_{Eij} = N\times P(R_i \cap C_j) = N\bigg(\frac{f_{Oi}f_{Oj}}{N^2}\bigg) =  \frac{f_{Oi}f_{Oj}}{N}$$

## [Expected Frequency Calculations]{style="font-size: 0.667em;"}

::: {style="font-size: 0.85em;"}
|                      | Death Sentence |         |           |
|:---------------------|:--------------:|:-------:|:---------:|
| **Defendant's Race** |    **Yes**     | **No**  | **Total** |
| Nonwhite             |       33       |   251   |  **284**  |
| White                |       33       |   508   |  **541**  |
| **Total**            |     **66**     | **759** |  **825**  |

$$f_{Eij} = \frac{f_{Oi}f_{Oj}}{N}$$

-   $f_{E}$ (nonwhite, death) = [$\frac{284 \times 66}{825} = \boldsymbol{22.72}$]{.answer .fragment .content-visible unless-meta="params.v0"}

-   $f_{E}$ (white, death) = [$\frac{541 \times 66}{825} = \boldsymbol{43.28}$]{.answer .fragment .content-visible unless-meta="params.v0"}

-   $f_{E}$ (nonwhite, no death) = [$\frac{284 \times 759}{825} = \boldsymbol{261.28}$]{.answer .fragment .content-visible unless-meta="params.v0"}

-   $f_{E}$ (white, no death) = [$\frac{541 \times 759}{825} = \boldsymbol{497.72}$]{.answer .fragment .content-visible unless-meta="params.v0"}
:::

## $\chi^2$ Test of Independence

The $\chi^2$ test statistic is computed in the same way as for the GoF test

$$\chi^2_{\text{obs}} = \sum_{i=1}^R \sum_{j=1}^C \frac{(f_{Oij} - f_{Eij})^2}{f_{Eij}}$$

-   sum over all cells

::: {style="font-size: 0.85em;"}
$$\frac{(33 - 22.72)^2}{22.72} + \frac{(33 - 43.28)^2}{43.28} + \frac{(251 - 261.28)^2}{261.28} + \frac{(508 - 497.72)^2}{497.72} = 7.71$$
:::

## Degrees of Freedom

Given the marginal counts, how many cell counts can be freely chosen?

|   | $C_1$ | $C_2$ | $C_3$ | **Total** |
|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
| $R_1$ | [**1**]{.fragment style="color: blue;"} | [**4**]{.fragment style="color: blue;"} |  | 10 |
| $R_2$ |  |  |  | 15 |
| **Total** | 5 | 5 | 15 | 25 |

## None Left

Given the marginal counts, how many cell counts can be freely chosen?

|   | $C_1$ | $C_2$ | $C_3$ | **Total** |
|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
| $R_1$ | [**1**]{style="color: blue;"} | [**4**]{style="color: blue;"} | [**5**]{style="color: red;"} | 10 |
| $R_2$ | [**4**]{style="color: red;"} | [**1**]{style="color: red;"} | [**5**]{style="color: red;"} | 15 |
| **Total** | 5 | 5 | 15 | 25 |

The degrees of freedom for the $\chi^2$ test of independence equals $(R-1)(C-1)$

## Our Test

$$df = (R-1)(C-1) = (2-1)(2-1) = 1$$

```{r}
pchisq(7.71, df = 1, lower.tail = FALSE)

```

## Visual of Our Test

```{r}


x <- data.frame(x = c(0, 8.5))

ggplot(x, aes(x)) +
  stat_function(fun = dchisq, args = list(df = 1), color = "black", linewidth = 0.6) +
  stat_function(fun = dchisq, args = list(df = 1), xlim = c(0, qchisq(.95, 1)), geom = "area", fill = my_blue, alpha = 0.8) +
  stat_function(fun = dchisq, args = list(df = 1), xlim = c(qchisq(.95, 1), 8.5), geom = "area", fill = my_red, alpha = 0.8) +
  geom_vline(xintercept = qchisq(.95, 1), color = my_orange, linewidth = 1) +
  geom_text(aes(label = paste0("Critical Value\n(", round(qchisq(.95, 1), 2), ")"), 
                x = qchisq(.95, 1) + 0.3, y = 0.18), size = 4, color = my_orange, hjust = 0) +

  geom_vline(xintercept = 7.71, color = my_purple, linewidth = 1, linetype = "dashed") +
  geom_text(aes(label = "Observed Statistic\n(x = 7.71, p = 0.005)", 
                x = 5, y = 0.5), size = 4, color = my_purple, hjust = 0) +
  geom_segment(aes(x = 6.7, xend = 7.5, y = 0.35, yend = 0.09), 
               color = my_purple, arrow = arrow(length = unit(.1, "inches"))) +
  labs(x = expression(chi^2), y = "density", title = expression(paste("Chi-Square Distribution (", df == 1, ")"))) +
 
  theme_light()
```

## Yates' Correction

::: {style="font-size: 0.8em;"}
With small sample sizes, the $\chi^2$ test of independence may be **overly liberal**

-   We will be more likely to reject the null hypothesis even if it's true than we want

Yates' correction attempts to correct for the "discontinuity" implied by having small frequencies

-   i.e., marginal probabilities that are closer to discrete (0 or 1 exactly) rather than continuous

$$\chi^2_{\text{obs}} = \sum_{i=1}^R \sum_{j=1}^C \frac{(|f_{Oij} - f_{Eij}| - 0.5)^2}{f_{Eij}}$$

Yates' correction is **conservative**

-   We will be more likely to retain the null hypothesis even if it's false than we want

The corrected $p$-value is necessarily higher than the uncorrected $p$-value
:::

## Father and Son

Ordinary $\chi^2$ test developed by Karl Pearson

Egon Pearson is K. Pearson's son, another leader of early statistics!

Same $H_0$ and $H_1$ as Pearson's $\chi^2$ test of independence

Test statistic:

$$\chi^2(df) = \frac{\text{Pearson's } \chi^2 \times N}{N - 1}$$

-   $\chi^2$: test statistic from ordinary test of independence

-   $df$: $(R - 1) \times (C - 1)$

-   $N$ total frequency

## Egon Pearson's Adjusted $\chi^2$ Test

E. Pearson's adjusted $\chi^2$ test statistic is always larger than Pearson's $\chi^2$

$p$-value of E. Pearson's test is always smaller than Pearson's $p$-value

-   We say that E. Pearson's test is **liberal**

-   E. Pearson's test has **higher statistical power** to reject the false $H_0$

$$\chi^2_{\text{KP}} < \chi^2_{\text{EP}}$$

$$p_{_\text{KP}} > p_{_\text{EP}}$$

Slightly better performance when \underline{one} cell has a small expected frequency

## Fisher's Exact Test

::: {style="font-size: 0.9em;"}
Yates' $\chi^2$ and E. Pearson's adjusted $\chi^2$ change the calculation of $\chi^2$. Fisher's exact test uses another method to overcome limitations of the $\chi^2$ test of independence.

1.  Treat the observed marginals as fixed

2.  Find all possible sets of cell counts

3.  Calculate the probability of each set of cell counts (hypergeometric distribution)

4.  Sum the probabilities corresponding to cell count sets that are more extreme (further from $H_0$) than the observed data $\rightarrow$ this is the $p$-value

Limitations:

-   Complicated for tables larger than $2 \times 2$
-   Can be very slow when marginal frequencies are large
:::

## Recommendations by Campbell (2007)

-   For large sample sizes, the standard $\chi^2$ test is best.

-   Default in R uses Yates' correction, which helps with smaller samples and will have minimal cost in larger samples.

-   When at least one cell has expected frequency \< 5, instead consider:

    -   If one cell has expected frequency = 1, choose E. Pearson's adjusted $\chi^2$

    -   If all cells have expected frequency \> 1, choose Fisher's exact test

::: {style="font-size: 0.65em;"}
Campbell, I. (2007). Chi-square and Fisher-Irwin tests of two-by-two tables with small sample recommendations. *Statistics in Medicine*, *26*, 3661-3675.
:::
