---
title: "Lecture 17: Individual and Multiple Comparisons"
author: "Mark Himmelstein"
format: 
  revealjs:
    slide-width: 1920
    slide-height: 1080
    theme: ../../../gt-theme.scss
    slide-number: true
    slide-number-format: "c/t"
    html-math-method: mathjax
    include-in-header:
      text: |
        <script>
        window.MathJax = {
          tex: {
            packages: {'[+]': ['cancel']}
          },
          loader: {
            load: ['[tex]/cancel']
          }
        };
        </script>
params:
  v0: false 
editor: source
---

```{r}
#| echo: false


library(knitr)
library(formatR)
library(tidyverse)
library(readxl)
library(gganimate)
library(gifski)
opts_chunk$set(out.width = '100%', options(width=60))

opts_knit$set(global.par = TRUE)
opts_chunk$set(echo=FALSE, warning = FALSE, message = FALSE, comment = '', class.output='sh',
               fig.width=8, fig.height=5, fig.align = "center")

my_red <- "#E41A1C"
my_blue <- "#377EB8"
my_green <- "#4DAF4A"
my_purple <- "#984EA3"
my_orange <- "#FF7F00"
my_yellow <- "#FFFF33"
my_red2 <- "#FBB4AE" 
my_blue2 <- "#B3CDE3" 
my_green2 <- "#CCEBC5" 
my_purple2 <- "#DECBE4"
my_orange2 <- "#FED9A6"
my_yellow2 <- "#FFFFCC"


bball <- read.csv("../../data/2026NBA_data.csv") |> filter(G >= max(G)/2,
                                                draft_year != "")

```

## Last Time

-   We learned how to use the ANOVA framework to test hypotheses that **all means** among many groups are equal
-   Aka the "omnibus test"


$$H_0: \mu_1 = \mu_2 = \mu_3 = \cdots$$

$$H_1: \text{at least one of the means are different}$$

- Also learned how to structure this procedure as a linear model and do corollary Bayesian analysis with Bayes factors to compare posterior likelihoods of competing models

$$y_{ig} = \color{#377EB8}{\beta_0} + \color{#4DAF4A}{\sum_{p=1}^{g-1}\beta_p x_{p}} + \color{#984EA3}{\epsilon_i}$$

- Use coding schemes like contrast or dummy coding for $\beta$'s so model can represent all group means as parsimoniously as possible

## Specific Comparisons

What if we want to do specific comparisons between individual groups (not all groups at once)? Let's revisit the first example from last time

Example: weight of plants under one of three conditions

- Control (ctrl), 10 observations
- Treatment 1 (trt1), 10 observations
- Treatment 2 (trt2), 10 observations

::: {style="font-size: 0.85em;"}

```{r}
#| echo: true
data("PlantGrowth", package = "datasets")
summary(PlantGrowth)
```

:::
    

## Group Means

```{r}
ggplot(PlantGrowth, aes(group, weight, fill = group)) + 
  geom_boxplot() + theme_light() + 
  scale_fill_manual(values = c(my_purple2, my_green2, my_blue2)) + 
  theme(legend.position = "none") + labs(x = "") + 
  stat_summary(fun=mean, geom="point", shape=7, size=4)
```

::: {.fragment style="text-align: center; margin-top: -10px;"}
Are these means equal **in the population?** Last time we did this by hand, now lets use `R`
:::


## ANOVA in `R`

```{R}
#| echo: true

a1 <- aov(weight ~ group, PlantGrowth)
anova(a1)

```

Tells us there are differences **somewhere**. But where? What if we want to check `treatment1` vs. `control`? What might we do? 


[An independent samples $t$-test, but how will it be differet from a regular independent samples $t$-test?]{.fragment }

[Do we just ignore `treatment2` entirely, or can we use it for something?]{.fragment }

## $t$ Revisited

If we want to compare `treatment1` to `control` in a $t$-test, we have the following formula

$$t = \frac{\bar{x}_{\text{treatment1}}- \bar{x}_{\text{control}}}{\sigma_{\bar{x}_{\text{treatment1}}- \bar{x}_{\text{control}}}}$$

My question: how will I define $\sigma_{\bar{x}_{\text{treatment1}}- \bar{x}_{\text{control}}}$?

I'll give you two options

- Pooled standard error for treatment and control groups

- Use $\sqrt{\text{MSW}}$ for something

Which do you want to use?

## Lets Simulate!

In this simulation, I am going to simulate 1000 samples from three groups, each with $N = 5$:

- $x_1 \sim \text{Normal(-5, 10)}$
- $x_2 \sim \text{Normal(0, 10)}$
- $x_3 \sim \text{Normal(5, 10)}$



Then I am going to calculate

- the pooled SD based on $x_1$ and $x_2$
- $\sqrt{\text{MSW}}$ based on all three groups

## Simulation Result

```{r}

iter <- 1000
N <- 5



sigmas <- data.frame(sigma_paired = rep(NA, iter),
                     sqrt_msw = rep(NA, iter))

set.seed(1024)
for(i in 1:iter){
  
  sim_dat <- data.frame(x1 = rnorm(N, -5, 10),
                      x2 = rnorm(N, 0, 10),
                      x3 = rnorm(N, 5, 10))
  
  
  sim_dat_long <- sim_dat |>  
    pivot_longer(cols = c("x1","x2","x3"), names_to = "group", values_to = "x")
  
  ttest_result <- t.test(sim_dat$x1, sim_dat$x2)
  
  anova_result <- aov(x ~ group, sim_dat_long)
  
  sigmas$sigma_paired[i] <- ttest_result$stderr/sqrt(1/5 + 1/5)

  sigmas$sqrt_msw[i] <- sigma(anova_result)

}

sigmas_long <- sigmas |> pivot_longer(1:2, names_to = "Estimator",
                                      values_to = "Estimate")

ggplot(sigmas_long, aes(x = Estimate, fill = Estimator)) + 
  geom_density(alpha = 0.2) + 
  geom_vline(xintercept = 10, linetype = 2) + 
  annotate(
    "text", 
    x = 10, 
    y = .22, 
    label = "sigma == 10", 
    parse = TRUE, 
    vjust = 2, 
    hjust = -0.1
  ) + 
  xlim(c(0, 20)) +
  scale_fill_discrete(
    labels = c("pooled SD", expression(sqrt(MSW)))
  ) + 
  theme_minimal() + 
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  )

```


## Discussion

Why did $\sqrt{\text{MSW}}$ provide a better estimate (on average) of the population pooled standard deviation than the sample pooled standard deviation for $x_1$ and $x_2$?

- [Does $x_3$ have a different standard deviation than $x_1$ or $x_2$?]{.answer .fragment .content-visible unless-meta="params.v0"}
- [Because we assumed **homogeneity of variance** we want to use **all of our data** to estimate the common standard deviation]{.answer .fragment .content-visible unless-meta="params.v0"}
- [Even if our specific comparison omits a group, that group provides more data to estimate the shared variance]{.answer .fragment .content-visible unless-meta="params.v0"}
- [And $\text{MSW}$ IS an estimate of the shared, within group variance! (all groups)]{.answer .fragment .content-visible unless-meta="params.v0"}

## Using MSW in $t$




Modified $t$ test statistic:

$$t(df) = \frac{\bar{x}_1 - \bar{x}_2}{s_{\bar{x}_1 - \bar{x}_2}}$$

where

$$s_{\bar{x}_1 - \bar{x}_2} = \sqrt{\frac{MSW}{n_1} + \frac{MSW}{n_2}}$$

- $df = dfW$
- $n_1, n_2$: sample sizes for groups 1 and 2



## Back to Plants

```{r}
#| echo: true
PlantGrowth |> group_by(group) |> summarize(weight = mean(weight),
                                            n = n())
anova(a1)

```

## `treatment1` vs `control`

In the `anova()` output, MSW is in the `Residuals` row under `Mean Sq`

- $s_{\bar{x}_1 - \bar{x}_2}  = \sqrt{\frac{MSW}{n_1} + \frac{MSW}{n_2}} = \sqrt{\frac{0.3886}{10} + \frac{0.3886}{10}} = 0.279$


- $t(df) = \frac{\bar{x}_1 - \bar{x}_2}{s_{\bar{x}_1 - \bar{x}_2}} =\frac{5.032 - 4.661}{0.279} = 1.33$


- Since $df = dfW$, we have $t(27) = 1.33$

We can look this up in our `pt()` $t$-table function

```{r}
pt(1.331, 27, lower.tail = FALSE) * 2
```

## 95% CI

How will we get the 95% confidence interval?

:::::: {.answer}
:::: {.content-visible unless-meta="params.v0"}
:::: {.fragment}

-  Find the critical value corresponding to $\alpha$ on our $t$-distribution

```{r}
qt(.05/2, 27, lower.tail = FALSE)
```
Then Calculate $95\% \text{CI} = \bar{x}_1 - \bar{x}_2 \pm t(df)_{\alpha/2} \times s_{\bar{x}_1 - \bar{x}_2} =$ 
$5.032 - 4.661 \pm 2.05 \times 0.279 = [-0.20,.0.94]$
::::
::::
::::::



## Cohen's $d$

How will we get Cohen's $d$?

:::::: {.answer}
:::: {.content-visible unless-meta="params.v0"}
:::: {.fragment}

-  Since MSW is our pooled variance, $\sqrt{\text{MSW}}$ is our pooled SD:

$d = \frac{ \bar{x}_1 - \bar{x}_2}{\sqrt{MSW}} =$
$\frac{5.032 - 4.661}{\sqrt{.389}} = 0.595$
::::
::::
::::::

## Multiple Comparisons


What if we want to make more than one comparison?

::: {.callout-note title="Family-Wise Error (FWE) Rate"}
* The probability of rejecting **at least one** true $H_0$ in a series of comparisons.
* The probability of committing **at least one** Type I error in a series of comparisons.
:::

[xkcd Significant Comic](http://xkcd.com/882/)

- If **no** jelly bean color causes acne, there is a high probability that we will reject $H_0$ for at least one test.

- To avoid high FWE, we test all the hypotheses at the same time with an omnibus test.

::: {.fragment}
- But what if we WANT to test the hypotheses individually?
:::


## Multiple Comparisons

So far, we have only learned about the omnibus test against the null hypothesis:

- Are **all group means** equal in the population?

Or the test of a single comparison

- Does one specific group differ from another

Often, we want to also know whether differences occur among *all* pairs of groups.

Running multiple tests on the same data leads to an **inflated Type I error rate**. If all of the tests are independent:

\[
  \alpha^\star = 1 - (1 - \alpha)^k
\]

- $k$: number of comparisons made or tests done
-  $\alpha$: specified Type I error rate
- $\alpha^\star$: actual Type I error rate (probability of rejecting at least one correct null)

For a study with $G$ groups, there are $k = \frac{G(G-1)}{2}$ pairs of groups. That means we would do $k$ different tests if we want to compare each individual group to each other group.

## Logic of Error Inflation I

Recall, for two independent events:

$$P(A \cap B) = P(A)P(B)$$

- In this case, the independent events are the probability of **correctly retaining** the null. Since the probability of incorrectly rejecting is $\alpha$, the probability of correctly retaining is $1 - \alpha$.

- If we have $k$ independent tests, then the probability of correctly retaining the null **for every single test jointly** is:

## Logic of Error Inflation II

$$P(\text{retain}_1 \cap \text{retain}_2 \cap \text{retain}_3 \cap \dots \cap \text{retain}_k) =$$ 
$$P(\text{retain}_1) \times P(\text{retain}_2) \times \dots \times P(\text{retain}_k) =$$
$$(1-\alpha)_1 \times (1-\alpha)_2 \times \dots \times (1-\alpha)_k =$$
$$(1-\alpha)^k$$

That means that the probability that we **incorrectly** reject **at least one** of those null hypotheses is:

$$1 - (1 - \alpha)^k$$

## Inflated Type I Errors for Jellybeans

[xkcd Significant](http://xkcd.com/882/)

In the jellybean comic, we aren't comparing groups to one another, but we do have $k = 20$ different tests.

- Actual Type I error rate $\alpha^\star = 1 - (1 - .05)^{20} = \mathbf{.64}$ !!!

**Interpretation:** There is a 64% chance we would reject the null hypothesis for **at least one** jellybean color just by chance even if **all null hypotheses were true\***.

_\* And if all of the tests were independent_

## Multiple Comparisons

Traditional Procedures:

- When ANOVA **is not** significant, we **DO NOT** conduct multiple comparisons of mean differences between pairs of groups.
- When ANOVA **is** significant, we might conduct multiple comparisons to examine which pairs of groups contribute to the significant results.

Modern Recommendations:

- Conduct multiple comparisons **regardless** of the significance of the simple ANOVA test.
- Multiple comparisons do not require a significant omnibus test.
- Multiple comparison procedures control the Type I error rate of each comparison.

## Post-Hoc vs. A Priori Comparisons

**Post-hoc** comparisons:

- Mean comparisons between all pairs of groups after data collection.
- Requires extra statistical control for Type I error rates.

**A priori** comparisons:

- Mean comparisons based on planned hypotheses (before data collection).
- Planned hypotheses are made before knowing ANOVA test results.
- Usually fewer comparisons.
    - Higher power than all pairwise comparisons.


## Bonferroni and Holm

::: {style="background-color: #f3e5f5; border-left: 4px solid #984EA3; padding: 10px; margin-bottom: 15px;"}
**Bonferroni Correction:**

$$\text{adjusted } \alpha = \alpha / k$$
:::

::: {style="background-color: #f3e5f5; border-left: 4px solid #984EA3; padding: 10px;"}
**Holm Correction:**

1. Sort $p$-values from low to high.
2. Compare to adjusted $\alpha = \frac{\alpha}{k}, \frac{\alpha}{k-1}, \cdots, \frac{\alpha}{2}, \frac{\alpha}{1}$
:::

- $\alpha$: desired Type I error rate (e.g., .05) | $k$: number of comparisons
- If $p < \text{adjusted } \alpha$: reject $H_0$ for that pair
- Holm's adjustment is more powerful than Bonferroni's




## Other Example


Original Study:

> Williams, L. E., & Bargh, J. A. (2008). Keeping one's distance: The influence of social cues on affect and evaluation. *Psychological Science*, 19, 302-308.

Replication study link: [OSF Link](https://osf.io/p87cn)

Does viewing objects that are close together affect your feelings of closeness toward people and places?




## More Details

::: {style="font-size: .8em;"}

- **Step 1**: subjects drew two points on a coordinate plane
  - **Closeness** prime: (2, 4) and (-3, 1)
  - **Intermediate** prime: (8, 3) and (-6, -5)
  - **Distant** prime: (12, 10) and (-11, -8)
- **Step 2**: subjects rated the strength of their bonds to:
  - siblings
  - parents
  - hometown


**IV: distance prime**

- 3 levels: close, intermediate, distant


**DV: bond ratings** (average rating for siblings, parents, hometown)

- 7-point scale (1 = "not at all strong", 7 = "extremely strong")

:::

## Visual

```{r}


dat <- read_xls("../../data/Williams data.xls")

dat <- dat %>% mutate(average_bond = (SIB_BOND + PAR_BOND + HOME_BOND) / 3) %>%
  filter(!is.na(average_bond)) %>%
  mutate(condition = recode(factor(COND), "1" = "close", "2" = "intermediate", "3" = "distant")) %>% dplyr::select(-COND)

dat2 <- dat %>% 
  pivot_longer(c(SIB_BOND:HOME_BOND, average_bond), names_to = "type", values_to = "strength") %>%
  mutate(type = recode(type, SIB_BOND = "Sibling", PAR_BOND = "Parent", HOME_BOND = "Home"))

ggplot(dat2 %>% filter(type == "average_bond")) + 
  geom_boxplot(aes(x = condition, y = strength, fill = condition)) + 
  scale_fill_manual(values = c(my_purple2, my_green2, my_blue2)) + 
  theme(legend.position = "bottom") + theme_light() + 
  facet_grid(. ~ "Average of Bond Types") + 
  labs(x = "", y = "average bond strength rating") + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) + 
  stat_summary(aes(x = condition, y = strength), fun=mean, geom="point", shape=7, size=4)
```

## Omnibus Result


```{r}
#| echo: true
a1 <- aov(average_bond ~ condition, dat)
anova(a1)
```



## Bonferroni and Holm Corrections in R

::: {style="font-size: 0.8em;"}

```{r}
#| echo: true
library(emmeans)
em1 <- emmeans(a1, specs = "condition") 
pairs(em1, adjust = "bonferroni")
pairs(em1, adjust = "holm")
```

:::

## Tukey's HSD I


::: {style="font-size: 0.85em;"}

Tukey's Honestly Significant Difference (HSD) is a slightly different approach. Null and alternative hypotheses are the same as other pairwise comparisons.

**Test statistic:**

$$q(df, G) = \frac{\bar{x}_1 - \bar{x}_2}{s_{\bar{x}_1 - \bar{x}_2}}$$

where

$$s_{\bar{x}_1 - \bar{x}_2} = \sqrt{\frac{\left(\frac{MSW}{n_1} + \frac{MSW}{n_2}\right)}{2}}$$

- $MSW$: mean of within-group effect (from ANOVA table)
- $df = dfW$
- $q$ follows a studentized range distribution (`ptukey` / `qtukey` in `R`)

:::

## Tukey's HSD II

$(1 - \alpha)\%$ confidence intervals for $(\mu_1 - \mu_2)$:

$$(\bar{x}_1 - \bar{x}_2) \pm q(df)_{\alpha_{\text{adj}}/2} \times s_{\bar{x}_1 - \bar{x}_2}$$

- $df = dfW = N - G$
- $s_{\bar{x}_1 - \bar{x}_2}$: standard error of $(\bar{x}_1 - \bar{x}_2)$
- $\alpha_{\text{adj}}$: $\alpha$ adjusted with either Bonferroni or Holm
- $q(df)_{\alpha_{\text{adj}}/2}$: $(\alpha_{\text{adj}}/2)^{\text{th}}$ quantile of the studentized range distribution with $df$ degrees of freedom

Tukey's HSD is more powerful than Bonferroni.

## Tukey in R

Tukey's HSD is the default for `pairs()` in `emmeans`.

::: {style="font-size: 0.85em;"}

```{r}
#| echo: true
pairs(em1, adjust = "tukey") # can leave out adjust since tukey is default
confint(pairs(em1))
```

:::

## Games-Howell

::: {style="font-size: 0.85em;"}

Same $H_0$ and $H_1$ as Bonferroni/Holm/Tukey.

**Test statistic:**

$$q(df, G) = \frac{\bar{x}_1 - \bar{x}_2}{s_{\bar{x}_1 - \bar{x}_2}}$$

where

$$s_{\bar{x}_1 - \bar{x}_2} = \sqrt{\frac{\left(\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}\right)}{2}}$$

- $s_1^2, s_2^2$: sample variances of groups 1 and 2
- $n_1, n_2$: sample sizes of groups 1 and 2
- $q$ follows the Tukey distribution with $df = dfW$ and $G$ groups

:::

## Games-Howell in R

::: {style="font-size: 0.85em;"}

```{r}
#| echo: true
library(rstatix)
dat %>% 
  ungroup() %>% # needed because we grouped the tibble earlier
  games_howell_test(average_bond ~ condition)
```

:::

## Dunnett C and Dunnett T3

- Neither procedure assumes homogeneity of variance
- Both procedures successfully correct the Type I error rate for multiple comparisons
- **Comparisons among the 3 procedures:**
  - Games-Howell performs better than Dunnett C and Dunnett T3
  - Dunnett C has higher statistical power than Dunnett T3 when $dfW > 75$ (Tabachnick & Fidell, 2001)
  
  

## Recommendations

**Recommended post-hoc procedures:**

- **Homogeneity of variance:** Tukey HSD
- **Heterogeneity of variance:** Games-Howell
- Work with the same assumptions for both the ANOVA itself and for post-hoc tests

## A Priori Comparisons

So far, we have learned about post-hoc tests (exploratory, look at all possible pairs).

**A priori comparisons:**

- Comparisons planned before data collection
- Guided by theory
- Only some group differences may truly be of interest
- Usually fewer comparisons
- More statistically powerful

## Linear Contrasts I

Example (4 groups): $H_1$: $\mu_1$ is different than the average of $\mu_2$, $\mu_3$, and $\mu_4$

Rewrite $H_0$ and $H_1$:

$$\begin{align*}
H_0: & \quad 1\mu_1 + -\frac{1}{3}\mu_2 + -\frac{1}{3}\mu_3 + -\frac{1}{3}\mu_4 = 0 \\
H_1: & \quad 1\mu_1 + -\frac{1}{3}\mu_2 + -\frac{1}{3}\mu_3 + -\frac{1}{3}\mu_4 \neq 0
\end{align*}$$

Based on these hypotheses, we can define a linear contrast ($\psi$):

$$\psi = a_1 \bar{x}_1 + a_2 \bar{x}_2 + a_3 \bar{x}_3 + a_4 \bar{x}_4 = \sum a_g \bar{x}_g$$

- Require $\sum a_g = 0$

## Linear Contrasts II

What would be a linear contrast to compare $\bar{x}_1$ to $\bar{x}_2$?

$$\begin{align*}
H_0: & \quad 1\mu_1 + -1\mu_2 + 0\mu_3 + 0\mu_4 = 0 \\
H_1: & \quad 1\mu_1 + -1\mu_2 + 0\mu_3 + 0\mu_4 \neq 0
\end{align*}$$

## Linear Contrasts III

**For hypothesis testing:**

$$t(df) = \frac{\psi}{s_\psi}$$

- $\psi$: linear contrast $= \sum a_g \bar{x}_g$
- $s_\psi$: standard error of contrast
- Test statistic follows a $t$ distribution with $dfW$ degrees of freedom
- Assumes homogeneity of variance

## Linear Contrasts IV

**Confidence interval:**

$$\psi + t(df)_{\alpha/2}\times s_\psi$$

**Effect size of linear contrast:** Cohen's $d$

$$d = \frac{\psi}{s}$$

Options for $s$:

- Sample SD of focal group of interest
- $\sqrt{MSW}$ (square root of mean squares within)

## Specifying Weights

In our example, we use the following weights:

$$a_1 = 1, \quad a_2 = -\frac{1}{3}, \quad a_3 = -\frac{1}{3}, \quad a_4 = -\frac{1}{3}$$

The following weights will give the same results:

$$a_1 = 3, \quad a_2 = -1, \quad a_3 = -1, \quad a_4 = -1$$

**Interpretation of contrast weights:**

- Group with positive weights will be compared to those with negative weights
- Groups with zero weight are not considered in the contrast

## Contrasts in R

::: {style="font-size: 0.85em;"}

$H_1$: "close" is different than "intermediate" and "distant"

```{r}
#| echo: true
#| message: false
contrast(em1, method = list(c(-2, 1, 1)))
```

$H_1$: "close" is different than "intermediate"

```{r}
#| echo: true
#| message: false
contrast(em1, method = list(c(1, -1, 0)))
```

:::

## Effect Size Way 1 (by hand)

Using $\sqrt{MSW}$ and $H_1$: "close" is different than "intermediate" and "distant"

```{r}
summary(a1)
contrast(em1, method = list(c(1, -1, 0)))
```

$$d = \frac{\psi}{s} = \frac{0.129}{\sqrt{1.03}} = 0.13$$

## Effect Size Way 2

Using $\sqrt{MSW}$ and $H_1$: "close" is different than "intermediate" and "distant"

We can use `eff_size()` from `emmeans` but need to specify `sigma = sqrt(MSW)` and `edf = dfW`.

::: {style="font-size: 0.85em;"}

```{r}
summary(a1)
```

```{r}
#| echo: true
eff_size(em1, sigma = sqrt(1.03), edf = 122, 
         method = list(c(1, -1, 0)))
```

:::

## Orthogonal Contrasts

Historically, researchers have conducted **orthogonal** contrasts:

- $G-1$ linear contrasts
- For each contrast, $\sum a_g = 0$
- For all pairs of contrasts (weights $a_g$ and $b_g$), $\sum (a_g b_g) = 0$

Modern practices generally don't require orthogonal contrasts:

- Theory should drive comparisons
- Orthogonality not required

**Correct for multiple comparisons when specified a priori?**

- Traditional/textbook: Bonferroni, Holm, or other correction of $\alpha$
- Modern: Open debate about whether correction is necessary

## Trend Analysis

Special application of linear contrasts where groups are ordered (e.g., dosage levels 0ml, 5ml, 10ml, 15ml, 20ml).

More interested in the form of the relationship:

- **Linear relationship:** (-2, -1, 0, 1, 2)
- **Quadratic relationship:** (-2, 1, 2, 1, -2)
- **Cubic relationship:** (-1, 2, 0, -2, 1)


## APA Writeup

A one-way ANOVA of social cues on affect evaluation was conducted. The omnibus test of priming condition was not significant $F(2, 122) = 0.24$, $p = 0.79$, $\omega^2 < .001$. The contrast between the Close and Intermediate conditions was not significant, $t(122) = 0.59$, $p = .55$, $d = .13$.

*Note: Multiple comparisons were not adjusted for here. If you adjust, state the method used!*


## The Bayesian Perspective

Do we need multiple comparisons adjustments when we use Bayesian methods? Why or why not?

- [Multiple adjustments are about adjusting the probability we make an incorrect **decision** based on a prescribed cutoff]{.answer .fragment .content-visible unless-meta="params.v0"}
- [If we use Bayes factors with a cutoff we might need something similar]{.answer .fragment .content-visible unless-meta="params.v0"}
- [Alternatively, we can just consider the posterior distribution]{.answer .fragment .content-visible unless-meta="params.v0"}
- [What range of effects are plausible based on posterior?]{.answer .fragment .content-visible unless-meta="params.v0"}

## Posterior Extraction

- **Step 1:** Set up linear model $y_{ig} = \color{#377EB8}{\beta_0} + \color{#4DAF4A}{\sum_{p=1}^{g-1}\beta_p x_{p}} + \color{#984EA3}{\epsilon_i}$
- **Step 2:** Plug in for $x$'s to get $\hat{y}$ for first (reference) group
  - If using dummy coding, this would be all $x_p = 0$
- **Step 3:** plug in for $x$'s to get $\hat{y}$ for group 2
  -  If using dummy coding, this would be all $x_1 = 1$ and all other $x_2 \dots x_p = 0$
- Repeat for all groups
- Do this across **all draws** from MCMC output
- Calculate pairwise differences (or linear contrasts) for **every draw**



## Example

```{r}
#| echo: true
library(rstanarm)
options(mc.cores = parallel::detectCores())


stan1 <- stan_glm(average_bond ~ condition, dat, family = "gaussian",
                  seed = 32590)
summary(stan1)


```





## Draw Extraction I

We can make predictions about $\hat{y}$ by creating a toy dataset that has the values of $x$ we want to make predictions about. In this example, that means a dataset where there are three rows: 

- one for `condition = close` 
- one for `condition = intermediate`
- one for `condition = distant`




I can then use `posterior_linpred()` to make predictions about each of these rows across the 4000 draws generated by `Stan`. This represents my posterior prediction about each group.

## Draw Extraction II

```{r}
#| echo: true
(toydata <- data.frame(condition = c("close","intermediate", "distant")))
(post_pred <- posterior_linpred(stan1, newdata = toydata))


```



## Calculate Differences

Calculating the differences between each column represents the posterior distribution of pairwise differences between each group!

```{r}
#| echo: true
post_diffs <- data.frame(post_pred) |> 
  mutate(close_intermediate = X1 - X2,
         close_distant = X1 - X3,
         intermediate_distant = X2 - X3)


summary(post_diffs |> 
          select(close_intermediate, close_distant, intermediate_distant))

```



## Credible Intervals

We can use the `quantile()` function to get posterior medians and equal tailed credible intervals

```{r}
#| echo: true
round(quantile(post_diffs$close_intermediate, probs = c(.025, .5, .975)), 2)
round(quantile(post_diffs$close_distant, probs = c(.025, .5,  .975)), 2)
round(quantile(post_diffs$intermediate_distant, probs = c(.025, .5,  .975)), 2)
```


## Using `emmeans()`

Of course, `R` makes this easier for us with `emmeans`, which does all this work for us.  This will give us the highest posterior density interval (slightly different from equal tailed output on prior slide)

```{r}
#| echo: true
em_stan1 <- emmeans(stan1, ~ condition)
pairs(em_stan1)

```

And we can also include linear contrasts if we have a specific linear hypothesis

```{r}
#| echo: true
contrast(em_stan1, method = list(c(-2, 1, 1)))
```

## Comparison To Frequentist Result

```{r}
#| echo: true
pairs(em1)
contrast(em1, method = list(c(-2, 1, 1)))
```

## Bayes APA Writeup

A one-way Bayesian ANOVA of social cues on affect evaluation was conducted. Pairwise comparisons between all groups were estimated. The contrast between the Close and Intermediate had a posterior mean $= 0.13, 95\% \text{CI} = [-0.33, 0.56]$. The contrast between the Close and Distant groups had posterior mean $=0.14, 95\% CI = [-0.32, 0.58]$. The contrast between Intermediate and Distant Groups had posterior mean $=0.00, 95\% CI = [-0.49, 0.43]$. These results fail to support an effect of social cues on affect evaluation. 