---
title: "Lecture 2: Exploring Data"
author: "Mark Himmelstein"
format: 
  revealjs:
    slide-width: 1920
    slide-height: 1080
    theme: ../../../gt-theme.scss
    slide-number: true
    slide-number-format: "c/t"
    html-math-method: mathjax
    include-in-header:
      text: |
        <script>
        window.MathJax = {
          tex: {
            packages: {'[+]': ['cancel']}
          },
          loader: {
            load: ['[tex]/cancel']
          }
        };
        </script>
params:
  v0: true 
editor: source
---

```{r}
#| echo: false


library(knitr)
library(formatR)
library(tidyverse)
opts_chunk$set(out.width = '100%', options(width=60))

opts_knit$set(global.par = TRUE)
opts_chunk$set(echo=FALSE, warning = FALSE, message = FALSE, comment = '', class.output='sh',
               fig.width=8, fig.height=5, fig.align = "center")

my_red <- "#E41A1C"
my_blue <- "#377EB8"
my_green <- "#4DAF4A"
my_purple <- "#984EA3"
my_orange <- "#FF7F00"
my_yellow <- "#FFFF33"
my_red2 <- "#FBB4AE" 
my_blue2 <- "#B3CDE3" 
my_green2 <- "#CCEBC5" 
my_purple2 <- "#DECBE4"
my_orange2 <- "#FED9A6"
my_yellow2 <- "#FFFFCC"

```

## What is this class about?

[A lot of people think statistics is math.]{.fragment}

[But it's so much more. It's also *science*, it's also *philosophy*, and even has roots in *psychology*]{.fragment}

## What is this class about?

It is the way we take information that is messy and difficult to learn anything from....

```{r fig.height=4}
#| echo: false


set.seed(1290)
dat_sim <- data.frame(age = round(runif(200, 13, 18), 2))
dat_sim$curric <- sample(c(1,2), 200, replace = TRUE)
set.seed(21512)
dat_sim$score <- 40 + 5 * dat_sim$curric + 2 * dat_sim$age + rnorm(200, 0, 7.5)

dat_sim$curric <- ifelse(dat_sim$curric == 1, "B","A")

ggplot(dat_sim, aes(x = age, y= score, color = factor(curric))) + geom_point(size=2) +  xlab("Age") + ylab("Math Score") + labs(color = "Curriculum") + 
  papaja::theme_apa() +
  theme(axis.text = element_text(size = 18), 
  axis.title = element_text(size = 24)
  )

```

## What is Statistics?

....and figure out what it can tell us if we look at it carefully enough....

```{r fig.height=4}
ggplot(dat_sim, aes(x = age, y= score, color = factor(curric))) + geom_point(size=2) +  xlab("Age") + ylab("Math Score") + labs(color = "Curriculum") + stat_smooth(method = "lm") +
  papaja::theme_apa() +
  theme(axis.text = element_text(size = 18), 
  axis.title = element_text(size = 24)
  )
```

## Data and Simulation

That data wasn't real. It was **simulated**

-   I gave the computer some **parameters**
-   It generated random **data** based on those parameters

Simulation is an increasingly important tool in the statisticians toolkit! You'll learn more about this throughout the class and in lab. For now, the important thing to understand....

## Probability and Statistics

```{r}
#| fig.width: 6
#| fig.height: 2
#| out.width: "100%"

# Zero out standard plot margins
par(mar = c(0, 0, 0, 0))

# Initialize empty canvas
plot(0, 0, col = "white", xlim = c(0, 5), ylim = c(0, 3.5),
     bty = "n", xlab = "", ylab = "", xaxt = "n", yaxt = "n")

# --- Left Box (Population) ---
polygon(x = c(0.2, 1.7, 1.7, 0.2), y = c(2, 2, 3, 3), border = my_blue, lwd = 2)

# --- Right Box (Sample) ---
polygon(x = c(3.3, 4.8, 4.8, 3.3), y = c(.5, .5, 1.5, 1.5), border = my_purple, lwd = 2)

# --- Top Curve (Probability) ---
x_top <- seq(0.95, 4.05, length = 30)
y_top <- -0.478 * x_top^2 + 1.905 * x_top + 1.621
points(x_top, y_top, type = 'l', col = my_blue, lwd = 2)

# Top Arrow (pointing to Sample)
arrows(x0 = x_top[28], x1 = x_top[30], 
       y0 = y_top[28], y1 = y_top[30], 
       col = my_blue, lwd = 2)

# --- Bottom Curve (Statistics) ---
x_bot <- seq(0.95, 4.05, length = 30)
y_bot <- 0.435 * x_bot^2 - 2.656 * x_bot + 4.132
points(x_bot, y_bot, type = 'l', col = my_purple, lwd = 2)

# Bottom Arrow (pointing to Population)
arrows(x0 = x_bot[3], x1 = x_bot[1], 
       y0 = y_bot[3], y1 = y_bot[1], 
       col = my_purple, lwd = 2)

# --- Text Labels ---
text(x = 0.95, y = 2.65, "Population", col = my_blue)
text(x = 0.95, y = 2.35, "(Parameter)", col = my_blue)

text(x = 4.05, y = 1.15, "Sample", col = my_purple)
text(x = 4.05, y = 0.85, "(Data)", col = my_purple)

text(x = 3.1, y = 3.3, "Probability", col = my_blue)
text(x = 1.9, y = 0.3, "Statistics", col = my_purple)
```

:::::: {style="margin-top: -10px;"}
::::: {.columns style="font-size: 0.75em;"}
::: {.column width="50%" style="color: #377EB8;"}
Probability tells us how data are *randomly* **generated** based on some underlying facts about a population.
:::

::: {.column width="50%" style="color: #984EA3;"}
Statistics are how we infer underlying facts about a **population** from a finite set of observed data (*sample*).
:::
:::::

This course will begin and end with [**statistics**]{style="color: #984EA3;"}, with many detours to [**probability**]{style="color: #377EB8;"} along the way
::::::

## Terminology

The difference between [**statistics**]{style="color: #984EA3;"} and [**probability**]{style="color: #377EB8;"} is in what is **known** or **held constant**

-   In [**probability** parameters are assumed to be known]{style="color: #377EB8;"}
    -   **Parameter:** A quantity that describes (a property of) an entire population
-   In [**statistics** data are assumed to be known]{style="color: #984EA3;"}

What IS/ARE data? (Singular: datum)

-   Distinct pieces of **observed** information
-   **Statistic:** Quantity calculated from a finite amount of data to *estimate a parameter*

We often arrange data into rectangular **datasets**

## An Example Dataset

-   Rows: persons

-   Columns: variables/responses

```{r}
library(dplyr, warn.conflicts = FALSE)
d <- data.frame(Participant_ID = 1:10,
                Num_Siblings = c(1, 2, 2, 1, 2, 1, 0, 3, 3, 0),
                Gender = factor(c("M", "M", "M", "F", "M", "F", "F", "F", "F", "M")),
                Age = c(18.1, 35.4, 42.6, 19.6, 20.4, 23.4, 26.8, 18.3, 30.6, 29.7),
                Shoe_Size = c(10.5, 9.5, 10, 7.5, 9, 7, 7.5, 8.5, 6.5, 9.5),
                IQ = c(121, 109, 113, 123, 95, 110, 103, 99, 120, 95))
d <- as_tibble(d)
d
```

## The Data Box

-   Cattell (1966, 1988) proposed the concept of a **data box** to describe the structure of data.
-   The data box can be useful in understanding the research question, research design, and statistical analysis.

**Suggested readings:**

::: {style="font-size: 0.8em;"}
-   Cattell, R. B. (1988). *The data box: Its ordering of total resources in terms of possible relational systems*. In J. R. Nesselroade & R. B. Cattell (Eds.), *Perspectives on individual differences. Handbook of multivariate experimental psychology* (pp. 69--130). Plenum Press. <https://doi.org/10.1007/978-1-4613-0893-5_3>
-   Biesanz, J. C., West, S. G., & Kwok, O. M. (2003). Personality over time: Methodological approaches to the study of short-term and long-term development and change. *Journal of Personality*, *71*(6), 905--942. <https://doi.org/10.1111/1467-6494.7106002>
-   Ozer, D. J. (1986). *Consistency in personality: A methodological framework*. Springer-Verlag.
:::

## The Data Box

:::::: {style="font-size: 0.8em;"}
**Four dimensions:**

::::: columns
::: {.column width="50%"}
1.  Person (P)
2.  Occasion (O)
:::

::: {.column width="50%"}
3.  Situation (S)
4.  Variable (V)
:::
:::::
::::::

```{r}
#| echo: false
#| fig.align: "center"
#| fig.width: 8
#| fig.height: 4
#| out.width: "90%"

# Remove plot margins
par(mar = c(1, 1, 1, 1))
plot(0, 0, type = "n", xlim = c(-1, 12.5), ylim = c(-0.8, 5.8), axes = FALSE, xlab = "", ylab = "")

draw_databox <- function(xoff, p_label) {
  # Front face
  rect(xoff, 0, xoff + 4, 4, border = "#377EB8", lwd = 2)
  for (i in 1:3) {
    lines(c(xoff + i, xoff + i), c(0, 4), col = "#B3CDE3", lty = 2)
    lines(c(xoff, xoff + 4), c(i, i), col = "#B3CDE3", lty = 2)
  }
  
  # 3D Perspective lines
  for (i in 0:4) {
    lines(c(xoff + i, xoff + i + 1), c(4, 5), col = "#377EB8", lwd = 1.5)
    lines(c(xoff + 4, xoff + 5), c(i, i + 1), col = "#377EB8", lwd = 1.5)
  }
  lines(c(xoff + 1, xoff + 5), c(5, 5), col = "#377EB8", lwd = 2)
  lines(c(xoff + 5, xoff + 5), c(1, 5), col = "#377EB8", lwd = 2)
  
  # Inner depth lines (Fixed line)
  lines(c(xoff + 0.5, xoff + 4.5), c(4.5, 4.5), col = "#B3CDE3", lty = 2)
  lines(c(xoff + 4.5, xoff + 4.5), c(0.5, 4.5), col = "#B3CDE3", lty = 2)
  
  # Labels
  text(xoff + 0.5 + 0:3, -0.4, paste0("O", 1:4), cex = 0.9, col = "#333333")
  text(xoff - 0.4, 3.5 - 0:3, paste0("V", 1:4), cex = 0.9, col = "#333333")
  text(xoff - 0.2, 4.35, "S1", cex = 0.8, col = "#333333")
  text(xoff + 0.2, 4.85, "S2", cex = 0.8, col = "#333333")
  text(xoff + 2.5, 5.4, p_label, cex = 1.2, font = 2, col = "#377EB8")
}

draw_databox(0, "Person 1")
draw_databox(6.8, "Person 2")
```

## The Data Box & Research Designs

:::::: {style="font-size: 0.65em;"}
**Example:**

::::: columns
::: {.column width="50%"}
-   **Persons:** individual children
-   **Occasion:** child age (e.g., 6, 8, 10)
:::

::: {.column width="50%"}
-   **Situation:** before/after a snack
-   **Variable:** scores on different recall tasks
:::
:::::

**Cross-Sectional Design** (single time point):

-   Person $\times$ Variable
    -   Test scores of a sample of 6-year-olds
-   Person $\times$ Situation $\times$ Variable
    -   Test scores of a sample of 6-year-olds measured before and after a snack

**Longitudinal Designs** (multiple time points):

-   Person $\times$ Occasion $\times$ Variable (single response)
    -   Test scores of the same children at ages 6, 8, and 10
-   Person $\times$ Occasion $\times$ Situation $\times$ Variable (multiple responses)
    -   Test scores of the same children at ages 6, 8, and 10, measured before and after a snack
::::::

## Measurement Scales

Stevens (1946) classified variables as being measured on one of four scales:

1.  **Nominal**: Variables with discrete categories

2.  **Ordinal**: Variables with ordering or ranking, distances between scale points are not equal

3.  **Interval**: Distances between scale points are equal (zero has no special meaning)

4.  **Ratio**: Distances between scale points are equal (zero has a special meaning, usually the absence of the trait)

::: {style="font-size: 0.8em;"}
Stevens, S. S. (1946). On the theory of scales of measurement. *Science*, *103*, 677--680. <https://doi.org/10.1126/science.103.2684.677>
:::

## Measurement Scales

\small

My recommendation: Don't take measurement scales too seriously

Perhaps a more useful distinction:

::::: columns
::: {.column width="50%"}
**Discrete/Categorical variables** - Frequencies

-   Percentages

-   Chi-square tests

-   Logistic Regression
:::

::: {.column width="50%"}
**Continuous variables**

-   Mean

-   Standard Deviation

-   $t$-tests

-   Correlation
:::
:::::

-   Middle ground: count data, Likert scales

Better to understand what statistical tests can and cannot tell you, and whether **interpretations** make sense

## Thinking it Through

It can also be helpful to think about how different types of data compare to one another. In what ways are the following variables different or similar in terms of their *data structure*?

-   [**The number of siblings a person has**]{style="color: red;"} vs. [**a person's age**]{style="color: blue;"}

-   [**the number of points the home team scores**]{style="color: red;"} vs. [**the difference in score between the two teams**]{style="color: blue;"}

-   [**the probability a political incumbent wins election**]{style="color: red;"} vs. [**the number of votes they receive**]{style="color: blue;"}

Thinking about these differences can be critical to understanding what statistical techniques and interpretations are appropriate.

## 2026 NBA Data

```{r}
#| echo: TRUE
bball <- read.csv("2026NBA_data.csv") |> filter(G >= max(G)/2,
                                                draft_year != "")
bball |> glimpse()
```

## 2026 NBA Data Sources

-   [Basketball Reference 2026 season statistics](https://www.basketball-reference.com/leagues/NBA_2026_totals.html)
-   [Kaggle dataset with demographic data on players through 2021](https://www.kaggle.com/datasets/justinas/nba-players-data/data)
    -   Means we are missing demographic data on some younger players
    -   I removed these players from our dataset for simplicity

## Variable Types in R

Which variables are discrete? Continuous? How can you tell?

-   [Continuous data are sometimes referred to as "numeric" in R. We see two types of numeric variable here]{.answer .fragment .content-visible unless-meta="params.v0"}
    -   [`int` refers to integers]{.answer .fragment .content-visible unless-meta="params.v0"}
    -   [`dbl` refers to "double" which for our purposes are simply real numbers]{.answer .fragment .content-visible unless-meta="params.v0"}
-   [Discrete data are currently all listed as `chr` for "character". R is just treating these as strings for now]{.answer .fragment .content-visible unless-meta="params.v0"}
    -   [We can convert variables to "factors" which is the more appropriate R type for discrete variables (R will do this automatically for `chr` variables in many analyses)]{.answer .fragment .content-visible unless-meta="params.v0"}
    -   [R also has a special binary variable called "logical" which can only be `TRUE` or `FALSE`]{.answer .fragment .content-visible unless-meta="params.v0"}

## 2026 NBA Data Variables {.scrollable}

::: {style="font-size: 0.8em;"}
```{r}


tab <- data.frame(Variable = names(bball),
                  Description = c("Player's Name",
                                  "Player's Age",
                                  "Player's Team",
                                  "Player's Height (cm)",
                                  "Player's Weight (kg)",
                                  "College Attended",
                                  "Country of Origin",
                                  "Year Drafted",
                                  "Round Drafted",
                                  "Primary Position",
                                  "Games Played",
                                  "Games Started",
                                  "Minutes Played",
                                  "Points Per Game",
                                  "Field Goals Per Game",
                                  "Field Goal Attempts Per Game",
                                  "Field Goal Percentage",
                                  "Three Pointers per Game",
                                  "Three Point Attempts per Game",
                                  "Three Point Percentage",
                                  "Free Throws per Game",
                                  "Free Throw Attempts per Game",
                                  "Free Throw Percentage",
                                  "Offensive Rebounds per Game",
                                  "Defensive Rebounds per Game",
                                  "Total Rebounds Per Game",
                                  "Assists per Game",
                                  "Steals per Game",
                                  "Blocks per Game"
                                  ))

knitr::kable(tab)
```
:::

## Factors and Logicals

```{r}
#| echo: TRUE
bball <- bball |> mutate(draft_round = factor(draft_round),
                         # new variable for whether player was drafted:
                         drafted = draft_round != "Undrafted")
bball |> 
  glimpse()

```

## [Descriptive Statistics: Discrete Data]{style="font-size: 0.7em;"}

**Frequencies:** Counts of each category

```{r}
#| echo: TRUE

table(bball$draft_round)
```

**Proportions (Relative Frequencies)**: Counts of each category

```{r}
#| echo: TRUE

round(prop.table(table(bball$draft_round)), 2)
```

**Percentages:** Proportions $\times$ 100

```{r}
#| echo: TRUE

round(prop.table(table(bball$draft_round)) * 100, 2)
```

## Frequency Table {.scrollable}

```{r}
#| label: tbl-1
#| tbl-cap: "Frequency Table of Players by Country"


library(flextable)


bbal_desc <- bball |> 
  group_by(draft_round, Country) |>
  summarize(Freq. = n(), .groups = "drop") %>%
  pivot_wider(names_from = draft_round, values_from = Freq., names_prefix = "Round ") %>%
  arrange(desc(`Round 1`)) |>
  rename("Undrafted" = "Round Undrafted")
  
bbal_desc[is.na(bbal_desc)] <- 0

bbal_desc <- bbal_desc |> 
  mutate(
    `Round 1` = paste0(`Round 1`, " (", round(`Round 1`/sum(`Round 1`), 4) * 100, "%)"),
    `Round 2` = paste0(`Round 2`, " (", round(`Round 2`/sum(`Round 2`), 4) * 100, "%)"),
    `Undrafted` = paste0(`Undrafted`, " (", round(`Undrafted`/sum(`Undrafted`), 4) * 100, "%)")
  )

# 2. APA Formatting using flextable
bbal_desc |>
  flextable() |>
  theme_apa() |>
  align(j = 2:4, align = "center", part = "all") |> # Centers the data columns
  set_caption("Frequency Table of Players by Country") |>
  autofit() # Automatically adjusts column widths


```

## [Descriptive Statistics: Continuous Data]{style="font-size: 0.667em;"}

-   **Central Tendency:** The **center** of the distribution, **most common values** in the distribution, and **expected value** based on the distribution.

    -   Mean (and trimmed mean)

    -   Median

    -   Mode

-   **Variability:** How much the data **varies** or is **spread** around the center of the distribution.

    -   Variance

    -   Standard Deviation

    -   Interquartile range

## Central Tendency

-   **Median**: Score that corresponds to the point at or below which 50% of scores fall (order data first)

    -   Also known as the 50$^{\text{th}}$ percentile

    -   **Percentile**: point at or below which specified % of scores fall

-   **Mean**: $\bar{x} = \sum x / N$. Anyone want to try and break down this notation? It's important!

    -   [$\sum x$: sum of all the $x$ scores across all observations]{.answer .fragment .content-visible unless-meta="params.v0"}

    -   [$N$: sample size]{.answer .fragment .content-visible unless-meta="params.v0"}

    -   [$\bar{x}$: sample mean calculated from data]{.answer .fragment .content-visible unless-meta="params.v0"}

    -   [$\mu$: population mean, expected value of probability distribution]{.answer .fragment .content-visible unless-meta="params.v0"}

## Central Tendency (con't)

The **trimmed mean** is a robust statistic that reduces the two disadvantages of the mean

1.  Sensitivity to asymmetric distributions

2.  Sensitivity to extreme scores

**5% trimmed mean** is the mean of the data that discards the lowest 5% and the highest 5% of scores in the original data

The **mode** is the value with the highest frequency

-   Often used for discrete data

-   Also used for describing continuous distributions (esp. graphically)

## Measures of Central Tendency: Summary

```{r}
library(gridExtra, warn.conflicts = FALSE)
library(ggplot2)

p1 <- ggplot() +
  # Draw the normal distribution curve from x = -3 to 3
  stat_function(fun = dnorm, xlim = c(-3, 3), linewidth = 1) +
  
  # Add the vertical center line
  annotate("segment", x = 0, xend = 0, y = 0, yend = 0.40, linetype = "dashed") +
  
  # Add top labels
  annotate("text", x = 0, y = 0.42, label = "mean") +
  annotate("text", x = 0, y = 0.44, label = "median") +
  annotate("text", x = 0, y = 0.46, label = "mode") +
  
  # Add bottom label below the x-axis origin
  annotate("text", x = 0, y = -0.03, label = "symmetric distribution") +
  
  # Adjust axes to ensure labels fit comfortably
  lims(x = c(-3, 3), y = c(-0.05, 0.5)) +
  
  # Clean layout
  theme_void()

library(ggplot2)

alpha <- 2
beta <- 5

# Pre-calculate central tendency metrics for clean code
mode_val   <- (alpha - 1) / (alpha + beta - 2)          # 0.200
median_val <- (alpha - 1/3) / (alpha + beta - 2/3)    # ~0.263 (approx)
mean_val   <- alpha / (alpha + beta)                   # ~0.286

p2 <- ggplot() +
  # Draw Beta distribution curve
  stat_function(
    fun = dbeta, 
    args = list(shape1 = alpha, shape2 = beta), 
    xlim = c(0, 1), 
    linewidth = 1
  ) +
  
  # Vertical dashed reference lines
  annotate("segment", x = mode_val, xend = mode_val, y = 0, yend = 2.55, linetype = "dashed") +
  annotate("segment", x = median_val, xend = median_val, y = 0, yend = 2.40, linetype = "dashed") +
  annotate("segment", x = mean_val, xend = mean_val, y = 0, yend = 2.25, linetype = "dashed") +
  
  # Labels for mode, median, and mean
  annotate("text", x = mode_val, y = 2.65, label = "mode") +
  annotate("text", x = median_val + 0.04, y = 2.50, label = "median") +
  annotate("text", x = mean_val + 0.05, y = 2.35, label = "mean") +
  
  # Label below the curve baseline
  annotate("text", x = 0.4, y = -0.15, label = "asymmetric distribution") +
  
  # Axis limits (extending negative y to prevent text clipping)
  lims(x = c(0, 0.8), y = c(-0.25, 2.8)) +
  
  # Clean layout
  theme_void()

# Display plot

grid.arrange(p1, p2, nrow = 1)
```

## Central Tendency: Properties

::: {style="font-size: 0.8em;"}
| Property | **Mode** | **Median** | **Mean** |
|:-----------------|:-----------------|:-----------------|:-----------------|
| **Score that is actually observed in the data** | Always | Not Necessarily | Not Necessarily |
| **Affected by extreme scores** | Less Affected | Less Affected | More Affected |
| **Affected by asymmetric distribution** | Least Affected | Less Affected | Most Affected |
| **Full use of all participants' scores** | No | No | Yes |
| **Applications in other statistical analyses** | Few | Few | Many |
:::

## Range and IQR

::: {style="font-size: 0.8em;"}
**Variability** measures how spread out the data are

-   **Range**: = maximum score - minimum score

    -   Highly sensitive to extreme scores

    -   Does not use all of the information in the data

-   **Interquartile Range (IQR)** = $75^\text{th}$ percentile - $25^\text{th}$ percentile

    -   $25^\text{th}$ percentile = $1^\text{st}$ quartile (1Q)

    -   $75^\text{th}$ percentile = $3^\text{rd}$ quartile (3Q)

    -   Less sensitive to extreme scores

    -   Does not use all of the information in the data

-   Some variables are commonly known to be asymmetric with extreme values. Example: household income

-   For these variables, best to report median + IQR

-   Example: median annual household income in Manhattan (as of 2014) = \$66,739, IQR = \$24,567
:::

## Variance and SD

**Variance** (sample: $s^2$, population: $\sigma^2$):

$$s^2 = \frac{\sum(x - \bar{x})^2}{N-1}\,\,\,\,\, \sigma^2 = \frac{\sum(x - \mu)^2}{N}$$

-   $(x - \bar{x})$: deviation scores (deviation from mean)

-   $(x - \bar{x})^2$: square deviation scores to make them positive

-   $\frac{\sum}{N-1}$: average

-   $(N-1)$ are **degrees of freedom** used in the sample calculation

**Standard Deviation** (sample: $SD$ or $s$, population: $\sigma$)

-   Square root of the variance
-   Variance is in squared units, SD is in the units of the variable itself.

## Pythathogrean Theorm of Statistics

-   Why $(x - \bar{x})^2$ instead of $|x - \bar{x}|$?

[Variance is *additive*]{.answer .fragment .content-visible unless-meta="params.v0"}

[$s^2_x + s^2_y - 2\text{cov}_{xy} = s^2_{x + y}$]{.answer .fragment .content-visible unless-meta="params.v0"}

Or

[$\sqrt{s_x^2 + s_y^2 - 2\text{cov}_{xy}} = SD_{x + y}$]{.answer .fragment .content-visible unless-meta="params.v0"}

[Where $\text{cov}_{xy}$ is the *covariance* (more later) between $x$ and $y$]{.answer .fragment .content-visible unless-meta="params.v0"}

## Degrees of Freedom

Degrees of freedom ($df$) are the number of elements in a statistic's calculation that are free to vary

-   For the variance calculation, we lose 1 $df$ by calculating the sample mean

Let's consider the following set of numbers \[1, 2, $x$\]. What is $x$?

[$x$ could be anything!]{.answer .fragment .content-visible unless-meta="params.v0"}

[How what if I told you that the mean of these three numbers was $\bar{X} = 2$. Now what is $x$?]{.fragment}

[$x$ must be 3!]{.answer .fragment .content-visible unless-meta="params.v0"}

## [Degrees of Freedom Explained]{style="font-size: 0.7em;"}

[This is what degrees of freedom are: The number of elements that can freely vary. In the first example, there is one degree of freedom, because $x$ can be anything. In the second, there are zero, because there are no elements that can freely vary.]{.answer .fragment .content-visible unless-meta="params.v0"}

[Because we already account for the sample mean in estimating the variance, we are calculating the variance based on a sample that has $N - 1$ elements that can freely vary, so that's our appropriate sample size to average over]{.answer .fragment .content-visible unless-meta="params.v0"}

## [Degrees of Freedom Simulated]{style="font-size: 0.7em;"}

::: {style="font-size: 0.9em;"}
Sometimes when you want to prove a concept to yourself, it helps to simulated it.

In the following slides I:

-   Set the population mean $\mu = 0$
-   Set the population variance $\sigma^2 = 10$
-   Randomly generated a sample of size $N = 5$ from these parameters
-   Calculated the variance using each of the following formulas
    -   $\frac{\sum(x - \bar{x})^2}{N}$
    -   $\frac{\sum(x - \bar{x})^2}{N - 1}$
-   Repeated the process 100,000 times
-   Plotted the results to compare how well the sample statistics from these two formulas reproduced the known variance parameter $\sigma^2 = 10$

Our goal is to have $s^2$ equal to the $\sigma^2$ \textit{on average}
:::

## Simulated Example

```{r, warning = FALSE}
#| cache: true
dt <- data.frame(N = rep(NA, 100000), `N - 1` = rep(NA, 100000), check.names = FALSE)

set.seed(10202)
for(i in 1:100000){
  vr <- rnorm(5, 0, sqrt(10))
  dt$N[i] <- sum((vr - mean(vr))^2) / 5
  dt$`N - 1`[i] <- sum((vr - mean(vr))^2) / 4
}

dt_l <- pivot_longer(dt, cols = 1:2, names_to = "Denominator", values_to = "Variance")

mb <- round(mean(dt_l$Variance[dt_l$Denominator == "N"]), 2)
mu <- round(mean(dt_l$Variance[dt_l$Denominator == "N - 1"]), 2)


# Plot 1
p1 <- ggplot(dt_l, aes(x = Variance, fill = Denominator)) + 
  geom_vline(xintercept = 10, linetype = 1) + 
  annotate(geom = "text", x = 14, y = .07, label = "sigma^2 == 10", parse = TRUE, size = 6) +
  ylab(NULL) +
  xlim(c(0, 50)) + ylim(c(0, .1)) + 
  theme_bw() + 
  theme(legend.position = "bottom")

# Plot 2
p2 <- ggplot(dt_l %>% filter(Denominator == "N"), aes(x = Variance, fill = Denominator)) + 
  geom_vline(xintercept = 10, linetype = 1) + 
  annotate(geom = "text", x = 14, y = .07, label = "sigma^2 == 10", parse = TRUE, size = 6) +
  annotate(geom = "text", x = mb - 5, y = .0975, label = paste0("s^2 == ", mb), parse = TRUE, size = 6, color = "red") +
  geom_density(alpha = .5) + 
  scale_fill_manual(values = c("red", "blue")) + 
  geom_vline(xintercept = mb, linetype = 2, color = "red") +
  ylab(NULL) + 
  xlim(c(0, 50)) + ylim(c(0, .1)) + 
  theme_bw() + 
  theme(legend.position = "bottom")

# Plot 3
p3 <- ggplot(dt_l, aes(x = Variance, fill = Denominator)) + 
  geom_vline(xintercept = 10, linetype = 1) + 
  annotate(geom = "text", x = 14, y = .07, label = "sigma^2 == 10", parse = TRUE, size = 6) +
  annotate(geom = "text", x = mb - 5, y = .0975, label = paste0("s^2 == ", mb), parse = TRUE, size = 6, color = "red") +
  annotate(geom = "text", x = 14, y = .055, label = paste0("s^2 == ", mu), parse = TRUE, size = 6, color = "blue") +
  geom_density(alpha = .5) + 
  scale_fill_manual(values = c("red", "blue")) + 
  geom_vline(xintercept = mb, linetype = 2, color = "red") + 
  geom_vline(xintercept = mu, linetype = 2, color = "blue") + 
  ylab(NULL) +
  xlim(c(0, 50)) + ylim(c(0, .1)) + 
  theme_bw() + 
  theme(legend.position = "bottom")
p1
```

## Simulated Example

```{r warning = FALSE}
p2
```

## Simulated Example

```{r warning = FALSE}
p3
```

## [Mean and Variance of Discrete Data]{style="font-size: 0.7em;"}

Consider the binary variable we created "Was the player drafted or not?"

-   What do the mean and variance mean?

**Mean** of a binary variable $p$

-   $p$ is the proportion in category "1"

-   In this example, $p$ is the proportion players who were drafted

**Variance** of the binary variable $= p \times (1 - p)$

-   Depends on the mean

-   Largest variance if $p = 0.5$

-   Smallest (zero) variance if $p = 0$ or $p = 1$

## Summary Table in R {.scrollable}

```{r}
#| echo: true

bball |>
  group_by(draft_round) %>% 
  summarize(N = n(),
    mean_Age = round(mean(Age), 2), 
    sd_Age   = round(sd(Age), 2),
    
    mean_Height = round(mean(Height), 2), 
    sd_Height   = round(sd(Height), 2), 
    
    mean_Weight = round(mean(Weight), 2), 
    sd_Weight   = round(sd(Weight), 2)
  )
```

## In APA Format

One hundred sixty players were drafted in the first round (mean age = 28.0 years, $SD$ = 4.16. From the second round, 52 players were drafted (mean age = 22.2 years, SD = 3.16). Forty-six were undrafted (mean age = 27.8 years, SD = 3.41).

-   If a number has only one digit or begins a sentence, write it out. Otherwise, use the numeral.
-   Round as much as possible while preserving relevant information
-   Be consistent in the number of decimals you report, more than 2 is rarely necessary
-   Always report the unit

## Skewness

Skewness is the degree of asymmetry in a variable

$$\text{Skew} = \frac{\sum(x - \bar{x})^3 / N}{[\sum(x - \bar{x})^2 / N]^{3/2}}$$

-   Skew = 0 indicates a symmetric distribution

-   Negative skew: tail skews to left

-   Positive skew: tail skews to right

-   Higher magnitude of skew: more asymmetric variable

## Skewness

```{r}
library(forcats)
skew <- function(alpha, beta){
  2 * (beta - alpha) * sqrt(alpha + beta + 1) / ((alpha + beta + 2) * sqrt(alpha * beta))
}
d <- tibble(x = seq(0, 1, length = 100))
d <- d %>% mutate("  0" = dbeta(x, 5, 5),
             " .6" = dbeta(x, 2, 5),
             "-.6" = dbeta(x, 5, 2)) %>%
    pivot_longer("  0":"-.6", names_to = "skewness", values_to = "density") %>%
  mutate(skewness = fct_relevel(factor(skewness), " .6", "  0", "-.6"))

ggplot(d, aes(x, density, col = skewness)) + geom_path(size = I(2)) + theme_light() +
  labs(x = "", y = "") + scale_color_manual(values = c(my_blue, my_green, my_purple)) +
  theme(axis.text = element_blank(), legend.title = element_text(size = 12),
        legend.text = element_text(size = 12))
```

## Kurtosis

Kurtosis is the degree of peakedness in the variable

$$\text{Kurtosis} = \left(\frac{\sum(x - \bar{x})^4}{N}\right) \bigg/ \left(\frac{\sum(x - \bar{x}^2)}{N}\right)^2 - 3$$

-   Kurtosis = 0: mesokurtic

-   Normal distribution is mesokurtic

-   Negative kurtosis: platykurtic (lighter tails)

-   Positive kurtosis: leptokurtic (heavier tails)

## Kurtosis

```{r}
drcosd <- function(x,mu,s) {
  ifelse(abs(mu - s) < abs(x), 0,
  1/(2*s)*(1+cos(pi*(x-mu)/s)))
}

rrcosd <- function(n){
  y <- runif(n)

}
kurt1 <- 6 * (90 - pi^4) / (5 * (pi^2 - 6)^2)


d <- tibble(x = seq(-3, 3, length = 100))
d <- d %>% mutate("  0" = dnorm(x),
             "-1" = drcosd(x, 0, 2),
             "  6" = dt(x, 5)) %>%
  pivot_longer("  0":"  6", names_to = "kurtosis", values_to = "density") %>%
  mutate(kurtosis = fct_relevel(factor(kurtosis), "  6", "  0", "-1"))

ggplot(d, aes(x, density, col = kurtosis)) + geom_path(size = I(2)) + theme_light() +
  labs(x = "", y = "") + scale_color_manual(values = c(my_blue, my_green, my_purple)) +
  theme(axis.text = element_blank(), legend.title = element_text(size = 12),
        legend.text = element_text(size = 12))
```

## Moments of Distributions

Why are skewness and kurtosis defined as they are?

| Metric | Population Expression | Sample Statistic |
|:-----------------------|:----------------------:|:----------------------:|
| **Mean**<br>*(Center)* | $\mu = E(X)$ | $\bar{x} = \frac{\sum x}{N}$ |
| **Variance**<br>*(Deviation)* | $\sigma^{2} = E[(X-\mu)^{2}]$ | $s^{2} = \frac{\sum(x-\bar{x})^{2}}{N-1}$ |
| **Skewness**<br>*(Asymmetry)* | $E[(X-\mu)^{3}]$ | $\frac{\sum(x-\bar{x})^{3}/N}{\left[\sum(x-\bar{x})^{2}/N\right]^{3/2}}$ |
| **Kurtosis**<br>*(Peakedness)* | $E[(X-\mu)^{4}]$ | $\frac{\sum(x-\bar{x})^{4}/N}{\left[\sum(x-\bar{x})^{2}/N\right]^{2}} - 3$ |

## Next Time

Much of statistics is about using numerical summaries of data, since what we're often ultimately interested in is estimating parameters.

But interpreting numbers is hard!

Next time we will look at another way of interpreting data....
