---
title: "Lab 03 Activity"
author: "[Your Name]"
date: "[Today's Date]"
self-contained: true
---


### Activity Objectives:

- Practice data visualization with the `tidyverse`

- Practice simulating data


### Instructions:

- Fill in the code chunks

- Render the Quarto document and submit both the `.qmd` and `.html` file on Canvas


## Activity

### 0) If you'd like to load packages with `library()`, please do so in this block. 

```{r}

```



## Simulation


### 1) Extend or modify the data simulation shown in class in some way. Feel free to choose something interesting to you, but some possible ideas are: adding an additional condition, modifying an existing condition, replacing some distributions with different ones, comparing distributions with different parameters (e.g., comparing different means, SDs, degrees of freedom, etc. within a distribution. Specific potential idea: comparing a $\chi^2$ distribution with different degrees of freedom). For this extension, write 1--2 questions or goals below.




### 2) Next, make your plan. This can be brief, but sketch out the steps you will take to program this. If your extension involves varying some parameters, distribution, etc., specify the different levels it will take (e.g., testing a $\chi^2$ distribution with degrees of freedom x, y, and z).




### 3) Make your modifications to the simulation and show the results! Please include results for both the sample and sampling distributions.

```{r}

```


### 4) Provide a brief (2--4 sentences) writeup of your method and results.





# Whole Game Of Data Science


## Import

### 5) The `tidyverse` package includes a dataset called `billboard` that has song rankings for the Billboard Top 100 songs from the year 2000. You can access the `billboard` data object by just calling `billboard`. Use the `head()` function to display the first six rows. Then, use the `names()` to look at the column names.

```{r}

```


## Tidy


### 6a) Is the `billboards` dataset in wide or long format? How can you tell?




### 6b) Using the `?` tool on the dataset and / or the output from \#3, please describe the contents of this dataset (specifically, an overview of what you can find in the columns) (1--2 sentences).

```{r}

```



### 7c) Given what you said in \#4a, would you want to use the `pivot_longer()` or `pivot_wider()` function on this data?




### 8) Using a pipe, let's do a few things: 

* Tidy the data as you planned to in \#4c, 

* Mutate the column with the numbered weeks so that it's just numeric (hint: the `parse_number()` function will extract numbers from a string. Feel free to review its documentation with `?`), 

* Assign this tidied version to a variable so we can continue using it.

After the pipe, print the first six rows from this tidied version. 

```{r}

```


## Transform

### 9) Let's answer some questions with this data! Our first question is: what song spent the most number of weeks in the Billboard top 10? Using a pipe and the tidied dataset from the previous question, let's do a few things: 

* Filter the data to just include 10 or lower in the `ranking` column

* Summarize by the song name to get the number of weeks. HINT: the `n()` function returns the number of rows in a `summarize(.by = group)` group.

* Arrange the data with from highest to lowest number of weeks in the top 10

What song spent the most number of weeks in the Billboard top 10?

```{r}

``` 


### 10) Come up with one question you'd like to ask of the data through data transformation.




### 11) Using a pipe, transform the data to address your question.





## Visualize (Extra Practice!)

If you'd like more practice doing data visualization, here are some extra exercises. These will not count towards your grade, but I will provide feedback if you choose to do them for practice. I'm also happy to accept responses for these at a later date than this assignment's due date. 


### B1) Let's plot the billboards data. Let's make a line graph that has time (week) on the x-axis and a song's ranking on the y-axis. Start by using the `ggplot()` function to make a grid with axes (no geoms yet).

```{r}

```


### B2) Now, let's add a `geom_line()` to show the rankings of the different songs over time. Remember you will need to add a `group` argument to your aes.

```{r}

```


### B3) Because higher is "worse" for rankings, let's reverse the y-axis so higher on the y is "better" by adding a scale layer. The function `scale_y_reverse()` will do this for us! Add it on to your plot just like you would add a new layer.

```{r}

```


### B4) Modify the `alpha` argument within your geom to make it easier to see the patterns.

```{r}

```


### B6) Make three more modifications to this plot.

```{r}

```


### B7) Describe what you see in this plot (1--2 sentences).






