---
title: "Lab 05 Activity"
author: "[Your Name]"
date: "[Today's Date]"
self-contained: true
---


### Activity Objectives:

- Conduct *t*-tests

- Independent samples *t*-test

- Paired samples *t*-test


### Instructions:

- Fill in the code chunks

- Render the Quarto document and submit both the `.qmd` and `.html` file on Canvas


## Activity

### 1) If you'd like to load packages with `library()`, please do so in this block. 

```{r}

```



## T-Tests


### 2) Perform an independent sample *t*-test on the difference between Petal Length and Petal Width for the `iris` dataset.

- Specify your null and alternative hypotheses

- Perform the t-test

- Check assumptions

```{r}

```


### 3) Perform an paired sample *t*-test on the difference between Petal Length and Petal Width for the `iris` dataset.

- Specify your null and alternative hypotheses

- Perform the t-test

- Check assumptions

```{r}

```



### 4) For this data, do you think an independent sample or paired sample *t*-test is more appropriate? (Hint: each row of petal measurements was taken on the same flower.)




### 5) For whichever analysis you chose in (#), provide a APA-style writeup of your results. Include your null and alternative hypotheses, *t*-statistic and confidence interval, and relevant details about assumptions.



