---
title: "Lab 01 Activity"
author: "[Your Name]"
date: "[Today's Date]"
---

[Link to Activity 01](https://github.com/jhelmer3/PSYC6019-2026/blob/main/lab/activities/lab_01_activity.qmd){.btn .btn-outline-primary target="_blank"}

### Activity Objectives:

- Do some basic R!

- Get familiar with Quarto

### Instructions:

1. Fill in the code chunks

2. Render the Quarto document and submit both the `.qmd` and `.html` file on Canvas


## Activity

### 1) Display the first few rows of the iris dataset.

```{r}

```


### 2) Calculate the mean and SD for the Sepal Width and Sepal Length variables. How do they compare to each other? (1--2 sentences).

```{r}

```


### 3) Three LaTeX equations are displayed on the slide ["Adding Math Equations"](https://jhelmer3.github.io/PSYC6019-2026/lab/slides/lab_01.html#/adding-math-equations). Pick two of them, and write their LaTeX equations below! Don't forget the dollar signs to make them render correctly.



### 4) Visit [Quarto's Markdown Basics page](https://quarto.org/docs/authoring/markdown-basics.html) and scroll through the features. Below, demonstrate three of Markdown's capabilities. One example would be: **bold text like this**.



### 5) The Quarto documentation has a [Gallery Page](https://quarto.org/docs/gallery/). Visit the link, and look through the neat things people have made with Quarto! For whichever is your favorite, include the link below with Markdown's link syntax. Also, check out its source code by visiting its accompanying `</>` "View Source Code" link.



### 6) What kinds of R topics are you hoping to learn this semester? If you have experience coding in R, are there any challenges you've encountered with it? Do you have experience in any other languages or statistical programming software?







