---
title: "Lab 04 Activity"
author: "[Your Name]"
date: "[Today's Date]"
self-contained: true
---


### Activity Objectives:

- Conduct $\chi^$-tests


### Instructions:

- Fill in the code chunks

- Render the Quarto document and submit both the `.qmd` and `.html` file on Canvas


## Activity

### 1) If you'd like to load packages with `library()`, please do so in this block. 

```{r}
```



## Chi-Squared Test of Proportions


### 2) Download and relocate the `leaves.csv` file [from the GitHub](https://github.com/jhelmer3/PSYC6019-2026/tree/main/lab/data/lab_04) to your folder for this class. Read in the file and display the first few rows of the data. This (simulated) dataset contains observations for 1,000 leaves on two variables:

- `color`: the color (green, red, orange, or brown) of the leaf

- `fallen`: whether or not the leaf had fallen from its tree

Use `mutate()` to change the `color` variable into an factor variable ordered by the leaf changing process (green -> red -> orange -> brown) first.

```{r}

```


### 3a) You are a tree researcher, and you would like to better understand leaves changing color in the fall. You've recorded the color of 1,000 leaves, and you'd like to test if there is a difference in the frequency of the colors of the leaves. You plan to conduct a $\chi^2$-Test of Goodness of Fit on the frequency of different color leaves you've observed in your dataset. First, what are the null and alternative hypotheses for this test?




### 3b) What is the expected frequency for leaves of each color? 




### 3c) What is your chi-square degrees of freedom and cutoff value?

```{r}

```



### 3d) What are the observed frequencies of each color?

```{r}

```


### 3e) What is the observed chi-squared statistic and associated p-value?

```{r}

```


### 3f) Provide a brief interpretation of these results.




### 3g) Create a barplot that displays the frequencies of each leaf color.

```{r}

```



## Chi-Squared Test of Independence


### 4a) You now want to see if leaf color and whether or not the leaf had fallen from the tree are independent variables. What are the null and alternative hypotheses for this test?




### 4b) What are the observed frequencies of each color given whether or not they had fallen?

```{r}

```


### 4c) What is the chi-squared statistic for this test?

```{r}

```


### 4d) Provide a brief interpretation of these results.





