---
title: "Lab 06 Activity"
author: "[Your Name]"
date: "[Today's Date]"
self-contained: true
---


### Activity Objectives:

- Calculate Cohen's $d$

- Perform a power analysis


### Instructions:

- Fill in the code chunks

- Render the Quarto document and submit both the `.qmd` and `.html` file on Canvas


## Activity

### 1) If you'd like to load packages with `library()`, please do so in this block. Also, please install and load the `palmerpenguins` R package! Comment out (or remove) your package installation line so it does not try to install when rendering.

```{r}

```


## Effect Size

### 2) The `palmerpenguins` dataset has data on the bill and flipper sizes of 434 penguins! Access the data object with the name `penguins`, and use a few functions to get a sense of the data.

NOTE: this dataset does contain some missing values! The `effectsize` package functions will handle those by dropping rows with missing values, but they may throw a warning.

```{r}

```


### 3) Your colleague is a penguin researcher, and they suspect that there is a difference in flipper length between male and female penguins, but they aren't sure how much. Calculate the Cohen's $d$ value of the difference between flipper lengths (`flipper_length_mm`) by penguin sex (`sex`). 

```{r}

```

### 4) Briefly interpret this effect size.



### 5) Next, your colleague wants to better understand the difference in size between penguins' bill length `bill_length_mm` and bill depth `bill_depth_mm`. Calculate the appropriate Cohen's $d$ value.

```{r}

```


### 6) Briefly interpret this effect size.




## Power Analysis

### 7) Calculate the power for a two-sample *t*-test with a true effect size of $\delta = 0.4$, sample size of 100 participants across groups, and significance level of $\alpha = .05$.

```{r}

```


### 8) You are about to run a study where you expect an effect size of $\delta = 0.4$ in your two-tailed t-test that you'll test at $\alpha = .05$. You want at least 85% power. Your PI asks you how many participants you'll need to collect which group, and would prefer to pay for the least amount of participants. 

- Calculate power for varying levels of $N$

- Create a power curve with sample size on the x-axis and power on the y-axis

- What sample size do you recommend?

```{r}

```














