---
title: "Lab 02 Activity"
author: "[Your Name]"
date: "[Today's Date]"
---

### Activity Objectives:

- Practice data visualization in `ggplot2`

### Instructions:

- Fill in the code chunks

- Render the Quarto document and submit both the `.qmd` and `.html` file on Canvas

*Note*: this activity is fairly open ended! As always, feel free to ask me (Jess) for help.


## Activity

### 1) Load in the `tidyverse` packages with `library(tidyverse)`. Then, read in the `world_cup_player_stats.csv` data and assign it to a variable.

```{r}

```


### 2) Use at least two functions to explore and 'get a sense of' the data (HINT: `glimpse()` can be one of them).

```{r}

```

### 3) Come up with one question you'd like to ask of the data through data visualization (feel free to use the slides as inspiration, but it shouldn't be the exact same), and write the question below.




### 4) Before you start plotting, have an idea of your end goal in mind. Below, describe what the plot should overall look like (ignoring the actual trends you'd need to see to describe). For example, what would be on the x-axis? The y-axis? What type of plot would it be?




### 5) Start by creating just the bare bones of the plot. Use the `ggplot()` function and set the `data` and `mapping` argument. For the `mapping` argument, think carefully about which variables you want to set on which axes using the `aes()` function.

```{r}

```


### 6) Next, add a `geom_` layer. 

```{r}

```


### 7) Briefly interpret this plot (1--2 sentences).




### 8) Make at least one structural modification to this plot. By structural modification, I mean changing either (1) something in the `data` being plotted, (2) something in the data mapping in the `aes()` argument, or (3) the `geom`(s) being plotted.

```{r}

```


### 9) Briefly interpret this new plot (1--2 sentences).




### 10) Finally, make at least three additional modifications to this plot. This can include changes to the theme, scales, labels, etc. Please provide a brief justification for each change (e.g., making the plot visually cleaner, clarifying expected breaks in scale, adding text to explain the plot).




### 11) The [R Graph Gallery](https://r-graph-gallery.com/) contains lots of excellent plots. Browse through, and include the link to two plots you like below! Provide some brief commentary / description as to why you like those specific ones, and for each, try to identify a code snippet that connects to (or creates) the reason why you like it. See [this section in the Quarto Markdown Guide](https://quarto.org/docs/authoring/markdown-basics.html#source-code) for how to use Markdown code styling for the snippets.




