---
title: "Lecture 16: One Way ANOVA"
author: "Mark Himmelstein"
format: 
  revealjs:
    slide-width: 1920
    slide-height: 1080
    theme: ../../../gt-theme.scss
    slide-number: true
    slide-number-format: "c/t"
    html-math-method: mathjax
    include-in-header:
      text: |
        <script>
        window.MathJax = {
          tex: {
            packages: {'[+]': ['cancel']}
          },
          loader: {
            load: ['[tex]/cancel']
          }
        };
        </script>
params:
  v0: false 
editor: source
---

```{r}
#| echo: false


library(knitr)
library(formatR)
library(readxl)
library(tidyverse)
opts_chunk$set(out.width = '100%', options(width=60))

opts_knit$set(global.par = TRUE)
opts_chunk$set(echo=FALSE, warning = FALSE, message = FALSE, comment = '', class.output='sh',
               fig.width=8, fig.height=5, fig.align = "center")

my_red <- "#E41A1C"
my_blue <- "#377EB8"
my_green <- "#4DAF4A"
my_purple <- "#984EA3"
my_orange <- "#FF7F00"
my_yellow <- "#FFFF33"
my_red2 <- "#FBB4AE" 
my_blue2 <- "#B3CDE3" 
my_green2 <- "#CCEBC5" 
my_purple2 <- "#DECBE4"
my_orange2 <- "#FED9A6"
my_yellow2 <- "#FFFFCC"

```

## Frequentist and Bayesian

Today will give us a picture for how the rest of the course will go

- We will learn about one way ANOVA as a method for comparing the means of multiple groups
- Review the "traditional" frequentist perspective
- Contrast this with the Bayesian approach
- Emphasize the model based perspective


## Analysis of Variance

::: {style="font-size: 0.95em;"}

Analysis of variance (ANOVA) is one of the most popular statistical tests.

Today, we'll learn about the simplest case of ANOVA (one-way ANOVA):

- Groups are defined between (not within) subjects
  - Separate individuals in separate groups
- One independent variable with 3+ independent groups
- Can be considered generalization of independent sample $t$-test to more than two groups

Although it's called the analysis of **variance**, we use ANOVA to test hypotheses about **means**. For our frequentist setup:

$$H_0: \mu_1 = \mu_2 = \mu_3 = \cdots$$

$$H_1: \text{at least one of the means are different}$$

:::


## Example Data

Example: weight of plants under one of three conditions

- Control (ctrl), 10 observations
- Treatment 1 (trt1), 10 observations
- Treatment 2 (trt2), 10 observations

::: {style="font-size: 0.85em;"}

```{r}
#| echo: true
data("PlantGrowth", package = "datasets")
summary(PlantGrowth)
```

:::

## Group Means

```{r}
ggplot(PlantGrowth, aes(group, weight, fill = group)) + 
  geom_boxplot() + theme_light() + 
  scale_fill_manual(values = c(my_purple2, my_green2, my_blue2)) + 
  theme(legend.position = "none") + labs(x = "") + 
  stat_summary(fun=mean, geom="point", shape=7, size=4)
```

::: {.fragment style="text-align: center; margin-top: -10px;"}
Are these means equal **in the population?**
:::


## The Traditional Model

Suppose subject $i$ belongs to group $g$. The observed response is $y_{ig}$.

The **structural model** decomposes observations into different sources of variability (hence Analysis of **Variance**):

$$\begin{align*}
y_{ig} &= \color{#377EB8}{\mu} + \color{#4DAF4A}{(\mu_g - \mu)} + \color{#984EA3}{(y_{ig} - \mu_g)} \\
y_{ig} &= \color{#377EB8}{\mu} + \color{#4DAF4A}{\tau_g} + \color{#984EA3}{\epsilon_{ig}}
\end{align*}$$

- $\mu$: grand mean, population mean disregarding group membership
- $\mu_g$: population group mean for group $g$
- $\tau_{g}$: deviation of group $g$'s mean from the grand mean
- $\epsilon_{ig}$: deviation of individual $i$ from group $g$'s mean



## [Map to General Linear Model]{style="font-size: 0.8em;"}

$$\begin{align*}
y_{ig} &= \color{#377EB8}{\beta_0} + \color{#4DAF4A}{\beta_1 x_{1i} + \beta_2 x_{2i}} + \color{#984EA3}{\epsilon_i} 
\end{align*}$$

Where $x_1$ and $x_2$ are defined as follows:

|  | $g_1$ | $g_2$ | $g_3$ |
| :---: | :---: | :---: | :---: |
| $x_1$ | 1 | 0 | -1 |
| $x_2$ | 0 | 1 | -1 |

::::: {.columns .v-center-container}
::: {.column width="30%"}
And

- $\beta_0: = \mu$
- $\beta_1: = \mu_{1} - \mu$
- $\beta_2: = \mu_{2} - \mu$

:::

::: {.column width="70%"}

::: {style="font-size: 0.75em;"}


[What about $\mu_{3}$?]{.fragment}



[Let's say $\mu = 2$, $\mu_{1} = 1$, and $\mu_{2} = 3$ (equal sample sizes in each group). What is $\mu_3?$]{.answer .fragment .content-visible unless-meta="params.v0"}



:::: {.content-visible unless-meta="params.v0"}
::: {.answer .fragment}
$\mu_{3} = \mu - (\mu_{1} - \mu) - (\mu_{2} - \mu) =$
$2 - (1 - 2) - (2 - 2) = 3$
:::
::::

:::: {.content-visible unless-meta="params.v0"}
::: {.answer .fragment}
$\mu_{3} = \mu - (\mu_{1} - \mu) - (\mu_{2} - \mu) =$
$\beta_0 - \beta_1 - \beta_2$
:::
[Coding $g - 1$ $x$'s give us all the info about $g$ group means]{.answer .fragment .content-visible unless-meta="params.v0"}
:::
::::



:::
::::: 


## Coding Options

::: {style="font-size: 0.75em;"}


Effects or contrast coding compares $g-1$ groups to $\beta_0 =$ **grand mean**:

| Predictor | $g_1$ | $g_2$ | $\cdots$ | $g_{G-1}$ | $g_G$ |
| :---: | :---: | :---: | :---: | :---: | :---: |
| $x_1$ | 1 | 0 | $\cdots$ | 0 | -1 |
| $x_2$ | 0 | 1 | $\cdots$ | 0 | -1 |
| $\vdots$ | $\vdots$ | $\vdots$ | $\ddots$ | $\vdots$ | $\vdots$ |
| $x_{G-1}$ | 0 | 0 | $\cdots$ | 1 | -1 |


Dummy coding compares $g-1$ groups to $\beta_0 =$ **reference group mean**:

| Predictor | $g_1$ | $g_2$ | $\cdots$ | $g_{G-1}$ | $g_G$ |
| :---: | :---: | :---: | :---: | :---: | :---: |
| $x_1$ | 1 | 0 | $\cdots$ | 0 | 0 |
| $x_2$ | 0 | 1 | $\cdots$ | 0 | 0 |
| $\vdots$ | $\vdots$ | $\vdots$ | $\ddots$ | $\vdots$ | $\vdots$ |
| $x_{G-1}$ | 0 | 0 | $\cdots$ | 1 | 0 |

Coding scheme of choice **only matters for interpreting coefficients** ($\beta$'s)

:::

## Coding Scheme Comparison

$$\begin{align*}
y_{ig} &= \color{#377EB8}{\beta_0} + \color{#4DAF4A}{\beta_1 x_{1i} + \beta_2 x_{2i}} + \color{#984EA3}{\epsilon_i} 
\end{align*}$$



::::: {.columns .v-center-container}
::: {.column width="50%"}
**Contrast/Effects coding:**

- $\beta_0: = \mu$
- $\beta_1: = \mu_{1} - \mu$
- $\beta_2: = \mu_{2} - \mu$

Compares groups to **grand mean**

:::

::: {.column width="50%"}

**Dummy coding:**

- $\beta_0: = \mu_3$
- $\beta_1: = \mu_{1} - \mu_3$
- $\beta_2: = \mu_{2} - \mu_3$

Compares groups to **reference group mean** (in this case group three)

:::
:::::


## Decomposition

Sticking with the traditional model structure:

$$y_{ig} = \color{#377EB8}{\mu} + \color{#4DAF4A}{\tau_g} + \color{#984EA3}{\epsilon_{ig}}$$

From this equation, scores are affected by three effects:

- [**Overall mean effect**]{style="color: #377EB8;"}
- [**Between Group (treatment) effect**]{style="color: #4DAF4A;"}
- [**Within Group effects (individual differences)**]{style="color: #984EA3;"}

In ANOVA, we consider the magnitude of these effects by **decomposing** the [**total variance**]{style="color: #377EB8;"} into [**Between**]{style="color: #4DAF4A;"} and [**Within**]{style="color: #984EA3;"} group effects.


## Decomposition Visual

```{r}
#| message: false
mn <- mean(PlantGrowth$weight)

p1 <- ggplot(PlantGrowth, aes(x = 1, y = weight, shape = group)) + 
  geom_point(size = 2, color = my_blue) + theme_light() +  
  scale_shape_manual(values = c(8, 18, 16)) + 
  ylim(c(3.5,7)) +
  theme(legend.position = "none", axis.text.x = element_blank(),
        strip.background = element_rect(fill=my_blue)) + 
  labs(x = "") + facet_wrap(~"Total Variance") + geom_hline(yintercept = mn, 
                                                            linetype = 2, color = my_blue)

p2 <- ggplot(PlantGrowth %>% group_by(group) %>% summarize(mean = mean(weight)),
       aes(group, mean, shape = group)) + geom_point(size = 8, color = my_green) + theme_light() + 
    scale_shape_manual(values = c(8, 18, 16)) + 
  ylim(c(3.5,7)) + xlab(NULL) +
  theme(legend.position = "none",
        strip.background = element_rect(fill=my_green))  + facet_wrap(~"Group Means")  + geom_hline(yintercept = mn, 
                                                            linetype = 2, color = my_blue)

PlantGrowth <- PlantGrowth %>% group_by(group) %>% mutate(mean = mean(weight))

plantmns <- PlantGrowth %>% group_by(group) %>%
  summarize(weight = 0)

p3 <- PlantGrowth %>% group_by(group) %>% 
  mutate(weight = weight - mean(weight)) %>% 
  ggplot(aes(group, weight, shape = group)) + 
  geom_point(size = 3, color = my_purple) + 
    geom_point(aes(group, weight, shape = group), data = plantmns, size = 8,
             color = my_green) + theme_light() + 
    scale_shape_manual(values = c(8, 18, 16)) + 
  labs(x = NULL, y = "deviation scores") + 
  ylim(c(-1.5,2)) +
  theme(legend.position = "none",
        strip.background = element_rect(fill=my_purple)) + 
  facet_wrap(~"Variance Within Groups") + geom_hline(yintercept = 0, 
                                                            linetype = 2, color = my_green)

gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```



## Variance of Means

::: {style="font-size: 0.8em;"}

In the one-way ANOVA, the null hypothesis for $G$ groups is

$$H_0: \mu_1 = \mu_2 = \cdots = \mu_G$$

An equivalent way to express this is:

$$H_0: \mu_1 - \mu_2 = \mu_1 - \mu_3 = \cdots = \mu_{-1} - \mu_G = 0$$

Or, more succinctly 

$$H_0: \frac{\sum_{g=1}^G (\mu_g - \mu)^2}{G-1} = 0$$

**In English:** 

- If **population** group means are equal, then the **population** variance of group means equals 0
- **The problem (the one we always have):** even if $H_0$ is true, not all **sample** group means will be equal due to random sampling variation

:::



## Sums of Squares I

We want to separate the variability in data into:

- Variability between different group means
- Variability within groups

**Step 1**: Sums of Squares (SS)

- Sound familiar? Think: numerator of the variance

$$s^2 = \frac{\sum(x - \bar{x})^2}{N-1}$$

$$SS = \sum(x - \bar{x})^2$$

## Sums of Squares II

We want to separate the variability in data into:

- Variability between different group means
- Variability within groups

**Step 1**: Sums of Squares

- In ANOVA we **decompose** the Sums of Squares into three components: 
  - Sum of squares total ($SST$)
  - Sum of squares between groups/treatments ($SSB$)
  - Sum of squares within groups (error, residuals) ($SSW$)

## Sums of Squares III

::: {style="font-size: 0.8em;"}

Individual scores $y_{ig}$:

- $i$: individual indicator
- $g$: group indicator
- $n_g$: number of observations in group $g$
- $\bar{y}$: sample grand mean (of all scores)

Sample variation of $y = \sum_g\sum_i(y_{ig} - \bar{y})^2$

$$\begin{align*}
\underbrace{\sum_g\sum_i(y_{ig} - \bar{y})^2}_{\color{#377EB8}{SST}} &= \underbrace{\sum_gn_g(\bar{y}_g - \bar{y})^2}_{\color{#4DAF4A}{SSB}} + \underbrace{\sum_g\sum_i(y_{ig} - \bar{y}_g)^2}_{\color{#984EA3}{SSW}}
\end{align*}$$

- [**$SST$**]{style="color: #377EB8;"}: total variation of participants scores (total effect)
- [**$SSB$**]{style="color: #4DAF4A;"}: total variation between groups (between-group effect)
- [**$SSW$**]{style="color: #984EA3;"}: variation within groups (within-group effect)

::: 

## Sums of Squares IV

::: {style="font-size: 0.8em;"}

Plant Data: Overall $\bar{y} = 5.073$

<style>
.grouped-table table {
  border-collapse: collapse;
}
/* Border before 1st column */
.grouped-table th:nth-child(1),
.grouped-table td:nth-child(1) {
  border-left: 2px solid #444;
}
/* Borders after 2nd, 4th, and 6th columns */
.grouped-table th:nth-child(2),
.grouped-table td:nth-child(2),
.grouped-table th:nth-child(4),
.grouped-table td:nth-child(4),
.grouped-table th:nth-child(6),
.grouped-table td:nth-child(6) {
  border-right: 2px solid #444;
}
</style>

::: {.grouped-table}
| Control | | Treatment 1 | | Treatment 2 | |
| :---: | :---: | :---: | :---: | :---: | :---: |
| 4.17 | 4.61 | 4.81 | 3.83 | 6.31 | 5.29 |
| 5.58 | 5.17 | 4.17 | 6.03 | 5.12 | 4.92 |
| 5.18 | 4.53 | 4.41 | 4.89 | 5.54 | 6.15 |
| 6.11 | 5.33 | 3.59 | 4.32 | 5.50 | 5.80 |
| 4.50 | 5.14 | 5.87 | 4.69 | 5.37 | 5.26 |
| **$\bar{y}_1 = 5.03$** | | **$\bar{y}_2 = 4.66$** | | **$\bar{y}_3 = 5.53$** | |
:::
$$SST = (4.17 - 5.073)^2 + (5.58 - 5.073)^2 + \cdots + (5.26 - 5.073)^2 = \mathbf{14.26}$$

$$SSB = 10(5.03 - 5.073)^2 + 10(4.66 - 5.073)^2 + 10(5.53 - 5.073)^2 = \mathbf{3.77}$$

$$SSW = (4.17 - 5.03)^2 + \cdots + (4.81 - 4.66)^2 + \cdots + (5.26 - 5.53)^2 = \mathbf{10.49}$$

$$SSW = SST - SSB = 14.26 - 3.77 = \mathbf{10.49}$$

:::

## Mean Squares I

**Step 2**: Find **degrees of freedom** and calculate **mean squares**

::: {style="font-size: 0.85em;"}

| Component | Sum of Squares | Degrees of Freedom | Mean Squares |
| :--- | :--- | :--- | :--- |
| **Between-group effect** | $\sum_{g}n_g(\bar{y}_{g}-\bar{y})^{2}$ | $dfB = G - 1$ | $MSB = \frac{SSB}{dfB}$ |
| **Within-group effect** | $\sum_{g}\sum_{i}(y_{ig}-\bar{y}_{g})^{2}$ | $dfW = N - G$ | $MSW = \frac{SSW}{dfW}$ |
| **Total effect** | $\sum_{g}\sum_{i}(y_{ig}-\bar{y})^{2}$ | $dfT = N - 1$ | $MST = \frac{SST}{dfT} = s^2_y$ |

$$SST = SSB + SSW \qquad dfT = dfB + dfW$$

- $G$: number of groups
- $N$: total sample size

:::

## Mean Squares II

::: {style="font-size: 0.7em;"}

Mean square $= SS / df$

**Interpretation:** Mean squares can be interpreted like variances

- $MST$: overall variance of $y$ across participants and groups
- $MSB$: variance of the between-group effect
- $MSW$: variance of the within-group effect (residual/error variance)

:::: {.columns}

::: {.column width="55%"}
::: {style="background-color: #e8f5e9; border-left: 4px solid #4DAF4A; padding: 10px; margin-bottom: 10px;"}
**Across repeated sampling:**

$$\text{mean}(MSB) = N\theta^2_B + \sigma^2_W$$
$$\text{mean}(MSW) = \sigma^2_W$$

- $\theta^2_B$: population variance of between-group effect
- $\sigma^2_W$: population variance of within-group effect
:::
:::

::: {.column width="45%"}
::: {style="background-color: #e3f2fd; border-left: 4px solid #377EB8; padding: 10px; margin-bottom: 10px;"}
**If $H_0$ is true:**

$$\theta^2_B = 0$$
:::

::: {style="background-color: #ffebee; border-left: 4px solid #E41A1C; padding: 10px;"}
**If $H_1$ is true:**

$$\theta^2_B > 0$$
:::
:::

::::

So what comparison does this imply for our test in terms of our **sample** statistics?



:::


## Mean Squares III

::: {style="font-size: 0.7em;"}

:::: {.columns}

::: {.column width="55%"}
::: {style="background-color: #e8f5e9; border-left: 4px solid #4DAF4A; padding: 10px;"}
**Across repeated sampling:**

$$\text{mean}(MSB) = N\theta^2_B + \sigma^2_W$$
$$\text{mean}(MSW) = \sigma^2_W$$

- $\theta^2_B$: population variance of between-group effect
- $\sigma^2_W$: population variance of within-group effect
:::
:::

::: {.column width="45%"}
::: {style="background-color: #e3f2fd; border-left: 4px solid #377EB8; padding: 10px; margin-bottom: 10px;"}
**If $H_0$ is true:**

$$\text{mean}(MSB) = N(0) + \sigma^2_W$$
$$\text{mean}(MSW) = \sigma^2_W$$

$$\frac{\text{mean}(MSB)}{\text{mean}(MSW)} = 1$$
:::

::: {style="background-color: #ffebee; border-left: 4px solid #E41A1C; padding: 10px;"}
**If $H_1$ is true:**

$$\text{mean}(MSB) = N(\theta^2_B) + \sigma^2_W$$
$$\text{mean}(MSW) = \sigma^2_W$$

$$\frac{\text{mean}(MSB)}{\text{mean}(MSW)} > 1$$
:::
:::

::::

:::


## ANOVA Table

::: {style="font-size: 1.1em;"}

| | **SS** | **df** | **MS** |
| :--- | :---: | :---: | :---: |
| **Between Groups** | 3.77 | 2 | 1.88 |
| **Within Groups** | 10.49 | 27 | 0.39 |
| **Total** | 14.26 | 29 | |

:::

## $F$ Distribution

The $F$ distribution includes two parameters:

- Numerator $df$ ($dfB$)
- Denominator $df$ ($dfW$)

Properties of the $F$ distribution:

- $F$ is defined on the interval $[0, \infty]$
- The $t$ distribution is equivalent to a special case of the $F$ distribution
  - If $dfB = 1$, $F(1, dfW) = t(dfW)^2$

## $F$ Distribution Visual

```{r}
x <- tibble(x = seq(.3, 5, length = 100))
x <- x %>% mutate("dfB = 1, dfW = 1" = df(x, 1, 1),
           "dfB = 1, dfW = 3" = df(x, 1, 3),
           "dfB = 3, dfW = 10" = df(x, 3, 10),
           "dfB = 3, dfW = 20" = df(x, 3, 20),
           "dfB = 20, dfW = 20" = df(x, 20, 20))
x <- x %>% pivot_longer("dfB = 1, dfW = 1":"dfB = 20, dfW = 20", 
                        names_to = "type", values_to = "density")
ggplot(x, aes(x, density, group = type, col = type)) + 
geom_path(linewidth = 1.5) + theme_light() + 
scale_color_manual(values = c(my_blue, my_purple, my_red, my_green, my_orange)) + 
  xlab(NULL) + 
theme(legend.title = element_blank())
```

## Hypothesis Testing

- $H_0: \mu_1 = \mu_2 = \cdots = \mu_G$
- $H_1$: at least one population mean is different

Test statistic: $F$-statistic

$$F(dfB, dfW) = \frac{MSB}{MSW}$$

- $dfB = G - 1 \qquad (G = \text{number of groups})$
- $dfW = N - G \qquad (N = \text{total sample size})$

When $H_0$ is true, $F$ test statistic follows the $F$ distribution with $dfB$ numerator and $dfW$ denominator degrees of freedom.

These calculations work for any number of groups and any group sizes.

## ANOVA Table and Test

::: {style="font-size: 1.1em;"}

| | **SS** | **df** | **MS** | **F** | **p** |
| :--- | :---: | :---: | :---: | :---: | :---: |
| **Between Groups** | 3.77 | 2 | 1.88 | 4.85 | .02 |
| **Within Groups** | 10.49 | 27 | 0.39 | | |
| **Total** | 14.26 | 29 | | | |

:::

**Interpretation:** Were the null hypothesis true, there is only a 2% chance we would have observed variance between group means at least this large.

## Effect Size


## Effect Sizes

**$d$-family**: standardized **difference** between groups
- Cohen's $d$: standardized mean differences between pairs of groups
- Will come back to this more next time

**$r$-family**: standardized **association** between two variables
- $\eta^2$ aka $R^2$: proportion of variance of outcome accounted for by predictor (group membership)
- $\omega^2$: adjusted $\eta^2$
- $f^2$: used for statistical power analysis


## $\eta^2$ Effect Size

$$\eta^2 = \frac{SSB}{SST} = \frac{SSB}{SSB + SSW} = R^2 = 1 - \frac{s^2_{\epsilon}}{s^2_y} = \frac{s_{\hat{y}}^2}{s_y^2}$$



**Interpretation:**

- Proportion of variation of outcome accounted for by group differences
- Usually $\eta^2$ in ANOVA and $R^2$ in regression (more later)
- Is a positively biased statistic to the population parameter
- Positive bias: on average, statistic is larger than the parameter
  - We are more likely to think our effect is too large than too small due to sample variation
  - Sometimes called "overfit"


## $f^2$ Effect Size

$r$-family effect size measure, mostly used for conducting power analyses.

$$f^2 = \frac{\eta^2}{1 - \eta^2} = \frac{SSB}{SSW}$$

**Interpretation:** Ratio of between-group variability to within-group variability.

Cohen's (1988, 1992) suggestions:

- $f = 0.10, f^2 = 0.0100, \eta^2 = 0.010$: small effect
- $f = 0.25, f^2 = 0.0625, \eta^2 = 0.059$: medium effect
- $f = 0.40, f^2 = 0.1600, \eta^2 = 0.138$: large effect

In practice, you may want to report effect size in terms of $\eta^2$ instead of $f^2$ because it is easier to interpret.

## $\omega^2$ Effect Size

$\omega^2$ is less biased than $\eta^2$:

$$\omega^2 = \frac{SSB - (G - 1)\times MSW}{SST + MSW}$$

- $G$: number of groups
- Similar (but not identical to) to adjusted $R^2$ in regression (more later)
- $\omega^2$ is not completely unbiased)

**Interpretation:** Same interpretation as $\eta^2$ (proportion of variation accounted for by group differences).

## The Story So Far

- This is known as the "omnibus" test
    - Is there **any** difference in group means **anywhere** among **all** groups?
    - Says nothing about which specific groups differ
    - Will get to that next time!
- This is a bit awkward from a Bayesian perspective
    - Some would argue all we really want to do in Bayes is obtain probability distributions about what the means of each group are likely to be
      - Can then directly compare groups to one another
    - Closest corollary is Bayes factor
    - How much more/less likely is it that the means differ than they do not?
    


## Bayes Factor Refresher

Bayes Factors are a way of comparing the support for different models in a Bayesian framework.

They are ratios of posterior likelihoods between competing models. If we have two possible models for the same data, $M_1$ and $M_2$.

$$BF = \frac{P(\boldsymbol{D}|M_1)}{P(\boldsymbol{D}|M_2)}$$

- When $BF = 1$ the two models are equally likely
- When $BF > 1$ then $M_1$ is more likely than $M_2$
  - When $BF = 3$, $M_1$ is three times as likely to be the "true" model as $M_2$
- When $BF < 1$, $M_2$ is more likely than $M_1$
    - When $BF = \frac{1}{3}$, $M_2$ is three times as likely to be the "true" model as $M_1$


## Bayes Factor Setup

**Traditional ANOVA Model Setup**

- $\boldsymbol{M_2}: y_{ig} = \color{#377EB8}{\mu} + \color{#984EA3}{\epsilon_{ig}}$
- $\boldsymbol{M_1}: y_{ig} = \color{#377EB8}{\mu} + \color{#4DAF4A}{\tau_g} + \color{#984EA3}{\epsilon_{ig}}$

**General Linear Model Setup**

- $\boldsymbol{M_2}: y_{ig} = \color{#377EB8}{\beta_0} + \color{#984EA3}{\epsilon_{ig}}$
-  $\boldsymbol{M_1}: y_{ig} = \color{#377EB8}{\beta_0} + \color{#4DAF4A}{\sum_{p=1}^{g-1}\beta_p x_{p}} + \color{#984EA3}{\epsilon_i}$

Where $x_p$ and are $g-1$ coded categorical variables representing group membership.

**Interpretation:** The means being different is $BF$ times as likely as the means being equal

## Jeffreys Lindley Paradox



When using Bayes factors to "test the null hypothesis" that all means are equal

- Using a "weaker" prior, e.g. $\text{normal}(0, 10)$, for the $\beta$'s can be more likely to support the null model $M_2$
- Using a "stronger" prior, e.g. $\text{normal}(0, 1)$, for the $\beta$'s can be more likely to support the full model $M_1$

Why would a stronger prior that the null hypothesis is likely to be **true** make it so a Bayes factor would say that a model that represents an **alternative** hypothesis is more likely than the null?

## Jeffreys Lindley Reason

Remember, priors are probability distributions, not point hypotheses

- Stronger prior means there is more support for no difference, but **also** that very small differences are more likely than very large differences
    - Will make posterior likelihood larger if there are only small differences in the sample means
- Weaker prior will give less support to very small differences
    - Posterior likelihood less likely to be concentrated in regions reflecting small differences

**If unsure, do analysis with a few different priors** (sensitivity analysis)


::: {style="font-size: 0.6em;"}
Robert, C. P. (2014). On the Jeffreys-Lindley paradox. Philosophy of Science, 81(2), 216-232. [https://doi.org/10.1086/675729](https://doi.org/10.1086/675729)
:::

## Posterior Predictive

Philosophically Bayesian inference is more about using probability to represent belief.

- The **posterior predictive distribution** is a probability distribution representing our belief about possible values $y$ values might take on conditional on given $x$ values
    - E.g., a probability distribution representing plausible values for a random member from a specific group
    - If posterior predictive distribution for all groups are very similar, it means we believe differences between groups are small
    - If posterior predictive distributions for some groups are very different, it means we believe differences between groups are large
    
## Bayesian $R^2$

::: {style="font-size: .9em;"}

Because Bayesian estimates aren't deterministic the way the simple frequentist ANOVA algorithm is, the traditional $R^2$ is not guaranteed to be $0 \leq R^2 \leq 1$

- Means it is possible for the model to explain more than the total variance
    - Not a sensible interpretation!
    
Alternative:

$$\text{Bayesian } R^2 = \frac{s_{\hat{y}}^2}{s_{\hat{y}}^2 + s_{\epsilon}^2}$$


- Is guaranteed to be $0 \leq \text{Bayesian } R^2 \leq 1$
- Same interpretation as $\eta^2$ and $R^2$

::: 



## Another Example

```{r}


dat <- read_xls("../../data/Williams data.xls")

dat <- dat %>% mutate(average_bond = (SIB_BOND + PAR_BOND + HOME_BOND) / 3) %>%
  filter(!is.na(average_bond)) %>%
  mutate(condition = recode(factor(COND), "1" = "close", "2" = "intermediate", "3" = "distant")) %>% dplyr::select(-COND)

dat2 <- dat %>% 
  pivot_longer(c(SIB_BOND:HOME_BOND, average_bond), names_to = "type", values_to = "strength") %>%
  mutate(type = recode(type, SIB_BOND = "Sibling", PAR_BOND = "Parent", HOME_BOND = "Home"))

ggplot(dat2 %>% filter(type == "average_bond")) + 
  geom_boxplot(aes(x = condition, y = strength, fill = condition)) + 
  scale_fill_manual(values = c(my_purple2, my_green2, my_blue2)) + 
  theme(legend.position = "bottom") + theme_light() + 
  facet_grid(. ~ "Average of Bond Types") + 
  labs(x = "", y = "average bond strength rating") + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) + 
  stat_summary(aes(x = condition, y = strength), fun=mean, geom="point", shape=7, size=4)
```



## ANOVA in R: Way 1

The default `R` way to estimate an ANOVA model is with the `aov()` function using formula syntax.

We can then get the $F$-test with the `summary()` function.

```{r}
#| echo: true
a1 <- aov(average_bond ~ condition, dat)
summary(a1)
```

## ANOVA in R: Way 2

Between subjects one-way ANOVA in R is easy. Later it gets confusing when we have within subjects designs. The `ez` package and `ezANOVA()` function can do both easily, though you have to make sure to specify a subject ID as `wid = ID` even for between subjects analysis.

```{r}
#| echo: true
#| warning: false
#| message: false
library(ez)
ez1 <- ezANOVA(data = dat, dv = average_bond, wid = SUBJ, 
               between = condition)
ez1$ANOVA
```

What do you think `ges` might be?

[Stands for **generalized** $\eta^2$ (same as our $\eta^2$) so far, slightly different in some other cases]{.answer .fragment .content-visible unless-meta="params.v0"}


## $\eta^2$ Example

```{r}
summary(a1)
```

$$\eta^2 = \frac{SSB}{SSB + SSW} = \frac{0.49}{0.49 + 125.9} = .004$$

```{r}
#| echo: true
#| message: false
#| warning: false
library(effectsize)
eta_squared(a1)
```



## $\omega^2$ Example

```{r}
summary(a1)
```

$$\omega^2 = \frac{SSB - (G - 1)\times MSW}{SST + MSW} = \frac{0.49 - (2)\times 1.03}{0.49 + 125.9 + 1.03} = -0.01$$

*(Yes, $\omega^2$ can be negative, but we generally just interpret that as $\omega^2 = 0$, or more precisely, $\omega^2 < .001$)*

```{r}
#| echo: true
#| message: false
#| warning: false
omega_squared(a1)
```

## Bayesian Version

By default, most `R` packages use dummy coding. What is the reference group in this output? ["Close", since the coefficients represent "intermediate" and "distant" (differences with "close")]{.answer .fragment .content-visible unless-meta="params.v0"}


```{r}
#| echo: true
library(rstanarm)
options(mc.cores = parallel::detectCores())
```

```{r}
#| echo: true
#| cache: true
stan_lm1 <- stan_glm(average_bond ~ condition, dat, family = "gaussian",
                     diagnostic_file = "df1.csv")

summary(stan_lm1)
```

## Posterior Predictive Example

For one way ANOVA, we can use the `emmeans::emmeans()` function to get posterior predictive distributions with 95% HPD credible intervals

```{r}
#| echo: true
#| cache: true
library(emmeans)
em1 <- emmeans(stan_lm1, ~ condition)
summary(em1, point.est = "mean")

```


## Bayesian $R^2$

```{r}
#| echo: true
#| cache: true
library(performance)
r2_bayes(stan_lm1)
```


## Bayes Factors

```{r}
#| echo: true
#| cache: true
library(bayestestR)
stan_lm2 <- stan_glm(average_bond ~ 1, dat, family = "gaussian",
                     diagnostic_file = "df2.csv")

(bfs1 <- bayesfactor_models(stan_lm2, stan_lm1))


```
## Jeffreys Lindley Paradox

Let's see what happens if I use a stronger prior that the $\beta$'s are 0 (more likely that there are no group differences)

```{r}
#| echo: true
#| cache: true
stan_lm1a <- stan_glm(average_bond ~ condition, dat, family = "gaussian",
                      prior = normal(0, .1),
                     diagnostic_file = "df1a.csv")

(bfs1 <- bayesfactor_models(stan_lm2, stan_lm1a))

```


Bayes factor now says the full model is about as likely than the null model!

